---
title: "Bayesian Analysis of Binomial Ordering Preferences"
author: "Zachary Houghton"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
params:
  slug:  "znhoughton_opt-babylm-125m-20eps-seed964"
  layer: "last"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = FALSE
)
set.seed(964)

# Helper: check Rhat, ESS, and divergences; write to warnings file and stop if bad
check_diagnostics <- function(fit, label, warn_file) {
  issues <- c()

  diag <- summarise_draws(fit, "rhat", "ess_bulk", "ess_tail") %>%
    filter(str_starts(variable, "b_") | str_starts(variable, "sd_") |
             variable == "sigma")

  max_rhat <- max(diag$rhat, na.rm = TRUE)
  min_ess  <- min(diag$ess_bulk, na.rm = TRUE)

  if (max_rhat > 1.01)
    issues <- c(issues, sprintf("HIGH Rhat = %.4f (threshold 1.01)", max_rhat))
  if (min_ess < 400)
    issues <- c(issues, sprintf("LOW ESS bulk = %.0f (threshold 400)", min_ess))

  np <- nuts_params(fit)
  n_div <- sum(np$Value[np$Parameter == "divergent__"])
  if (n_div > 0)
    issues <- c(issues, sprintf("%d divergent transitions", n_div))

  if (length(issues) > 0) {
    msg <- paste0(
      "\n!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!\n",
      "WARNING [", label, "]:\n",
      paste(" -", issues, collapse = "\n"), "\n",
      "!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!\n"
    )
    cat(msg)
    cat(paste0(Sys.time(), "  ", msg), file = warn_file, append = TRUE)
  } else {
    cat(sprintf("[%s] diagnostics OK  (Rhat=%.4f, min ESS=%.0f, divergences=0)\n",
                label, max_rhat, min_ess))
  }
}
```

```{r libraries}
library(here)
library(tidyverse)
library(brms)
library(BayesFactor)
library(bayestestR)   # rope(), describe_posterior()
library(posterior)    # summarise_draws()
library(ggplot2)
library(knitr)
library(kableExtra)
```

```{r paths}
BASE        <- here::here()
SLUG        <- if (exists("params")) params$slug  else "znhoughton_opt-babylm-125m-20eps-seed964"
LAYER       <- if (exists("params")) params$layer else "last"
RESULTS_DIR <- file.path(BASE, "Results", SLUG, paste0("layer_", LAYER))

DELTA_F    <- file.path(RESULTS_DIR, "delta_features.csv")
NOVEL_F    <- file.path(RESULTS_DIR, "novel_pls_scores.csv")
CORPUS_F   <- file.path(RESULTS_DIR, "corpus_pls_scores.csv")
WARN_FILE  <- file.path(RESULTS_DIR, "brms_warnings.txt")

# Clear warnings file at the start of each run
if (file.exists(WARN_FILE)) file.remove(WARN_FILE)

cat("Model slug:", SLUG, "\n")
```

```{r load-data}
delta_df  <- read_csv(DELTA_F,  show_col_types = FALSE)
novel_df  <- read_csv(NOVEL_F,  show_col_types = FALSE)
corpus_df <- read_csv(CORPUS_F, show_col_types = FALSE)

cat("Delta features:  ", nrow(delta_df),  "pairs\n")
cat("Novel PLS scores:", nrow(novel_df),  "pairs\n")
cat("Corpus PLS scores:", nrow(corpus_df), "pairs\n")
```

---

# 1. Bayesian Feature–Component Correlations

The frequentist analysis reports Pearson *r* with a t-test p-value. Here we replace
p-values with **Bayes Factors** (BF₁₀) from `BayesFactor::correlationBF()`, which
uses a JZS prior (Cauchy with scale = 0.333 on ρ). BF₁₀ > 10 is conventionally
"strong" evidence for H₁; BF₁₀ < 1/10 favours H₀.

For each BF we also extract the posterior median and 95% credible interval on ρ
from 10,000 posterior samples.

```{r bayes-correlations}
components  <- paste0("C", 1:15)
targets     <- c(components, "preference")
delta_cols  <- names(delta_df)[startsWith(names(delta_df), "delta_")]

run_bf_cor <- function(x_vec, y_vec, min_n = 100) {
  keep <- complete.cases(x_vec, y_vec)
  x <- x_vec[keep]; y <- y_vec[keep]
  if (length(x) < min_n) return(NULL)

  bf_obj  <- correlationBF(y = y, x = x, nullInterval = NULL)
  bf10    <- extractBF(bf_obj)$bf

  samps   <- posterior(bf_obj, iterations = 10000)
  rho_samp <- samps[, "rho"]
  ci      <- quantile(rho_samp, probs = c(0.025, 0.5, 0.975))

  tibble(
    n          = length(x),
    r_pearson  = cor(x, y, method = "pearson"),
    BF10       = bf10,
    rho_median = ci[2],
    rho_lo95   = ci[1],
    rho_hi95   = ci[3]
  )
}

bf_results <- list()

for (d in delta_cols) {
  feat  <- sub("^delta_", "", d)
  x_vec <- delta_df[[d]]
  for (tgt in targets) {
    y_vec  <- delta_df[[tgt]]
    result <- run_bf_cor(x_vec, y_vec)
    if (!is.null(result)) {
      bf_results[[length(bf_results) + 1]] <-
        bind_cols(tibble(feature = feat, target = tgt), result)
    }
  }
}

bf_df <- bind_rows(bf_results)
```

## 1.1 BF₁₀ heatmap (feature × component)

```{r bf-heatmap, fig.width=12, fig.height=4}
bf_comp <- bf_df %>%
  filter(target %in% components) %>%
  mutate(
    log_bf = log10(pmax(BF10, 1e-3)),   # clip for display
    target = factor(target, levels = components)
  )

ggplot(bf_comp, aes(x = target, y = feature, fill = log_bf)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = ifelse(abs(log_bf) > 1,
                               sprintf("%.1f", log_bf), "")),
            size = 3) +
  scale_fill_gradient2(
    low = "steelblue", mid = "white", high = "firebrick",
    midpoint = 0,
    name = "log₁₀(BF₁₀)"
  ) +
  labs(
    title = "Bayes Factors: feature × PLS component",
    x = "Component", y = "Feature",
    caption = "Cell values shown where |log₁₀(BF₁₀)| > 1 (BF > 10 or < 0.1)"
  ) +
  theme_minimal(base_size = 13)
```

## 1.2 Full results table

```{r bf-table}
bf_df %>%
  mutate(
    BF10       = formatC(BF10,       format = "e", digits = 2),
    r_pearson  = round(r_pearson, 3),
    rho_median = round(rho_median, 3),
    CI_95      = sprintf("[%.3f, %.3f]", rho_lo95, rho_hi95)
  ) %>%
  select(feature, target, n, r_pearson, BF10, rho_median, CI_95) %>%
  arrange(feature, target) %>%
  kable(
    col.names = c("Feature", "Target", "n", "Pearson r",
                  "BF₁₀", "Post. median ρ", "95% CI"),
    align = "llrrrrl"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## 1.3 Feature–preference Bayes Factors

```{r bf-preference}
bf_df %>%
  filter(target == "preference") %>%
  arrange(desc(BF10)) %>%
  mutate(
    BF10 = formatC(BF10, format = "e", digits = 2),
    r_pearson = round(r_pearson, 3),
    CI_95 = sprintf("[%.3f, %.3f]", rho_lo95, rho_hi95)
  ) %>%
  select(feature, n, r_pearson, BF10, CI_95) %>%
  kable(
    col.names = c("Feature", "n", "Pearson r", "BF₁₀", "95% CI"),
    caption = "Bayesian correlations between Δfeature and model ordering preference"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

---

# 2. Mixed-Effects Regression: Novel Set

The cross-validation analyses treat each pair as an independent observation, but
pairs sharing a word are not independent — if *cat* tends to go first, all pairs
containing *cat* will show a similar bias. This inflates precision.

We fit a Bayesian mixed-effects model with **crossed random intercepts** for word1
and word2, which absorbs word-level ordering biases and gives unbiased estimates
of the fixed effects (PLS components).

$$\text{preference}_{ij} = \beta_0 + \sum_{k=1}^{15} \beta_k C_{kij} + u_i + v_j + \varepsilon_{ij}$$

where $u_i \sim \mathcal{N}(0, \sigma^2_{\text{w1}})$ and
$v_j \sim \mathcal{N}(0, \sigma^2_{\text{w2}})$ are word-level random intercepts.

## 2.1 Priors

```{r novel-priors}
# Weakly informative priors.
# PLS scores are z-scored, so beta ~ N(0,1) is reasonable.
# Random effect SDs: exponential(1) concentrates near 0 but has a long tail.
novel_priors <- c(
  prior(normal(0, 1),    class = b),
  prior(normal(0, 2),    class = Intercept),
  prior(exponential(1),  class = sd),
  prior(exponential(1),  class = sigma)
)
```

## 2.2 Fit (novel set)

> **Note:** This model has ~8,955 unique words, giving ~17,910 random effects.
> Expect ~30–60 min on a modern CPU with 4 chains. Results are cached.

```{r fit-novel, cache=FALSE}
comp_formula <- paste(paste0("C", 1:15), collapse = " + ")
novel_formula <- as.formula(
  paste("preference ~", comp_formula, "+ (1 | word1) + (1 | word2)")
)

fit_novel <- brm(
  formula  = novel_formula,
  data     = novel_df,
  prior    = novel_priors,
  family   = gaussian(),
  chains   = 4,
  iter     = 4000,
  warmup   = 2000,
  cores    = 4,
  seed     = 964,
  backend  = "cmdstanr",    # faster than rstan; fall back to "rstan" if needed
  file     = file.path(RESULTS_DIR, "brms_novel_fit")   # saves/reloads automatically
)
```

## 2.3 Model diagnostics

```{r novel-diagnostics, fig.width=8, fig.height=5}
# Rhat and ESS summary
diag <- summarise_draws(fit_novel, "rhat", "ess_bulk", "ess_tail") %>%
  filter(str_starts(variable, "b_") | str_starts(variable, "sd_") |
           variable == "sigma")

cat("Max Rhat (should be < 1.01):", round(max(diag$rhat, na.rm = TRUE), 4), "\n")
cat("Min ESS bulk (should be > 400):", round(min(diag$ess_bulk, na.rm = TRUE), 0), "\n")
check_diagnostics(fit_novel, paste(SLUG, "novel"), WARN_FILE)

# Posterior predictive check
pp_check(fit_novel, ndraws = 100) +
  labs(title = "Posterior predictive check — novel set",
       x = "Ordering preference", y = "Density") +
  theme_minimal()
```

## 2.4 Fixed effects

```{r novel-fixed-effects}
fe <- fixef(fit_novel) %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  filter(term != "Intercept") %>%
  mutate(across(where(is.numeric), ~round(.x, 4)))

fe %>%
  kable(
    col.names = c("Component", "Estimate", "Est. Error", "Q2.5", "Q97.5"),
    caption = "Fixed effects: PLS components predicting novel ordering preference"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r novel-fe-plot, fig.width=8, fig.height=5}
fe %>%
  mutate(term = factor(term, levels = rev(term))) %>%
  ggplot(aes(x = Estimate, xmin = Q2.5, xmax = Q97.5, y = term)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_linerange() +
  geom_point(size = 2) +
  labs(
    title = "Fixed effects with 95% credible intervals",
    x = "Coefficient estimate", y = NULL
  ) +
  theme_minimal(base_size = 13)
```

## 2.5 Random effects and ICC

The **intraclass correlation coefficient (ICC)** is the proportion of variance
attributable to word-level effects. A high ICC means ordering preferences are
largely lexically specific — knowing a word's tendency from one pair predicts
its tendency in others.

```{r novel-icc}
vc <- VarCorr(fit_novel, summary = TRUE)

sigma2_w1  <- vc$word1$sd[1, "Estimate"]^2
sigma2_w2  <- vc$word2$sd[1, "Estimate"]^2
sigma2_res <- vc$residual__$sd[1, "Estimate"]^2
sigma2_tot <- sigma2_w1 + sigma2_w2 + sigma2_res

icc_w1  <- sigma2_w1  / sigma2_tot
icc_w2  <- sigma2_w2  / sigma2_tot
icc_tot <- (sigma2_w1 + sigma2_w2) / sigma2_tot

cat(sprintf("σ²(word1)   = %.4f  ICC = %.3f\n", sigma2_w1, icc_w1))
cat(sprintf("σ²(word2)   = %.4f  ICC = %.3f\n", sigma2_w2, icc_w2))
cat(sprintf("σ²(residual)= %.4f\n", sigma2_res))
cat(sprintf("Total ICC (word-level) = %.3f\n", icc_tot))
cat(sprintf("\nInterpretation: %.1f%% of variance in ordering preference is attributable\n",
    icc_tot * 100))
cat("to word-level idiosyncratic effects rather than pair-level structure.\n")
```

## 2.6 Bayesian R²

```{r novel-r2}
r2_novel <- bayes_R2(fit_novel, summary = FALSE, ndraws = 500)
cat(sprintf("Bayesian R² (novel, with word random effects):\n"))
cat(sprintf("  Median = %.3f  [%.3f, %.3f]\n",
            median(r2_novel[, "R2"]),
            quantile(r2_novel[, "R2"], 0.025),
            quantile(r2_novel[, "R2"], 0.975)))
rm(fit_novel); gc()
```

---

# 3. Mixed-Effects Regression: Corpus (In-Sample)

The same model fit on the 212,946 attested corpus pairs. This gives the Bayesian
in-sample R² comparable to the novel transfer result.

> **Note:** With 212,946 pairs this will be substantially slower than the novel
> model. Consider running overnight or on a computing cluster.

```{r fit-corpus, cache=FALSE}
corpus_formula <- as.formula(
  paste("preference ~", comp_formula, "+ (1 | word1) + (1 | word2)")
)

# Corpus has many more unique words — increase adapt_delta to reduce divergences
corpus_ctrl <- list(adapt_delta = 0.95)

fit_corpus <- brm(
  formula   = corpus_formula,
  data      = corpus_df,
  prior     = novel_priors,
  family    = gaussian(),
  chains    = 4,
  iter      = 4000,
  warmup    = 2000,
  cores     = 4,
  seed      = 964,
  backend   = "cmdstanr",
  control   = corpus_ctrl,
  file      = file.path(RESULTS_DIR, "brms_corpus_fit")
)
```

```{r corpus-r2}
r2_corpus <- bayes_R2(fit_corpus, summary = FALSE, ndraws = 500)
cat(sprintf("Bayesian R² (corpus, in-sample, with word random effects):\n"))
cat(sprintf("  Median = %.3f  [%.3f, %.3f]\n",
            median(r2_corpus[, "R2"]),
            quantile(r2_corpus[, "R2"], 0.025),
            quantile(r2_corpus[, "R2"], 0.975)))
check_diagnostics(fit_corpus, paste(SLUG, "corpus"), WARN_FILE)
```

```{r corpus-icc}
vc_c <- VarCorr(fit_corpus, summary = TRUE)

s2_w1  <- vc_c$word1$sd[1, "Estimate"]^2
s2_w2  <- vc_c$word2$sd[1, "Estimate"]^2
s2_res <- vc_c$residual__$sd[1, "Estimate"]^2
s2_tot <- s2_w1 + s2_w2 + s2_res

cat(sprintf("Corpus ICC (word-level) = %.3f\n", (s2_w1 + s2_w2) / s2_tot))
rm(fit_corpus); gc()
```

---

# 4. Corpus Word-Level Cross-Validation

The corpus word-level CV mirrors the novel word-level CV: unique words are split
into 10 folds; a corpus pair is testable only when both words fall in the same
held-out fold (~10% of pairs). PLS is refit from scratch on each training fold.

This allows direct comparison of word-level generalization within the attested
corpus vs. within the novel set — testing whether the lexical specificity pattern
(larger models → worse word-level CV) holds in both datasets.

```{r load-corpus-wordcv}
corpus_wcv_summary <- read_csv(
  file.path(RESULTS_DIR, "corpus_wordcv_summary.csv"),
  show_col_types = FALSE
)
corpus_wcv_folds <- read_csv(
  file.path(RESULTS_DIR, "corpus_wordcv_fold_stats.csv"),
  show_col_types = FALSE
)
corpus_wcv_preds <- read_csv(
  file.path(RESULTS_DIR, "corpus_wordcv_predictions.csv"),
  show_col_types = FALSE
)

cat("Corpus word-level CV summary:\n")
print(corpus_wcv_summary)
```

## 4.1 Bayesian correlation: corpus word CV predictions vs. observed

```{r corpus-wcv-bayes-cor}
bf_corpus_wcv <- correlationBF(
  y = corpus_wcv_preds$preference,
  x = corpus_wcv_preds$cv_pred,
  nullInterval = NULL
)

bf10_cwcv <- extractBF(bf_corpus_wcv)$bf
samps_cwcv <- posterior(bf_corpus_wcv, iterations = 10000)
ci_cwcv    <- quantile(samps_cwcv[, "rho"], probs = c(0.025, 0.5, 0.975))

cat(sprintf("Corpus word-level CV\n"))
cat(sprintf("  BF10 = %.2e\n", bf10_cwcv))
cat(sprintf("  Posterior median r = %.3f  [%.3f, %.3f]\n",
            ci_cwcv[2], ci_cwcv[1], ci_cwcv[3]))
cat(sprintf("  r² = %.3f\n", corpus_wcv_summary$cv_r2))
```

## 4.2 Compare corpus vs. novel word-level CV

```{r wcv-comparison}
novel_wcv_summary <- read_csv(
  file.path(RESULTS_DIR, "novel_wordcv_summary.csv"),
  show_col_types = FALSE
)

tibble(
  Set          = c("Corpus word-level CV", "Novel word-level CV"),
  n_tested     = c(corpus_wcv_summary$n_tested, novel_wcv_summary$n_tested),
  n_total      = c(corpus_wcv_summary$n_total,  novel_wcv_summary$n_total),
  pct_testable = round(100 * n_tested / n_total, 1),
  cv_r         = c(corpus_wcv_summary$cv_r,  novel_wcv_summary$cv_r),
  cv_r2        = c(corpus_wcv_summary$cv_r2, novel_wcv_summary$cv_r2)
) %>%
  kable(
    col.names = c("Set", "n tested", "n total", "% testable", "CV r", "CV r²"),
    caption   = "Word-level CV: corpus vs. novel"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## 4.3 Per-fold r² comparison

```{r wcv-fold-plot, fig.width=8, fig.height=4}
bind_rows(
  corpus_wcv_folds %>% mutate(set = "Corpus"),
  read_csv(file.path(RESULTS_DIR, "novel_wordcv_fold_stats.csv"),
           show_col_types = FALSE) %>% mutate(set = "Novel")
) %>%
  ggplot(aes(x = fold, y = r2, colour = set, group = set)) +
  geom_line() +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    title = "Word-level CV r² per fold: corpus vs. novel",
    x = "Fold", y = "r²", colour = NULL
  ) +
  theme_minimal(base_size = 13)
```

---

# 5. Fixed-Effects-Only Models (Comparison)

To quantify how much the word random effects change the fixed-effect estimates,
we fit the same regression without random effects.

```{r fit-novel-fe-only, cache=FALSE}
fe_formula <- as.formula(paste("preference ~", comp_formula))

fit_novel_fe <- brm(
  formula = fe_formula,
  data    = novel_df,
  prior   = c(prior(normal(0, 1), class = b),
              prior(normal(0, 2), class = Intercept),
              prior(exponential(1), class = sigma)),
  family  = gaussian(),
  chains  = 4,
  iter    = 4000,
  warmup  = 2000,
  cores   = 4,
  seed    = 964,
  backend = "cmdstanr",
  file    = file.path(RESULTS_DIR, "brms_novel_fe_fit")
)
```

```{r compare-fe-re}
r2_fe   <- bayes_R2(fit_novel_fe, summary = FALSE, ndraws = 500)
r2_re   <- r2_novel   # computed above

cat("Novel set — fixed effects only:\n")
cat(sprintf("  R² = %.3f [%.3f, %.3f]\n",
            median(r2_fe[,"R2"]),
            quantile(r2_fe[,"R2"], 0.025),
            quantile(r2_fe[,"R2"], 0.975)))

cat("\nNovel set — with word random effects:\n")
cat(sprintf("  R² = %.3f [%.3f, %.3f]\n",
            median(r2_re[,"R2"]),
            quantile(r2_re[,"R2"], 0.025),
            quantile(r2_re[,"R2"], 0.975)))

cat("\nDifference in R² attributable to word random effects:",
    round(median(r2_re[,"R2"]) - median(r2_fe[,"R2"]), 3), "\n")
check_diagnostics(fit_novel_fe, paste(SLUG, "novel-fe"), WARN_FILE)
rm(fit_novel_fe); gc()
```

---

# 6. MLP Neural Baseline

The MLP is a two-layer network (`Linear(p, 15) → Tanh → Linear(15, 1)`) trained to
predict ordering preference from the diff-vector (or concatenated individual vectors),
with hidden dim = 15 matching PLS K. Results are loaded from the pre-computed CSVs
produced by `Scripts/lib/mlp_comparison.py`.

```{r mlp-load}
mlp_files <- list.files(RESULTS_DIR, pattern = "^mlp_.*\\.csv$",
                         full.names = TRUE)
mlp_files <- mlp_files[!grepl("_(preds|fold_stats|loss_curves)\\.csv$", mlp_files)]

if (length(mlp_files) == 0) {
  cat("No MLP result files found in", RESULTS_DIR,
      "— run mlp_comparison.py first.\n")
  mlp_df <- tibble()
} else {
  mlp_df <- map_dfr(mlp_files, function(f) {
    d <- read_csv(f, show_col_types = FALSE)
    # CV splits use mean_* column names; normalise to match non-CV splits
    if ("mean_test_r" %in% names(d)) {
      d <- d %>% rename(
        test_r   = mean_test_r,
        test_r2  = mean_test_r2,
        test_rho = mean_test_rho,
        train_r  = mean_train_r,
        n_train  = mean_n_train,
        n_test   = mean_n_test
      ) %>% mutate(train_r2 = round(train_r^2, 6))
    }
    d <- d %>% mutate(n_train = as.integer(round(n_train)),
                      n_test  = as.integer(round(n_test)))
    stem  <- tools::file_path_sans_ext(basename(f))
    parts <- str_split(stem, "_")[[1]]
    d$input <- parts[2]
    d$split <- paste(parts[3:length(parts)], collapse = "_")
    d
  })
  cat("MLP results loaded:", nrow(mlp_df), "rows\n")
  print(mlp_df)
}
```

## 6.1 MLP vs. PLS comparison table

```{r mlp-vs-pls}
if (nrow(mlp_df) > 0) {
  # PLS reference r² for each split type.
  # transfer: corpus PLS applied frozen to novel (computed directly).
  # pair_novel: 10-fold pair-level CV within novel.
  # word_novel: 10-fold word-level CV within novel.
  # word_strict: no direct PLS equivalent -> NA.
  lm_pls <- lm(reformulate(comp_formula, "preference"), data = corpus_df)
  pls_transfer_r2 <- cor(novel_df$preference,
                         predict(lm_pls, newdata = novel_df))^2

  pls_refs <- tribble(
    ~split,        ~pls_r2,
    "transfer",    pls_transfer_r2,
    "pair_novel",  read_csv(file.path(RESULTS_DIR, "novel_cv_summary.csv"),
                            show_col_types = FALSE) %>%
                   pull(cv_r2) %>% first(),
    "word_novel",  read_csv(file.path(RESULTS_DIR, "novel_wordcv_summary.csv"),
                            show_col_types = FALSE) %>%
                   pull(cv_r2) %>% first()
  )

  mlp_df %>%
    left_join(pls_refs, by = "split") %>%
    mutate(
      test_r2    = round(test_r2, 3),
      train_r2   = round(train_r2, 3),
      pls_r2     = round(pls_r2, 3),
      delta_pls  = round(test_r2 - pls_r2, 3)
    ) %>%
    select(input, split, n_train, n_test, train_r2, test_r2, pls_r2, delta_pls) %>%
    arrange(split, input) %>%
    kable(
      col.names = c("Input", "Split", "N train", "N test",
                    "Train r²", "Test r²", "PLS r²", "Δ vs PLS"),
      caption = paste("MLP vs. PLS comparison —", SLUG)
    ) %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
}
```

## 6.2 MLP predicted vs. observed

```{r mlp-pred-obs, fig.width=12, fig.height=6, warning=FALSE, message=FALSE}
mlp_pred_files <- list.files(RESULTS_DIR, pattern = "^mlp_.*_preds\\.csv$",
                              full.names = TRUE)

if (length(mlp_pred_files) > 0) {
  mlp_preds <- map_dfr(mlp_pred_files, function(f) {
    d <- read_csv(f, show_col_types = FALSE) %>%
      rename(predicted = mlp_pred)
    stem  <- tools::file_path_sans_ext(basename(f))
    # stem is mlp_{input}_{split}_preds — remove trailing _preds
    stem  <- str_remove(stem, "_preds$")
    parts <- str_split(stem, "_")[[1]]
    d$label <- paste0("MLP-", parts[2], " (", paste(parts[3:length(parts)],
                                                      collapse = "_"), ")")
    d
  })

  ggplot(mlp_preds %>% group_by(label) %>% slice_sample(n = 5000),
         aes(x = preference, y = predicted)) +
    geom_point(alpha = 0.15, size = 0.4, colour = "darkorange") +
    geom_smooth(method = "lm", colour = "firebrick", linewidth = 0.8, se = TRUE) +
    facet_wrap(~label, scales = "free", ncol = 3) +
    labs(
      title = "MLP: predicted vs. observed ordering preference",
      subtitle = paste("Model:", SLUG),
      x = "Actual preference (log-prob ratio)",
      y = "MLP predicted preference"
    ) +
    theme_minimal(base_size = 12) +
    theme(strip.text = element_text(face = "bold"))
} else {
  cat("No MLP prediction files found — run mlp_comparison.py first.\n")
}
```

---

# 7. Summary Table

```{r summary-table}
tibble(
  Model      = SLUG,
  Set        = c("Corpus (in-sample)", "Novel (frozen)", "Novel (frozen)"),
  RE         = c("word1 + word2", "word1 + word2", "none"),
  R2_median  = c(
    round(median(r2_corpus[,"R2"]), 3),
    round(median(r2_novel[,"R2"]),  3),
    round(median(r2_fe[,"R2"]),     3)
  ),
  R2_lo = c(
    round(quantile(r2_corpus[,"R2"], 0.025), 3),
    round(quantile(r2_novel[,"R2"],  0.025), 3),
    round(quantile(r2_fe[,"R2"],     0.025), 3)
  ),
  R2_hi = c(
    round(quantile(r2_corpus[,"R2"], 0.975), 3),
    round(quantile(r2_novel[,"R2"],  0.975), 3),
    round(quantile(r2_fe[,"R2"],     0.975), 3)
  )
) %>%
  mutate(CI_95 = sprintf("[%.3f, %.3f]", R2_lo, R2_hi)) %>%
  select(Model, Set, RE, R2_median, CI_95) %>%
  kable(
    col.names = c("Model", "Set", "Random effects", "R² (median)", "95% CI"),
    caption = "Bayesian R² summary — all values with 95% credible intervals"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

---

# 8. Predicted vs. Observed Preference Plots

These scatter plots address the concern that r² ≈ .14 could be an artifact of
preferences clustering near 50/50. If preferences had near-zero variance, both
axes would be compressed near 0. Visible spread on the x-axis (actual preference)
alongside a positive correlation confirms that the structure is real.

```{r pred-vs-obs, fig.width=12, fig.height=8, warning=FALSE, message=FALSE}
corpus_pls  <- read_csv(CORPUS_F, show_col_types = FALSE)
novel_pls   <- read_csv(NOVEL_F,  show_col_types = FALSE)

comp_cols <- paste0("C", 1:15)

# Corpus in-sample predictions from PLS regression
lm_corpus   <- lm(reformulate(comp_cols, "preference"), data = corpus_pls)
corpus_pls$predicted <- predict(lm_corpus)

# Novel frozen predictions
novel_pls$predicted  <- predict(lm_corpus, newdata = novel_pls)

# CV prediction files
novel_pair_cv  <- read_csv(file.path(RESULTS_DIR, "novel_cv_predictions.csv"),
                           show_col_types = FALSE)
novel_word_cv  <- read_csv(file.path(RESULTS_DIR, "novel_wordcv_predictions.csv"),
                           show_col_types = FALSE)
corpus_word_cv <- read_csv(file.path(RESULTS_DIR, "corpus_wordcv_predictions.csv"),
                           show_col_types = FALSE)

plot_data <- bind_rows(
  corpus_pls     %>% select(preference, predicted) %>% mutate(condition = "Corpus (in-sample)"),
  novel_pls      %>% select(preference, predicted) %>% mutate(condition = "Novel (frozen)"),
  novel_pair_cv  %>% select(preference, predicted = cv_pred) %>% mutate(condition = "Novel pair-level CV"),
  novel_word_cv  %>% select(preference, predicted = cv_pred) %>% mutate(condition = "Novel word-level CV"),
  corpus_word_cv %>% select(preference, predicted = cv_pred) %>% mutate(condition = "Corpus word-level CV")
) %>%
  mutate(condition = factor(condition, levels = c(
    "Corpus (in-sample)", "Novel (frozen)",
    "Novel pair-level CV", "Novel word-level CV", "Corpus word-level CV"
  )))

ggplot(plot_data %>% group_by(condition) %>% slice_sample(n = 5000),
       aes(x = preference, y = predicted)) +
  geom_point(alpha = 0.15, size = 0.4, colour = "steelblue") +
  geom_smooth(method = "lm", colour = "firebrick", linewidth = 0.8, se = TRUE) +
  facet_wrap(~condition, scales = "free", ncol = 3) +
  labs(
    title = "Predicted vs. observed ordering preference",
    subtitle = paste("Model:", SLUG),
    x = "Actual preference (log-prob ratio)",
    y = "Predicted preference"
  ) +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold"))
```

## 7.2 Binned means

Bins actual preferences into 20 equal-width intervals and plots the mean
prediction per bin. A flat line near zero would indicate the correlation is
driven only by outliers; a monotonic line across the full range confirms
systematic structure throughout the distribution.

```{r binned-pred-vs-obs, fig.width=12, fig.height=8, warning=FALSE, message=FALSE}
binned <- plot_data %>%
  group_by(condition) %>%
  mutate(bin = cut(preference, breaks = 20)) %>%
  group_by(condition, bin) %>%
  summarise(
    mean_actual    = mean(preference),
    mean_predicted = mean(predicted),
    n              = n(),
    se             = sd(predicted) / sqrt(n),
    .groups = "drop"
  ) %>%
  filter(n >= 10)

ggplot(binned, aes(x = mean_actual, y = mean_predicted)) +
  geom_ribbon(aes(ymin = mean_predicted - 1.96 * se,
                  ymax = mean_predicted + 1.96 * se),
              alpha = 0.2, fill = "steelblue") +
  geom_line(colour = "firebrick", linewidth = 0.9) +
  geom_point(aes(size = n), colour = "firebrick", alpha = 0.7) +
  scale_size_continuous(range = c(1, 4), guide = "none") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  facet_wrap(~condition, scales = "free", ncol = 3) +
  labs(
    title = "Mean predicted preference by bin of actual preference",
    subtitle = paste("Model:", SLUG, "| Each point = mean of ~N pairs; ribbon = 95% CI"),
    x = "Actual preference (log-prob ratio)",
    y = "Mean predicted preference"
  ) +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold"))
```

---

# Session Info

```{r session-info}
sessionInfo()
```
