---
title: "Stored or Computed? Separating Compositional from Holistic Representations of Binomials in Language Models"
acl-mode: review
format:
  acl-pdf:
    keep-tex: true
    cite-method: natbib
  html: default
editor: visual
author:
  - name:
      literal: Zachary Nicholas Houghton
    email: znh@uoregon.edu
    affiliations:
      - name: University of Oregon
      - name: Vail Systems, Inc.

abstract: |
  Humans balance abstract and item-specific knowledge when ordering binomials: abstract constraints shape the ordering preferences for infrequent binomials, while idiosyncratic knowledge drives the ordering preferences for frequent ones. Evaluating 16 language models across five families, we uncover a parallel trade-off. Probing hidden-state representations shows that language models do learn abstract preferences, recovering the ordering of binomials built from words they never saw together. But a binomial's own attested order in the training corpus predicts a model's preference over and above anything those abstract preferences supply, and predicts it more strongly the more often the model encountered it. To locate that item-specific component, we zero out cross-word attention, forcing the representation into a strictly additive function of the two words. The intervention removes roughly half of it while the abstract component survives undiminished, showing that what is stored requires the two words to be encoded together. Tracking training checkpoints shows that abstract preferences are present early while the item-specific component is absent at first and accumulates with exposure, its onset governed by training tokens rather than model scale.

bibliography: references.bib
tbl-cap-location: bottom
header-includes:
  - \usepackage{float}
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)
library(kableExtra)
library(scales)

RES <- "../Results"

# Keep large intermediates off the system drive. The .xz prediction files
# decompress through vroom's temp storage at roughly 277 MB apiece, which has
# exhausted C: during rendering. The project's .Renviron sets this too, but R
# reads .Renviron from the working directory and Quarto renders from Writeup/,
# so it is set here as well. The path must contain no spaces: R rejects a
# temp directory with a space in it.
local({
  tmp <- Sys.getenv("VROOM_TEMP_PATH")
  if (!nzchar(tmp)) {
    tmp <- "D:/rtmp"
    if (dir.exists(dirname(tmp))) {
      dir.create(tmp, showWarnings = FALSE, recursive = TRUE)
      Sys.setenv(VROOM_TEMP_PATH = tmp)
    }
  }
})

# ── Slug → label map ──────────────────────────────────────────────────────────
all_slugs <- c(
  "znhoughton_opt-babylm-125m-20eps-seed964" = "BabyLM-125M",
  "znhoughton_opt-babylm-350m-20eps-seed964" = "BabyLM-350M",
  "znhoughton_opt-babylm-1_3b-20eps-seed964" = "BabyLM-1.3B",
  "EleutherAI_pythia-160m"                   = "Pythia-160M",
  "EleutherAI_pythia-410m"                   = "Pythia-410M",
  "EleutherAI_pythia-1b"                     = "Pythia-1B",
  "gpt2"                                     = "GPT-2",
  "gpt2-medium"                              = "GPT-2-M",
  "gpt2-large"                               = "GPT-2-L",
  "gpt2-xl"                                  = "GPT-2-XL",
  "allenai_OLMo-1B-hf"                       = "OLMo-1B",
  "allenai_OLMo-7B-hf"                       = "OLMo-7B",
  "allenai_OLMo-2-0425-1B"                   = "OLMo-2-1B",
  "allenai_OLMo-2-1124-7B"                   = "OLMo-2-7B",
  "meta-llama_Llama-3.2-1B"                  = "Llama-1.3B",
  "meta-llama_Meta-Llama-3-8B"               = "Llama-3-8B"
)

family_map <- c(
  "BabyLM-125M" = "BabyLM", "BabyLM-350M" = "BabyLM", "BabyLM-1.3B" = "BabyLM",
  "Pythia-160M" = "Pythia", "Pythia-410M"  = "Pythia", "Pythia-1B"   = "Pythia",
  "GPT-2"       = "GPT-2",  "GPT-2-M"      = "GPT-2",  "GPT-2-L"     = "GPT-2",
  "GPT-2-XL"   = "GPT-2",
  "OLMo-1B"    = "OLMo",   "OLMo-7B"      = "OLMo",   "OLMo-2-1B"   = "OLMo",
  "OLMo-2-7B"  = "OLMo",
  "Llama-1.3B" = "Llama",  "Llama-3-8B"   = "Llama"
)

# ── Training-scale metadata ───────────────────────────────────────────────────
model_meta <- tribble(
  ~label,         ~params_M,  ~training_tokens_B,
  "BabyLM-125M",  125,         0.15,
  "BabyLM-350M",  350,         0.15,
  "BabyLM-1.3B",  1300,        0.15,
  "GPT-2",        117,         8.0,
  "GPT-2-M",      345,         8.0,
  "GPT-2-L",      762,         8.0,
  "GPT-2-XL",     1500,        8.0,
  "Pythia-160M",  160,         300.0,
  "Pythia-410M",  410,         300.0,
  "Pythia-1B",    1000,        300.0,
  "OLMo-1B",      1000,        3000.0,
  "OLMo-7B",      7000,        2500.0,
  "OLMo-2-1B",    1000,        4000.0,
  "OLMo-2-7B",    7000,        4000.0,
  "Llama-1.3B",   1200,        9000.0,
  "Llama-3-8B",   8000,        15000.0
)

# ── Load MLP R² from per-model CSVs ──────────────────────────────────────────
load_mlp <- function(slug, label, is_control = FALSE) {
  fname <- if (is_control) "by_layer_mlp_control.csv" else "by_layer_mlp.csv"
  path  <- file.path(RES, slug, fname)
  if (!file.exists(path)) return(NULL)
  read_csv(path, show_col_types = FALSE) |>
    mutate(model = label, family = family_map[label], is_control = is_control)
}

all_r2 <- map_dfr(names(all_slugs), function(slug) {
  bind_rows(
    load_mlp(slug, all_slugs[[slug]], FALSE),
    load_mlp(slug, all_slugs[[slug]], TRUE)
  )
}) |>
  filter(split %in% c("pair_novel", "word_novel")) |>
  group_by(model, family, condition, mode, split, is_control) |>
  filter(layer == max(layer)) |>
  ungroup()

# ── Checkpoint slug registries ────────────────────────────────────────────────
babylm_ckpt_registry <- bind_rows(
  tibble(slug  = paste0("znhoughton_opt-babylm-125m-20eps-seed964_step",
                        c(24, 48, 144, 384, 912, 2280)),
         label = "BabyLM-125M",
         step  = c(24, 48, 144, 384, 912, 2280),
         tokens_M = c(24, 48, 144, 384, 912, 2280) * 0.8192),  # 1024 * 400 * 2 / 1e6
  tibble(slug  = paste0("znhoughton_opt-babylm-350m-20eps-seed964_step",
                        c(48, 96, 288, 768, 1824, 4560)),
         label = "BabyLM-350M",
         step  = c(48, 96, 288, 768, 1824, 4560),
         tokens_M = c(48, 96, 288, 768, 1824, 4560) * 0.4096),  # 1024 * 200 * 2 / 1e6
  tibble(slug  = paste0("znhoughton_opt-babylm-1.3b-20eps-seed964_step",
                        c(97, 194, 582, 1455, 3686, 9021)),
         label = "BabyLM-1.3B",
         step  = c(97, 194, 582, 1455, 3686, 9021),
         tokens_M = c(97, 194, 582, 1455, 3686, 9021) * 0.2048)  # 1024 * 100 * 2 / 1e6
)

pythia_ckpt_registry <- bind_rows(
  tibble(slug  = paste0("EleutherAI_pythia-160m_step", c(16, 32, 64, 256, 512, 1000)),
         label = "Pythia-160M",
         step  = c(16, 32, 64, 256, 512, 1000),
         tokens_M = c(16, 32, 64, 256, 512, 1000) * 2.097152),
  tibble(slug  = paste0("EleutherAI_pythia-410m_step", c(16, 32, 64, 256, 512, 1000)),
         label = "Pythia-410M",
         step  = c(16, 32, 64, 256, 512, 1000),
         tokens_M = c(16, 32, 64, 256, 512, 1000) * 2.097152),
  tibble(slug  = paste0("EleutherAI_pythia-1b_step",   c(16, 32, 64, 256, 512, 1000)),
         label = "Pythia-1B",
         step  = c(16, 32, 64, 256, 512, 1000),
         tokens_M = c(16, 32, 64, 256, 512, 1000) * 2.097152),
)

# ── UO-brand colour palettes ──────────────────────────────────────────────────
# Oregon Green: #154733 / #007030  UO Gold: #F0B323  Amber: #875E00  UO Blue: #005581
family_colors <- c(
  "BabyLM" = "#154733",
  "GPT-2"  = "#F0B323",
  "Pythia" = "#007030",
  "OLMo"   = "#875E00",
  "Llama"  = "#005581"
)

babylm_size_colors <- c(
  "BabyLM-125M" = "#F0B323",
  "BabyLM-350M" = "#007030",
  "BabyLM-1.3B" = "#154733"
)

# the two predictors of @eq-relfreq: abstract component vs the corpus's own
# attested order for that binomial. Figures use prose labels; tables keep the
# variable names, which is where a reader looking up a coefficient expects them.
term_labels <- c("y_pred"   = "Abstract preference (probe)",
                 "rel_freq" = "Relative frequency")
# UO green and UO gold, the same two the checkpoint figure uses for size
term_colors <- c("Abstract preference (probe)" = "#154733",
                 "Relative frequency"          = "#F0B323")
relabel_term <- function(x) {
  factor(unname(term_labels[as.character(x)]), levels = unname(term_labels))
}

# ── Factor order for model labels ──────────────────────────────────────────────
model_order <- c(
  "BabyLM-125M", "BabyLM-350M", "BabyLM-1.3B",
  "GPT-2", "GPT-2-M", "GPT-2-L", "GPT-2-XL",
  "Pythia-160M", "Pythia-410M", "Pythia-1B",
  "OLMo-1B", "OLMo-7B", "OLMo-2-1B", "OLMo-2-7B",
  "Llama-1.3B", "Llama-3-8B"
)

# ascending params within family (for dot-plot row order)
dumbbell_order <- c(
  "BabyLM-125M", "BabyLM-350M", "BabyLM-1.3B",
  "GPT-2", "GPT-2-M", "GPT-2-L", "GPT-2-XL",
  "Pythia-160M", "Pythia-410M", "Pythia-1B",
  "OLMo-1B", "OLMo-2-1B", "OLMo-7B", "OLMo-2-7B",
  "Llama-1.3B", "Llama-3-8B"
)

condition_labels <- c(
  "default"     = "Full context",
  "attn_zeroed" = "Attention-zeroed"
)

# UO green and gold, as elsewhere in the paper.
condition_colors <- c("Full context" = "#154733", "Attention-zeroed" = "#F0B323")

recode_condition <- function(x) {
  factor(recode(x, !!!condition_labels),
         levels = c("Full context", "Attention-zeroed"))
}

mode_labels <- c(
  "mean_pooled" = "Binomial-level",
  "words_only"  = "Word-level"
)

theme_paper <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      text             = element_text(color = "black"),
      axis.text        = element_text(color = "black", size = rel(0.9)),
      axis.title       = element_text(color = "black", size = rel(1.0)),
      legend.text      = element_text(color = "black", size = rel(0.9)),
      legend.title     = element_text(color = "black", size = rel(1.0)),
      strip.text       = element_text(color = "black", face = "bold", size = rel(1.0)),
      panel.grid.minor = element_blank(),
      axis.ticks       = element_line(color = "black"),
      axis.ticks.length = unit(3, "pt")
    )
}

# collapse_rows() emits \nopagebreak after EVERY row so that a collapsed group
# is never split. In a longtable that makes the whole table effectively
# unbreakable: LaTeX cannot begin it on a partly filled page, so it is pushed
# to the next page and leaves the remainder of the previous one blank. This
# strips \nopagebreak only where it follows a group-final row (the row carrying
# the \multirow label) at the OUTERMOST level, whose line begins with
# \multirow rather than being nested after an &. Breaking inside an outer
# group splits its \multirow label across pages and renders it on top of the
# continuation header, so only whole top-level groups may be separated.
allow_group_breaks <- function(x) {
  y <- gsub("(?m)^(\\\\multirow\\{-[0-9]+\\}[^\n]*\n)\\\\nopagebreak\n",
            "\\1", x, perl = TRUE)
  attributes(y) <- attributes(x)
  y
}
```

# Introduction

Language users are constantly faced with a trade-off between holistically stored, item-specific knowledge and abstract, generative knowledge. For example, the meaning of a phrase often depends on whether it is composed from its individual components or retrieved as a whole. Idioms, such as *kicked the bucket*, are the clearest form of this: the compositional meaning refers to the action of kicking a bucket whereas the idiomatic meaning refers to someone passing away. However, this trade-off isn't limited to meaning. More generally, the representation of a multi-word expression may be composed from the representations of its individual pieces, or the expression may be stored and retrieved holistically. How, and crucially when, this trade-off emerges in learning has been a question that linguists have been asking for nearly a century [@berkoChildsLearningEnglish1958; @morgan2015; @morgan2024productiveknowledgeitemspecific; @houghton2023doespredictabilitydrive; @houghton2024frequencydependentpreferenceextremity; @houghton2025boltsnutslanguage; @houghton2025multiwordrepresentationsminds; @houghton2025role; @houghton2026holisticstorageverbupphrases; @houghton2026exemplarsdisguisepure; @stembergerAreInflectedForms2004; @kapatsinskiFrequencyEmergencePrefabs2009].

## Abstract Preferences and Holistic Storage

Historically, these questions have been largely addressed at the behavioral level. @berkoChildsLearningEnglish1958 famously demonstrated that children could produce the correct plural of nouns that they had never heard before rather early in life; because the plural form was never available to be stored, it must instead have been generated using abstract knowledge of the language. Evidence for abstract knowledge of this kind has been relatively uncontroversial, and it does not rest on child language alone. The priming literature, for instance, has demonstrated that sentences sharing syntactic structure [e.g., @bock1986syntacticpersistencelanguage] or semantic structure [e.g., @meyer1971facilitationrecognizingpairs] facilitate the processing of later sentences sharing those features, which requires that the shared structure be represented independently of the particular words that realize it.

Evidence for holistic storage, on the other hand, has been somewhat more controversial. Nonetheless, evidence of holistic storage has been accumulating over the last few decades [@stembergerAreInflectedForms2004; @kapatsinskiFrequencyEmergencePrefabs2009; @bybeeEffectUsageDegrees1999; @morgan2016]. Perhaps the most famous example comes from @bybeeEffectUsageDegrees1999. In their seminal study, they demonstrated that the extent to which the word *don't* is reduced is determined by the frequency of the phrase it occurs in. *I don't know* can be reduced to a single nasalized vowel, while *don't* in less frequent phrases, such as *I don't walk*, is unable to be reduced as much. The context-specificity of this reduction demonstrates that the representation of certain high-frequency phrases cannot be reduced to the simple composition of the representations of the individual words.

Binomial expressions (e.g., *cats and dogs*) are notable for their relationship between abstract preferences and holistic storage: they show evidence of being driven by both kinds of knowledge. Speakers have strong intuitions about the order of the two words: *salt and pepper* sounds far more natural than *pepper and salt*, even though there's nothing ungrammatical about the latter. For lower frequency binomials, @morgan2016 demonstrated that these preferences are driven by abstract constraints such as the length of a word, the semantic content of a word, etc [@benor2006]. However, that same study also showed that those constraints don't account for the ordering preferences of high-frequency binomials like *salt and pepper*. For those binomials, the preferences can't be predicted solely from the properties of the individual words, suggesting that the binomial is instead stored *holistically*. Binomial ordering preferences therefore show sensitivity to both abstract preferences and holistic storage depending on their frequency, which makes them a particularly useful case for studying the trade-off between the two.

## Abstract Preferences in Language Models

Language models offer a way to study this trade-off that behavioral work cannot. When an expression stops being formed compositionally and starts being stored holistically is fundamentally a question about experience, but a human speaker's history with any particular expression can only be estimated reasonably using corpus frequency. In a language model, that history is the training corpus itself, and it can be counted exactly. The representations are also directly inspectable, and the computation that produces them can be intervened on. The same is not true for humans, at least not without throwing away our ethics.

The evidence for abstract preferences and holistic storage in language models has been much more mixed than in the human behavioral literature. Models clearly generalize beyond what they have seen: @misraLanguageModelsLearn2024 show that language models can produce the rare *a beautiful five days* construction even when it is withheld from training, generalizing to it from related constructions that are better attested, and @yaoBothDirectIndirect2025 find that dative alternation preferences are shaped by abstract constraints rather than direct experience alone. At the same time, models retain a great deal item-specifically. @mccoy2023howmuchlanguage find that while model-generated text is often as structurally novel as human text, the same models also reproduce passages of over a thousand words verbatim from their training data. As such, it is clear that both abstract preferences and holistic storage are clearly at play within language models, however what remains unclear is how these types of knowledge trade off and when they emerge during training.

Perhaps most related to the present study, language models' ordering preferences for binomials appear not to be driven by abstract preferences at all [@houghton2025role]. @houghton2025role examined whether language models showed the same abstract preferences that humans show in the ordering of binomials, and found that the models' ordering preferences were driven exclusively by frequency statistics, with no evidence of an effect of the abstract preferences that affect human ordering preferences. However, that finding faces limitations along two main dimensions. First, the study can only claim to find no evidence that the models' ordering preferences were driven by *human* ordering preferences; it may very well be that their preferences are driven by abstract preferences that simply differ from those humans show. Second, the examination was conducted on model behavior, in the form of output probabilities, rather than over the representations that produce it. A model could in principle derive an ordering from abstract properties of the binomial and still produce output statistics that look frequency-driven.

## Present Study

The present study thus asks whether binomial ordering preferences in language models are driven by abstract structure or if they are simply holistically stored. We make two changes to the design of @houghton2025role. First, rather than testing model behavior against a fixed inventory of human constraints, we ask whether *any* systematic structure is recoverable at all, by training a probe to predict a model's ordering preference from its hidden representations and evaluating it on binomials that the probe was never trained on. Second, we separate the two components architecturally rather than statistically. Zeroing out cross-word attention forces the model's preference to be purely compositional: a strictly additive function of the individual words evaluated separately. Interventions on attention have been used to trace factual recall and to isolate circuits [@geva-etal-2023-dissecting; @wang2023interpretability], but to our knowledge this is the first use of one to separate holistic storage from abstract preferences. We apply this to 16 models across five families, to models trained on an amount of data comparable to humans as well as on models trained on trillions of tokens, and to training checkpoints for two families of models matched roughly in terms of numbers of parameters but differing in the size of their training corpus.

Our main contributions are:

-   **An architectural intervention separating holistic storage from abstract preferences.** Zeroing out cross-word attention between the individual words isolates a model's abstract preferences from its binomial-specific representations.

-   **Empirical evidence for both abstract preferences and frequency-driven holistic storage.** We show that ordering preferences are recoverable from a model's representations and generalize to binomials composed of words the probe never saw, establishing that they are driven by abstract preferences. At the same time, the order a binomial's own training corpus attests for it predicts a model's preference over and above anything those abstract preferences supply, and does so more strongly the more often the model encountered it, showing that frequent items are additionally stored. 
-   **Demonstrating that exposure, not scale, drives the shift.** Across a tenfold range in parameters, the stored component becomes detectable at nearly the same point in training: between 298M and 315M tokens for the three BabyLM models. What governs the transition is how much text a model has read rather than its representational capacity.

# Methods {#sec-methods}

All three experiments apply the same procedure and differ only in which models and checkpoints it is applied to; Appendix\ \ref{sec-appendix-models} lists all sixteen with their parameter counts and training-token estimates. This section describes that shared procedure: the datasets, the probes, and the analyses.

```{r}
#| label: exp2-novel-pile-stats
#| include: false

# Counts are from DOLMA, not the Pile. Dolma is OLMo's training corpus; it stands
# in here as a large-web-corpus proxy for the other families, whose exact training
# data is either different (Pile, WebText) or undisclosed (Llama).
novel_dolma_freqs <- read_csv(
  file.path(RES, "..", "Results", "novel_binomials_infinigram_dolma.csv"),
  show_col_types = FALSE
)
dolma_nonzero     <- novel_dolma_freqs$freq_total[novel_dolma_freqs$freq_total > 0]
dolma_pct_nonzero <- round(mean(novel_dolma_freqs$freq_total > 0) * 100, 1)
dolma_q025 <- round(quantile(dolma_nonzero, 0.025))
dolma_q975 <- round(quantile(dolma_nonzero, 0.975))
```

## Datasets

In order to examine the ordering preferences for binomials, we created two binomial corpora: a novel set and an attested set. The **novel set** comprises approximately 340,000 binomials from the Wikitext corpus [@merity2016pointer] selected to be absent from BabyLM's training corpus. The **attested corpus set** comprises approximately 49,000 binomials from the BabyLM corpus [@warstadt-etal-2023-findings].

The novel set is genuinely novel only for the BabyLM models; for the large-scale models it consists of infrequent attested binomials rather than unseen ones. We retain the label *novel* throughout, and return to what it licenses in the Limitations.

Both sets were extracted automatically, using spaCy [@spacy2] part-of-speech tags to find candidate sequences and the Berkeley Neural Parser [@kitaev-klein-2018-constituency; @kitaev-etal-2019-multilingual] to confirm that the two words form a minimal coordinate phrase; Appendix\ \ref{sec-appendix-extraction} gives the full criteria.

For each corpus binomial and each model, we also calculate the ordering preference, which is operationalized as the model's own log-probability ratio: $y_{\text{true}} = \log P(W_1\text{ and }W_2) - \log P(W_2\text{ and }W_1)$.

For the developmental analysis in Experiment 3, we probe six checkpoints each from the BabyLM and Pythia models. BabyLM checkpoints were sampled at approximately even intervals across training; Pythia checkpoints were instead chosen so that each model had seen roughly the same number of tokens as the corresponding BabyLM. These span roughly 20M to 2B effective training tokens.

## Representation Extraction

For each model we extract final-layer hidden states, presenting each binomial in both orderings (*W*~1~ and *W*~2~; *W*~2~ and *W*~1~) within the same sentence frame. Each binomial is represented in two ways:

-   **Binomial-level**: $[\bar{h}(W_1\text{ and }W_2); \bar{h}(W_2\text{ and }W_1)]$, where $\bar{h}$ denotes the mean of token representations across all tokens in the binomial in each ordering.
-   **Word-level**: $[h(W_1); h(W_2)]$, where $h(W_k)$ is the hidden state of word $W_k$ taken from the sentence in which it appears in first position.

## Attention Conditions

We additionally probe representations under two attention conditions:

-   **Full context**: standard forward pass with full cross-token attention. Each word's representation reflects contextual influences from all other (previous) tokens in the sentence.
-   **Attention-zeroed**: attention from the second word and the conjunction to the first word is blocked by setting the corresponding attention logits to $-\infty$ before the softmax, at every transformer layer, so no cross-word information reaches $W_2$ or the conjunction at any depth (Appendix\ \ref{sec-appendix-implementation}). This condition tests whether the second word's representation is specifically shaped by direct cross-word attention to the first word.

This gives three representation types per model: binomial-level under each attention condition, and full-context word-level.

Because $y_{true}$ is a difference of log-probabilities over the two orderings, and because causal masking already prevents $W_1$ from attending to $W_2$, blocking attention from $W_2$ and the conjunction to $W_1$ leaves the model blind to the other word. The resulting preference is necessarily of the form $g(W_1) - g(W_2)$: an additive function of the two words evaluated separately, with no interaction term. The attention-zeroed model therefore reports not a degraded preference but the preference it would hold given only per-word information.

## Probes

For each representation type, we train a two-layer MLP classifier (hidden dimension 128, ReLU activation) to predict binomial ordering preferences from the extracted representations. The probe takes the concatenated representations for a binomial and outputs a single continuous value $\hat{y}$, trained under mean-squared-error loss against $y_{true}$, with antisymmetric augmentation so that a swapped ordering carries the negated label. Optimizer settings, early stopping, and the augmentation rate are given in Appendix\ \ref{sec-appendix-implementation}. We report cross-validated R² (the squared Pearson correlation between $\hat{y}$ and $y_{true}$ on held-out folds), averaged across 10 folds.[^1]

[^1]: The results below do not depend on the MLP's capacity. A ridge regression on the same antisymmetric features, using the same folds and a single tuned hyperparameter, recovers most of the MLP's performance and orders the models the same way; on the word-held-out split the two agree most closely. Full comparison in Appendix\ \ref{sec-appendix-linear}.

We evaluate generalization under two cross-validation splits:

-   **Binomial-held-out**: binomials are held out as units; each binomial in the test set is excluded from the training set, however overlapping words are allowed (e.g., *radios and computers* is not excluded from the training even if *radios and televisions* is in the test set).
-   **Word-held-out**: the vocabulary is partitioned into ten folds; the test set for a fold is the binomials whose *both* words fall in that fold, and the training set is the binomials whose *neither* word does. Binomials with one word on each side are discarded for that fold. This tests generalization to completely novel lexical items.

Both splits are applied to the novel set; the attested set is not cross-validated, and enters only as the transfer target described below. Word-held-out yields an order of magnitude fewer test items per fold, and its R² is correspondingly more variable, which is why we report fold standard deviations throughout (Appendix\ \ref{sec-appendix-cv}).

Finally, following @hewitt-liang-2019-designing, we train a matched control probe on randomly permuted labels under identical settings; near-zero control R² indicates the probe is recovering linguistic structure rather than label artifacts or probe capacity.

## Analyses {#sec-methods-analyses}

We report two measurements. The first is the probe's cross-validated $R^2$ under the two splits above, of which word-held-out is the stricter: binomial-held-out $R^2$ may be inflated by lexical preferences, such as a general tendency for *bread* to come first.

The second measurement uses the probe rather than evaluating it. A probe trained on the full novel set is applied to the attested set. Since that probe has seen only binomials the model encountered rarely or not at all, its prediction $\hat{y}$ for an attested binomial estimates what generalizable knowledge alone would predict, and cannot reflect the probe having learned that binomial.

Two things could govern the model's preference for an attested binomial: the abstract component $\hat{y}$, and the order the corpus itself attests for that particular pair. The latter is item-specific and external to the model, so it is the natural measure of stored knowledge. Writing each pair with $w_1$ alphabetically first, we measure it as $r = \text{freq}_{w_1 \text{and} w_2} / (\text{freq}_{w_1 \text{and} w_2} + \text{freq}_{w_2 \text{and} w_1}) - 0.5$, the share of the corpus's occurrences that place the alphabetically first word first, centered so that $r$ runs from $-0.5$ to $+0.5$ and zero means the corpus attests both orders equally often. Alphabetical order is an arbitrary reference frame rather than a preferred one: it is the majority order for almost exactly half of the binomials in each corpus, and $r$, $\hat{y}$, and $y_{\text{true}}$ are all signed the same way, so a positive value on any of the three means the alphabetically first word comes first. Entering both predictors together, and letting each be moderated by how often the model saw the binomial ($f$, log total frequency), gives

$$
\begin{aligned}
y_{\text{true}} &\sim \beta_0 + \beta_1\hat{y} + \beta_2 r + \beta_3 f\\
                &\quad + \beta_4\,(\hat{y} \times f) + \beta_5\,(r \times f)\\
\log \sigma     &= \gamma_0 + \gamma_1 f
\end{aligned}
$$ {#eq-relfreq}

fit in *brms* [@burkner2017brms] with default priors. Holistic storage makes two predictions about the moderators: as a binomial becomes entrenched, the abstract component should lose ground ($\beta_4 < 0$) while its own attested order should gain ground ($\beta_5 > 0$). Attention-zeroing should attenuate both, since it prevents any information about the two words jointly from reaching the probe.

The residual scale is modeled rather than assumed constant. The spread of $y_{\text{true}}$ varies with frequency, and an unmodeled trend there would inflate the slopes at one end of the range; $\gamma_1$ absorbs it, leaving the interaction terms to capture changes in how strongly each predictor governs the ordering.

Frequency is each binomial's exposure in the model's own training corpus: exact counts in the BabyLM corpus [@warstadt-etal-2023-findings] for the BabyLM models, and infini-gram counts in the Pile [@gao2020pile; @liu2024infinigram] for the rest, which are exact only for Pythia and stand in as a proxy for GPT-2, OLMo, and Llama. Reduplicatives such as *again and again* are excluded, since the two orderings are the same string (Appendix\ \ref{sec-appendix-relfreq}).

# Experiment 1: BabyLM Representations {#sec-exp1}

In this first experiment, we probe three BabyLMs (125M, 350M, and 1.3B parameters)[^2] at their final training checkpoint. Because these models were trained on an amount of data comparable to humans, and because the novel set excludes binomials in their training corpus, they give the cleanest test of whether a language model encodes abstract binomial ordering preferences: any preference the probe recovers for a novel binomial must have been generalized rather than retrieved.

[^2]: The BabyLM models used here are OPT-based models trained for 20 epochs on a 150M-token subset of the BabyLM benchmark corpus [@houghton2026holisticstorageverbupphrases; @warstadt-etal-2023-findings].

```{r}
#| label: load-brms-babylm
#| include: false

babylm_labels <- c("BabyLM-125M", "BabyLM-350M", "BabyLM-1.3B")
# ── Main model: y_true ~ y_pred*log_freq + rel_freq*log_freq, sigma ~ log_freq ──
# (Scripts/analysis/corpus_relfreq_brms_all.R). All four variables z-scored within each
# cell, so coefficients are standardized betas and comparable across cells; the
# file holds final checkpoints (step = NA) and training checkpoints together.
relfreq_all <- if (file.exists("../Data/brms_relfreq_prop.rds")) {
  readRDS("../Data/brms_relfreq_prop.rds") |>
    filter(label %in% model_order) |>
    mutate(model = factor(label, levels = model_order),
           family = family_map[label])
} else tibble()

# Experiments 1 and 2: final checkpoints only
relfreq <- if (nrow(relfreq_all)) filter(relfreq_all, is.na(step)) else tibble()
# Experiment 3: training checkpoints
relfreq_ckpt <- if (nrow(relfreq_all)) filter(relfreq_all, !is.na(step)) else tibble()

# pull one coefficient, formatted; term is the stem used in the summary columns
rf <- function(lbl, cond, term, digits = 3) {
  r <- relfreq[relfreq$label == lbl & relfreq$condition == cond, ]
  if (!nrow(r)) return("--")
  sprintf(paste0("%+.", digits, "f [%+.", digits, "f, %+.", digits, "f]"),
          r[[paste0(term, "_mean")]], r[[paste0(term, "_lo")]], r[[paste0(term, "_hi")]])
}
rfm <- function(lbl, cond, term, digits = 3) {
  r <- relfreq[relfreq$label == lbl & relfreq$condition == cond, ]
  if (!nrow(r)) return("--") else sprintf(paste0("%+.", digits, "f"), r[[paste0(term, "_mean")]])
}
# how many models have this coefficient credibly on the given side of zero
rf_n <- function(cond, term, side = "pos", fam = NULL) {
  d <- relfreq[relfreq$condition == cond, ]
  if (!is.null(fam)) d <- d[d$family %in% fam, ]
  if (!nrow(d)) return(NA_integer_)
  if (side == "pos") sum(d[[paste0(term, "_lo")]] > 0) else sum(d[[paste0(term, "_hi")]] < 0)
}
rf_tot <- function(cond, fam = NULL) {
  d <- relfreq[relfreq$condition == cond, ]
  if (!is.null(fam)) d <- d[d$family %in% fam, ]
  nrow(d)
}
# percent change in a main effect from full context to attention-zeroing
rf_drop <- function(lbl, term) {
  a <- relfreq[relfreq$label == lbl & relfreq$condition == "default", ]
  b <- relfreq[relfreq$label == lbl & relfreq$condition == "attn_zeroed", ]
  if (!nrow(a) || !nrow(b)) return(NA_real_)
  100 * (1 - b[[paste0(term, "_mean")]] / a[[paste0(term, "_mean")]])
}
rf_drop_rng <- function(term, fam = NULL) {
  labs <- unique(relfreq$label)
  if (!is.null(fam)) labs <- unique(relfreq$label[relfreq$family %in% fam])
  v <- vapply(labs, rf_drop, numeric(1), term = term)
  range(v[is.finite(v)])
}
large_fams <- c("GPT-2", "Pythia", "OLMo", "Llama")
large_models <- model_order[!model_order %in% c("BabyLM-125M","BabyLM-350M","BabyLM-1.3B")]

# ── Model-implied simple slopes (Scripts/analysis/prepare_relfreq_slopes.R) ───────────
# beta_1 + beta_4*f and beta_2 + beta_5*f from each fit's posterior, evaluated
# at each decile's median frequency. Drawn from the model rather than refit per
# decile, so the lines carry the pooling the model performs and have credible
# intervals.─
relfreq_curve <- if (file.exists("../Data/relfreq_slope_curve.csv")) {
  read_csv("../Data/relfreq_slope_curve.csv", show_col_types = FALSE) |>
    filter(label %in% model_order) |>
    mutate(model = factor(label, levels = model_order),
           family = factor(family_map[label],
                           levels = c("BabyLM","GPT-2","Pythia","OLMo","Llama")),
           term = relabel_term(term))
} else tibble()

# The frequency axis. prepare_relfreq_slopes.R undoes the z-score and supplies
# mid_freq, a raw corpus count, which is what a reader can actually interpret;
# on a log axis the simple-slope line stays straight, since @eq-relfreq is
# linear in log_freq. If that column is absent the figures fall back to the
# standardized column so the document still renders, just with a less readable
# axis. Both branches plot the same quantity.
has_raw_freq <- nrow(relfreq_curve) > 0 && "mid_freq" %in% names(relfreq_curve)
if (nrow(relfreq_curve)) {
  relfreq_curve <- relfreq_curve |>
    mutate(freq_x = if (has_raw_freq) mid_freq else mid_log_freq_z)
  # Older runs of the slopes script emitted decile rows only; newer ones add a
  # dense grid spanning the fitted frequency range. Lines come from the grid
  # where it exists, points always from the deciles.
  if (!"kind" %in% names(relfreq_curve)) relfreq_curve$kind <- "decile"
}
curve_line  <- function(d) if (any(d$kind == "grid")) filter(d, kind == "grid") else d
curve_point <- function(d) filter(d, kind == "decile")
freq_axis <- if (has_raw_freq) {
  list(scale_x_log10(breaks = scales::breaks_log(4),
                     labels = scales::label_number(big.mark = ",", accuracy = 1)),
       labs(x = "Binomial frequency in training corpus (log scale)"))
} else {
  list(scale_x_continuous(breaks = c(-1, 0, 1)),
       labs(x = "Binomial frequency (standardized log)"))
}

# observed standardized slope for one model / term / decile
# trm is given as the variable name ("y_pred"/"rel_freq"); the table stores the
# display label, so translate before matching
rc <- function(lbl, trm, dec, cond = "default") {
  r <- relfreq_curve[relfreq_curve$label == lbl &
                     relfreq_curve$term == relabel_term(trm) &
                     relfreq_curve$decile == dec & relfreq_curve$condition == cond, ]
  if (!nrow(r)) "--" else sprintf("%.2f", r$beta[1])
}

```

```{r}
#| label: ctrl-stats
#| include: false

ctrl_r2_exp1 <- all_r2 |>
  filter(family == "BabyLM", is_control, mode == "mean_pooled") |>
  pull(mean_r2)

ctrl_r2_exp2 <- all_r2 |>
  filter(is_control, mode == "mean_pooled") |>
  pull(mean_r2)

ctrl_mean_exp1 <- mean(ctrl_r2_exp1)
ctrl_sd_exp1   <- sd(ctrl_r2_exp1)
ctrl_mean_exp2 <- mean(ctrl_r2_exp2)
ctrl_sd_exp2   <- sd(ctrl_r2_exp2)
```

## Results

### Probe Performance

Results are shown in @fig-r2-babylm; full numerical results are in Appendix\ \ref{sec-appendix-exp1} (@tbl-r2-babylm). The control probe R² values were consistently near zero across all conditions and all three models (mean ± SD = `r sprintf("%.4f ± %.4f", ctrl_mean_exp1, ctrl_sd_exp1)`; full control results are in @tbl-ctrl-r2-babylm), confirming that the probe doesn't recover spurious signal from label noise.

```{r}
#| label: fig-r2-babylm
#| fig-cap: "Cross-validated probe R² at the final hidden layer for the BabyLM models, binomial level, by attention condition. Bars = ±1 SD across 10 folds (minimum width enforced for visibility). The drop from the left to the right of each panel is the lexical generalization cost."
#| fig-width: 3.3
#| fig-height: 2.3
#| eval: !expr nrow(all_r2) > 0

all_r2 |>
  filter(family == "BabyLM", mode == "mean_pooled", !is_control) |>
  mutate(
    model      = factor(model, levels = model_order),
    condition  = recode_condition(condition),
    split_lbl  = factor(recode(split,
                               "pair_novel" = "Binomial-
held-out",
                               "word_novel" = "Word-
held-out"),
                        levels = c("Binomial-
held-out", "Word-
held-out")),
    display_sd = pmax(sd_r2, 0.010)
  ) |>
  ggplot(aes(x = split_lbl, y = mean_r2, color = model)) +
  geom_pointrange(aes(ymin = mean_r2 - display_sd, ymax = mean_r2 + display_sd),
                  position = position_dodge(0.55), linewidth = 0.5, size = 0.35) +
  facet_wrap(~condition, nrow = 1) +
  # size only: the full model names overflow the single-column width and the
  # last label gets clipped
  scale_color_manual(values = babylm_size_colors, name = NULL,
                     labels = c("125M", "350M", "1.3B")) +
  labs(x = NULL, y = expression(R^2)) +
  theme_paper(base_size = 9) +
  theme(legend.position = "bottom",
        legend.key.width = unit(10, "pt"),
        legend.margin = margin(t = -4),
        legend.text = element_text(size = 7),
        axis.text.x = element_text(size = 7))
```

For the binomial-level conditions, binomial-held-out R² ranges from 0.24 to 0.34; word-held-out R² is consistently lower, ranging from 0.12 to 0.23. The drop reflects a lexical generalization cost: the probe is more accurate when a test binomial's words appeared in other training binomials than when both are novel, so some preference signal is tied to word identity rather than being fully abstract. Probe accuracy is non-monotonic with model size, and the full-context condition yields higher binomial-held-out R² than attention-zeroed, a difference we report but do not build on.

### Abstract and Stored Contributions

Both sources predict the model's ordering preference, and both do so credibly in all three models. Under full context, $\hat{y}$ carries a coefficient between `r sprintf("%+.3f", min(sapply(babylm_labels, function(l) relfreq$pred_mean[relfreq$label==l & relfreq$condition=="default"])))` and `r sprintf("%+.3f", max(sapply(babylm_labels, function(l) relfreq$pred_mean[relfreq$label==l & relfreq$condition=="default"])))`, and $r$ one between `r sprintf("%+.3f", min(sapply(babylm_labels, function(l) relfreq$rel_mean[relfreq$label==l & relfreq$condition=="default"])))` and `r sprintf("%+.3f", max(sapply(babylm_labels, function(l) relfreq$rel_mean[relfreq$label==l & relfreq$condition=="default"])))`, with 95% credible intervals excluding zero throughout (Appendix\ \ref{sec-appendix-relfreq}). A BabyLM's preference for an attested binomial is thus governed partly by what generalizes from binomials it never saw, and partly by the order its own corpus happens to attest for that specific pair. Neither alone accounts for the ordering.

Attention-zeroing separates the two. Blocking cross-word attention reduces the contribution of the corpus's attested order by `r sprintf("%.0f", min(sapply(c("BabyLM-125M","BabyLM-350M","BabyLM-1.3B"), rf_drop, term="rel")))`–`r sprintf("%.0f", max(sapply(c("BabyLM-125M","BabyLM-350M","BabyLM-1.3B"), rf_drop, term="rel")))`%, against `r sprintf("%.0f", min(sapply(c("BabyLM-125M","BabyLM-350M","BabyLM-1.3B"), rf_drop, term="pred")))`–`r sprintf("%.0f", max(sapply(c("BabyLM-125M","BabyLM-350M","BabyLM-1.3B"), rf_drop, term="pred")))`% for the abstract component. The two ranges overlap, but the asymmetry holds within each model taken separately: the attested order loses more than the abstract component does in all three. Since the zeroed representation is additive in the two words and can carry no information about them jointly, the component that gives way is the one that required them to interact.

The frequency moderation is present here too. The attested order gains ground with exposure in all three models (`r rf("BabyLM-125M","default","rel_x")`, `r rf("BabyLM-350M","default","rel_x")` and `r rf("BabyLM-1.3B","default","rel_x")`), and in each case that gain is roughly halved under attention-zeroing (`r rfm("BabyLM-125M","attn_zeroed","rel_x")`, `r rfm("BabyLM-350M","attn_zeroed","rel_x")` and `r rfm("BabyLM-1.3B","attn_zeroed","rel_x")`). These are the same magnitudes the large-scale models show in Experiment 2, so the models trained on a human-scale corpus need no separate account: the frequency-dependent trade-off appears at 100M tokens and does not require web-scale exposure to emerge. The caveat is on precision rather than direction. 70% of BabyLM's binomials occur exactly once, so $r$ sits at a boundary value for most of the set and its lowest frequency deciles consist entirely of singletons, leaving the measure coarse at the low-frequency end (Appendix\ \ref{sec-appendix-relfreq}). Experiment 2 supplies the frequency range these models lack.

```{r}
#| label: fig-relfreq-babylm
#| fig-cap: "How strongly each source contributes to the BabyLM models' ordering preference, by binomial frequency. Values are simple slopes from the fitted posteriors of @eq-relfreq, evaluated at each decile's median frequency and plotted against that frequency. The lines are straight because @eq-relfreq is linear in `log_freq`, so a simple slope is exactly linear in it; points mark the decile medians, and their spacing shows where the binomials actually are. Bands are 95\\% credible intervals, widest at the extremes because a simple slope is least precisely estimated far from the mean. Under full context relative frequency gains ground as frequency rises while the abstract component loses it; under attention-zeroing relative frequency sits lower throughout."
#| fig-width: 3.3
#| fig-height: 3.0
#| eval: !expr nrow(relfreq_curve) > 0

# x is the decile's median log frequency, not the decile index. @eq-relfreq is
# linear in log_freq, so the simple slope beta_1 + beta_4*f is exactly linear in
# f; plotting it against decile RANK re-expresses that straight line on a
# nonlinear axis and draws it as an S-curve, because ranking compresses the
# middle of a skewed frequency distribution and stretches its tails. Points
# still sit at decile medians, so their spacing shows where the data actually
# is. The intervals are widest at the ends because a simple slope is estimated
# least precisely far from the mean.
bl <- relfreq_curve |>
  filter(family == "BabyLM") |>
  mutate(condition = recode_condition(condition))

ggplot(mapping = aes(freq_x, beta, color = term, fill = term, linetype = model)) +
  geom_ribbon(data = curve_line(bl),
              aes(ymin = lo, ymax = hi, group = interaction(model, term)),
              alpha = 0.13, colour = NA) +
  geom_line(data = curve_line(bl), linewidth = 0.5) +
  geom_point(data = curve_point(bl), size = 0.7) +
  facet_wrap(~condition, nrow = 1) +
  scale_color_manual(values = term_colors, name = NULL) +
  scale_fill_manual(values = term_colors, guide = "none") +
  # size only: the full labels overflow the single-column width
  scale_linetype_manual(values = c("solid", "42", "12"), name = NULL,
                        labels = c("125M", "350M", "1.3B")) +
  freq_axis +
  labs(y = "Standardized slope on ordering preference") +
  guides(linetype = guide_legend(order = 1), color = guide_legend(order = 2)) +
  theme_paper(base_size = 9) +
  theme(legend.position = "bottom", legend.box = "vertical",
        legend.key.width = unit(12, "pt"),
        legend.spacing.y = unit(0, "pt"),
        legend.text = element_text(size = 7),
        legend.margin = margin(t = -4))
```

# Experiment 2: Cross-Family Final-Layer Representations {#sec-exp2}

Experiment 1 demonstrates that BabyLMs rely on both abstract preferences and holistic storage for binomial ordering. Those models are small in both parameters and training data, so Experiment 2 applies the same probe to 13 further models across four large-scale families: Pythia, GPT-2, OLMo, and Llama, with the BabyLM results carried over for comparison (Appendix\ \ref{sec-appendix-models}).

## Results

### Probe Performance

@fig-r2 shows both cross-validation splits for all 16 models; full numbers, control probes, and the word-level replication are in Appendix\ \ref{sec-appendix-exp2}. Control R² was near zero throughout (mean ± SD = `r sprintf("%.4f ± %.4f", ctrl_mean_exp2, ctrl_sd_exp2)`). All large-scale models substantially outperform BabyLM: full-context binomial-held-out R² spans 0.43 to 0.65 against 0.29–0.34, and word-held-out R² is lower than binomial-held-out for every model and condition. Attention-zeroing raises binomial-held-out R² in every large-scale family, most sharply in GPT-2, where it reaches 0.79 at 117M against 0.65 under full context; Llama-3-8B is the sole exception. Scale effects within families are non-monotonic, with GPT-2 R² *decreasing* with size, from 0.65 at 117M to 0.49 at XL.

```{r}
#| label: fig-r2
#| fig-cap: "Cross-validated R² for all 16 models, by attention condition and cross-validation split. Circles = binomial-held-out; triangles = word-held-out. Bars = 95\\% CI across 10 folds. Models ordered by parameter count within each family and colored by family. Control R² $\\approx$ 0 (Appendix\\ \\ref{sec-appendix-exp2})."
#| fig-width: 6.5
#| fig-height: 2.5
#| fig-env: "figure*"
#| eval: !expr nrow(all_r2) > 0

size_labels <- c(
  "BabyLM-125M" = "125M", "BabyLM-350M" = "350M", "BabyLM-1.3B" = "1.3B",
  "GPT-2" = "S", "GPT-2-M" = "M", "GPT-2-L" = "L", "GPT-2-XL" = "XL",
  "Pythia-160M" = "160M", "Pythia-410M" = "410M", "Pythia-1B" = "1B",
  "OLMo-1B" = "1B", "OLMo-7B" = "7B", "OLMo-2-1B" = "2-1B", "OLMo-2-7B" = "2-7B",
  "Llama-1.3B" = "1.3B", "Llama-3-8B" = "8B"
)

all_r2 |>
  filter(mode == "mean_pooled", !is_control) |>
  mutate(
    condition = recode_condition(condition),
    split_lbl = factor(recode(split,
                              "pair_novel" = "Binomial-held-out",
                              "word_novel" = "Word-held-out"),
                       levels = c("Binomial-held-out", "Word-held-out")),
    family    = factor(family, levels = c("BabyLM", "GPT-2", "Pythia", "OLMo", "Llama")),
    model_fct = factor(model, levels = dumbbell_order),
    ci_half   = qt(0.975, df = 9) * sd_r2 / sqrt(n_folds),
    r2_lo     = mean_r2 - pmax(ci_half, 0.02),
    r2_hi     = mean_r2 + pmax(ci_half, 0.02)
  ) |>
  ggplot(aes(x = model_fct, y = mean_r2, color = family, shape = split_lbl)) +
  geom_linerange(aes(ymin = r2_lo, ymax = r2_hi), linewidth = 0.7,
                 position = position_dodge(width = 0.6)) +
  geom_point(size = 1.8, position = position_dodge(width = 0.6)) +
  facet_grid(condition ~ family, scales = "free_x", space = "free_x") +
  scale_color_manual(values = family_colors, guide = "none") +
  scale_shape_manual(values = c("Binomial-held-out" = 19, "Word-held-out" = 17),
                     name = NULL) +
  scale_x_discrete(labels = size_labels) +
  scale_y_continuous(breaks = c(0.25, 0.5, 0.75)) +
  labs(y = expression(R^2), x = NULL) +
  theme_paper(base_size = 9) +
  theme(
    panel.spacing.x = unit(0.35, "lines"),
    legend.position  = "bottom",
    legend.margin    = margin(t = -4),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6.5),
  
    # Make the right-hand facet labels smaller
    strip.text.y.right = element_text(size = 7),
  
    # Don't clip the facet-strip text if the rows get shorter
    strip.clip = "off"
  )
```

```{r}
#| label: load-brms-all
#| include: false

```

### Abstract and Stored Contributions

The frequency range BabyLM lacks sharpens the moderation and separates the two moderators. Under full context, the contribution of the corpus's attested order grows with exposure in all `r rf_tot("default", large_fams)` large-scale models, and the abstract component's shrinks in all `r rf_tot("default", large_fams)`, with 95% credible intervals excluding zero in every case (@tbl-relfreq; @fig-relfreq-coefs). The two moderators run in opposite directions on the same items: the more often a model saw a binomial, the more its ordering is governed by that binomial's own attested order and the less by what generalizes from binomials it never saw.

The effect is substantial rather than marginal. In GPT-2 the two sources cross over: the attested order's standardized slope climbs from `r rc("GPT-2","rel_freq",1)` in the rarest frequency decile to `r rc("GPT-2","rel_freq",10)` in the most frequent, while the abstract component's falls from `r rc("GPT-2","y_pred",1)` to `r rc("GPT-2","y_pred",10)`, so that for the most frequent binomials the corpus's own attested order is the better predictor of the two (@fig-relfreq-pile). The crossing is not universal: in Llama-3-8B the attested order climbs from `r rc("Llama-3-8B","rel_freq",1)` to `r rc("Llama-3-8B","rel_freq",10)` while the abstract component falls from `r rc("Llama-3-8B","y_pred",1)` to `r rc("Llama-3-8B","y_pred",10)`, converging without meeting. The direction is the same in every model; whether the two actually cross depends on where each starts.

Attention-zeroing removes this selectively, which is what identifies where the stored component lives. The attested order's contribution falls by `r sprintf("%.0f", min(sapply(large_models, rf_drop, term="rel")))`–`r sprintf("%.0f", max(sapply(large_models, rf_drop, term="rel")))`% across the large-scale models, and its moderation by frequency falls by roughly `r sprintf("%.0f", median(sapply(large_models, rf_drop, term="rel_x")))`% alongside it, though it remains credible in `r rf_n("attn_zeroed","rel_x","pos",large_fams)` of `r rf_tot("attn_zeroed", large_fams)` models rather than vanishing outright. The abstract component moves the other way: it is *larger* under attention-zeroing in every one of the `r rf_tot("default", large_fams)`, by `r sprintf("%.0f", -max(sapply(large_models, rf_drop, term="pred")))`–`r sprintf("%.0f", -min(sapply(large_models, rf_drop, term="pred")))`%. Since the zeroed representation is additive in the two words, no information about them jointly can reach the probe, so what the intervention strips is the component that required the two words to be encoded together. A blunt degradation of the representation would have reduced both terms; this reduces one and raises the other. It is worth being clear about what remains: roughly half of the attested order's contribution survives an additive representation, so the item-specific knowledge is not carried by a pair representation alone. Experiment 3 shows this surviving portion is the part that appears first, and @sec-discussion takes up what it is likely to be.

One term does not behave as the storage account predicts. The abstract component's moderation stays negative under attention-zeroing in all `r rf_tot("attn_zeroed", large_fams)` large-scale models rather than flattening, so its decline with frequency is not caused by stored orderings displacing it. The probe is trained on novel binomials, which by construction have no corpus frequency; the words of frequent attested binomials have had their representations shaped by those binomial contexts, and a mapping learned from nonce items transfers less well to them. That is a claim about lexical representations rather than phrasal storage, and the intervention leaves it intact because word-additive representations survive the mask.

```{r}
#| label: fig-relfreq-pile
#| fig-cap: "How strongly each source contributes to the ordering preference, by binomial frequency, for the 13 large-scale models. Values are simple slopes from the fitted posteriors of @eq-relfreq, evaluated at each decile's median frequency and plotted against that frequency; faint lines are individual models with their 95\\% credible intervals, bold lines their mean. The lines are straight because @eq-relfreq is linear in `log_freq`. Under full context the abstract component weakens and relative frequency strengthens in all 13; the two cross in Pythia, GPT-2 and Llama-1.3B but not in OLMo or Llama-3-8B. Under attention-zeroing relative frequency sits far lower throughout."
#| fig-width: 3.3
#| fig-height: 2.8
#| eval: !expr nrow(relfreq_curve) > 0
#| fig-pos: 't'

# One faint line per model behind a bold mean. The individual traces show that
# the aggregate is not hiding disagreement, without needing thirteen legend
# entries or a facet per family.
#
# x is the decile's median log frequency rather than the decile index: the
# simple slope is exactly linear in log_freq, and plotting it against decile
# rank would bend that straight line into an S-curve (see @fig-relfreq-babylm).
lg <- relfreq_curve |>
  filter(family %in% large_fams) |>
  mutate(condition = recode_condition(condition))

ggplot(mapping = aes(freq_x, beta, color = term, fill = term)) +
  # per-model posterior intervals, faint; between-model spread dominates them
  geom_ribbon(data = curve_line(lg),
              aes(ymin = lo, ymax = hi, group = interaction(label, term)),
              alpha = 0.10, colour = NA) +
  geom_line(data = curve_line(lg), aes(group = interaction(label, term)),
            alpha = 0.22, linewidth = 0.3) +
  stat_summary(data = curve_line(lg), aes(group = term), fun = mean,
               geom = "line", linewidth = 1) +
  stat_summary(data = curve_point(lg), aes(group = term), fun = mean,
               geom = "point", size = 1.1) +
  facet_wrap(~condition, nrow = 1) +
  scale_fill_manual(values = term_colors, guide = "none") +
  scale_color_manual(values = term_colors, name = NULL) +
  freq_axis +
  labs(y = "Standardized slope on ordering preference") +
  theme_paper(base_size = 9) +
  theme(legend.position = "bottom",
        legend.key.width = unit(12, "pt"),
        legend.margin = margin(t = -4))
```

# Experiment 3: Developmental Trajectories {#sec-exp3}

Experiments 1 and 2 establish that language models encode binomial preferences at the end of training, but leave open when those representations emerge, and whether the trade-off between abstract preferences and holistic storage is present early or accumulates with exposure. Experiment 3 examines the same probe at multiple training checkpoints.

The probe, input modes, attention conditions, and cross-validation splits are as described in @sec-methods. We refit @eq-relfreq at each saved checkpoint of the three BabyLM models and the three Pythia models for which checkpoints were available, in both attention conditions: 72 fits in all. Because every variable is standardized within each fit, coefficients are comparable across checkpoints as well as across models.

The two sets of checkpoints cover very different portions of training and should not be read as a common axis. BabyLM's six steps span roughly the first two-thirds of its run, whereas Pythia's stop at step 1000 of about 143,000, or the first 0.7%. Pythia therefore shows what happens very early, and its final-checkpoint models from Experiment 2 show where that leads, with nothing in between.

## Results

```{r}
#| label: load-r2-checkpoints
#| cache: false
#| include: false

load_ckpt_r2 <- function(registry) {
  map_dfr(seq_len(nrow(registry)), function(i) {
    slug  <- registry$slug[i]
    label <- registry$label[i]
    step  <- registry$step[i]
    tok   <- registry$tokens_M[i]
    bind_rows(
      load_mlp(slug, label, FALSE),
      load_mlp(slug, label, TRUE)
    ) |>
      mutate(step = step, tokens_M = tok)
  }) |>
    filter(split %in% c("pair_novel", "word_novel")) |>
    group_by(model, condition, mode, split, is_control, step) |>
    filter(layer == max(layer)) |>
    ungroup()
}

babylm_ckpt_r2  <- load_ckpt_r2(babylm_ckpt_registry)
pythia_ckpt_r2  <- load_ckpt_r2(pythia_ckpt_registry)

ckpt_r2 <- bind_rows(
  babylm_ckpt_r2 |> mutate(family_lbl = "BabyLM"),
  pythia_ckpt_r2 |> mutate(family_lbl = "Pythia")
) |> mutate(family_lbl = factor(family_lbl, levels = c("BabyLM", "Pythia")))
```

```{r}
#| label: load-brms-checkpoints
#| include: false

# Checkpoint fits of @eq-relfreq, joined to the token counts each step
# corresponds to so the trajectories can be plotted on an exposure axis.
ckpt_tokens <- bind_rows(
  babylm_ckpt_registry |> select(label, step, tokens_M),
  pythia_ckpt_registry |> select(label, step, tokens_M)
)

ckpt_coefs <- relfreq_ckpt |>
  inner_join(ckpt_tokens, by = c("label", "step")) |>
  mutate(family_lbl = factor(ifelse(family == "BabyLM", "BabyLM", "Pythia"),
                             levels = c("BabyLM", "Pythia")),
         condition_lbl = recode_condition(condition))

# how much of the stored contribution the intervention removes, per checkpoint
ckpt_gap <- ckpt_coefs |>
  select(label, step, tokens_M, family_lbl, condition, rel_mean, rel_x_mean) |>
  tidyr::pivot_longer(c(rel_mean, rel_x_mean), names_to = "term") |>
  tidyr::pivot_wider(names_from = condition, values_from = value) |>
  mutate(gap = default - attn_zeroed,
         term = factor(term, levels = c("rel_mean", "rel_x_mean"),
                       labels = c("rel_freq", "rel_freq : log_freq")))

# ── accessors for the Experiment 3 prose ─────────────────────────────────────
# BabyLM and Pythia are checkpointed at different step numbers, so "first" and
# "last" are resolved per model by rank rather than by a shared step. Term stems
# are the summary-column stems: pred, rel, pred_x, rel_x.
ck <- ckpt_coefs |>
  group_by(label) |>
  mutate(ck_rank = dense_rank(step), ck_last = ck_rank == max(ck_rank)) |>
  ungroup()

babylm_ck <- c("BabyLM-125M", "BabyLM-350M", "BabyLM-1.3B")
pythia_ck <- c("Pythia-160M", "Pythia-410M", "Pythia-1B")
N_EARLY   <- 3          # "the first three checkpoints" in the prose below

fmt3 <- function(x) sprintf("%.3f", x)
# "a, b and c"
serial <- function(v) paste0(paste(head(v, -1), collapse = ", "), " and ", tail(v, 1))
rng3 <- function(x) paste0(fmt3(min(x)), "--", fmt3(max(x)))

# one coefficient; rank is an integer or "last"
ck_num <- function(lbl, rank, term, cond = "default") {
  r <- if (identical(rank, "last")) ck[ck$label == lbl & ck$ck_last & ck$condition == cond, ]
       else ck[ck$label == lbl & ck$ck_rank == rank & ck$condition == cond, ]
  if (!nrow(r)) NA_real_ else r[[paste0(term, "_mean")]]
}
ck_v <- function(lbl, rank, term, cond = "default") fmt3(ck_num(lbl, rank, term, cond))
ck_list <- function(labs, rank, term, cond = "default")
  serial(fmt3(vapply(labs, ck_num, numeric(1), rank = rank, term = term, cond = cond)))

# condition gap: full context minus attention-zeroed
ck_gap <- function(lbl, rank, term)
  ck_num(lbl, rank, term, "default") - ck_num(lbl, rank, term, "attn_zeroed")
ck_gap_list <- function(labs, rank, term)
  serial(fmt3(vapply(labs, ck_gap, numeric(1), rank = rank, term = term)))
ck_gap_rng <- function(labs, rank, term)
  rng3(vapply(labs, ck_gap, numeric(1), rank = rank, term = term))
# Share of rel_freq's contribution that requires the two words encoded jointly,
# i.e. the fraction attention-zeroing removes. The remainder survives a strictly
# additive representation, so it has to be carried by the individual word
# vectors rather than by any interaction between them.
#
# Only quotable once the main effect is credibly positive. Before that the
# denominator is indistinguishable from zero and the ratio is noise: at the
# first three checkpoints it takes values like 2697% and -291%.
ck_joint <- function(lbl, rank)
  100 * (1 - ck_num(lbl, rank, "rel", "attn_zeroed") / ck_num(lbl, rank, "rel", "default"))
ck_joint_rng <- function(labs, rank) {
  v <- vapply(labs, ck_joint, numeric(1), rank = rank)
  sprintf("%.0f--%.0f", min(v), max(v))
}
fin_joint <- function(lbl) {
  r <- relfreq[relfreq$label == lbl, ]
  100 * (1 - r$rel_mean[r$condition == "attn_zeroed"] / r$rel_mean[r$condition == "default"])
}
fin_joint_rng <- function(labs) {
  v <- vapply(labs, fin_joint, numeric(1))
  sprintf("%.0f--%.0f", min(v), max(v))
}

# largest condition gap anywhere in the early window, across the given terms
ck_gap_maxabs <- function(terms = c("rel", "rel_x"), ranks = 1:4) {
  v <- unlist(lapply(terms, function(t)
    vapply(unique(ck$label), function(l)
      max(abs(vapply(ranks, function(r) ck_gap(l, r, t), numeric(1)))), numeric(1))))
  fmt3(max(v))
}
# is the coefficient monotone past the early window within one condition?
ck_mono_cond <- function(term, cond) {
  all(vapply(unique(ck$label), function(l) {
    g <- ck[ck$label == l & ck$condition == cond, ]
    v <- g[[paste0(term, "_mean")]][order(g$ck_rank)][-seq_len(N_EARLY)]
    all(diff(v) > 0)
  }, logical(1)))
}

# cell counts over checkpoint subsets
ck_sub    <- function(early = TRUE, cond = NULL) {
  d <- if (early) ck[ck$ck_rank <= N_EARLY, ] else ck[ck$ck_rank > N_EARLY, ]
  if (!is.null(cond)) d <- d[d$condition == cond, ]
  d
}
n_cred    <- function(d, term) sum(d[[paste0(term, "_lo")]] > 0 | d[[paste0(term, "_hi")]] < 0)
n_cred_p  <- function(d, term) sum(d[[paste0(term, "_lo")]] > 0)
n_cred_n  <- function(d, term) sum(d[[paste0(term, "_hi")]] < 0)
ck_maxabs <- function(d, term) fmt3(max(abs(d[[paste0(term, "_mean")]])))
# does the coefficient rise at every step past the early window, in every cell?
# Guards the word "monotonically": it was not true of the previous fits, where
# two attention-zeroed Pythia cells fall back at the last checkpoint.
ck_mono <- function(term) {
  all(vapply(split(ck, list(ck$label, ck$condition), drop = TRUE), function(g) {
    g <- g[order(g$ck_rank), ]; v <- g[[paste0(term, "_mean")]][g$ck_rank > N_EARLY]
    all(diff(v) > 0)
  }, logical(1)))
}
# a final-checkpoint (non-checkpoint) coefficient, for contrast with the trajectory
fin_v <- function(lbl, term, cond = "default") {
  r <- relfreq[relfreq$label == lbl & relfreq$condition == cond, ]
  if (!nrow(r)) "--" else fmt3(r[[paste0(term, "_mean")]])
}
```

Probe trajectories are shown in @fig-checkpoints and the two sources' contributions in @fig-ckpt-sources. BabyLM binomial-held-out R² is already high at the earliest checkpoint (\~20M tokens: 0.67–0.72) and broadly declines to 0.32–0.38 by the end of training. At the earliest checkpoint, the attention-zeroed condition outperforms full context (0.78–0.80 vs. 0.67–0.72); this ordering reverses between \~120M and \~750M tokens, so that full context exceeds attention-zeroed at the final checkpoint for all three model sizes. Pythia shows an inverted-U trajectory instead: R² rises from 0.68–0.74 at 33.6M tokens to a peak of 0.80–0.86 at 537M tokens, then declines to 0.71–0.76 at 1074M–2097M tokens, consistent with large-scale models still consolidating preference representations during this early window (the first 0.7% of 300B tokens). The attention condition flip for Pythia follows the same qualitative pattern as BabyLM. Full numerical and word-level results are in Appendix\ \ref{sec-appendix-exp3}.

The stored contribution accumulates. Across the first three checkpoints `rel_freq` is negligible in every model, never exceeding `r ck_maxabs(ck_sub(TRUE), "rel")` in absolute value and credibly different from zero in only `r n_cred(ck_sub(TRUE), "rel")` of `r nrow(ck_sub(TRUE))` cells, in both directions. From the fourth checkpoint it is credibly positive in `r n_cred_p(ck_sub(FALSE), "rel")` of `r nrow(ck_sub(FALSE))` model-by-condition cells and climbs `r if (ck_mono("rel")) "monotonically " else ""`thereafter: to `r ck_list(babylm_ck, "last", "rel")` in the three BabyLM models and to `r ck_list(pythia_ck, "last", "rel")` in the three Pythia models. This holds in both attention conditions, so what grows is not solely the part that requires the two words to interact. Item-specific ordering information is not present from the outset; it is acquired, and it keeps accruing for as long as we can observe.

What the intervention removes appears later than the contribution itself, and the lag is informative. At the fourth checkpoint, the first at which `rel_freq` is credibly positive in every model, attention-zeroing removes none of it: the gap between conditions is `r ck_gap_maxabs()` at most and as often slightly negative as positive, in both corpora and on both terms. Item-specific ordering information is therefore already measurable at a point where it survives a strictly additive representation intact. Whatever carries it at this stage is not holistic storage, which by definition requires the pair: it is the individual lexical representations.

Holistic storage then accumulates on top of it. The share of `rel_freq` that attention-zeroing removes rises from nothing at the fourth checkpoint to `r ck_joint_rng(pythia_ck, "last")`% by the last in the three Pythia models and `r ck_joint_rng(babylm_ck, "last")`% in the three BabyLM models, and reaches `r fin_joint_rng(large_models)`% in the fully trained large-scale models. The frequency moderation follows the same course, its condition gap reaching `r ck_gap_list(babylm_ck, "last", "rel_x")` in BabyLM and `r ck_gap_list(pythia_ck, "last", "rel_x")` in Pythia. Repeated exposure to a binomial thus leaves item-specific knowledge in two forms, and they arrive in order: first in the lexical representations, and only later as holistic storage.
The abstract component moves independently of all this. `y_pred` declines steadily across BabyLM training in both conditions (`r ck_v("BabyLM-125M", 1, "pred")` to `r ck_v("BabyLM-125M", "last", "pred")` in BabyLM-125M under full context) while rising over Pythia's early window and then falling below its starting value by the final model (`r ck_v("Pythia-160M", 1, "pred")` at step 16, `r ck_v("Pythia-160M", 4, "pred")` at step 256, `r fin_v("Pythia-160M", "pred")` fully trained). Its moderation, `y_pred`$\times$`log_freq`, is credibly negative under attention-zeroing in `r n_cred_n(ck[ck$condition == "attn_zeroed", ], "pred_x")` of the `r sum(ck$condition == "attn_zeroed")` checkpoint cells, but with no trend common to the two corpora: it drifts further from zero across BabyLM training and back toward zero across Pythia's. Full numerical results are in Appendix\ \ref{sec-appendix-exp3}.

```{r}
#| label: fig-checkpoints
#| fig-cap: "Cross-validated probe R² across BabyLM and Pythia training checkpoints, by attention condition and cross-validation split. BabyLM's checkpoints span most of its training run, Pythia's only the first 0.7\\% of its own, so the two x-axes are not on a common footing. Control R² $\\approx$ 0."
#| fig-width: 3.3
#| fig-height: 2.6
#| eval: !expr nrow(ckpt_r2) > 0


# Family is the row strip, so colour carries size only. Labels are approximate,
# since the two families' sizes are close but not identical at each step.
rank_of <- c("BabyLM-125M" = "~150M",  "Pythia-160M" = "~150M",
             "BabyLM-350M" = "~400M",  "Pythia-410M" = "~400M",
             "BabyLM-1.3B" = "~1000M", "Pythia-1B"   = "~1000M")
rank_colors <- c("~150M" = "#F0B323", "~400M" = "#4E9D69", "~1000M" = "#154733")
mk_rank <- function(v) factor(rank_of[as.character(v)], levels = names(rank_colors))

split_lt <- c("Binomial-held-out" = "solid", "Word-held-out" = "dashed")

p_r2 <- ckpt_r2 |>
  filter(mode == "mean_pooled", !is_control) |>
  mutate(size_rank = mk_rank(model),
         condition = recode_condition(condition),
         split_lbl = factor(recode(split, "pair_novel" = "Binomial-held-out",
                                   "word_novel" = "Word-held-out"),
                            levels = names(split_lt))) |>
  ggplot(aes(tokens_M, mean_r2, color = size_rank, linetype = split_lbl,
             group = interaction(model, split_lbl))) +
  geom_line(linewidth = 0.5) + geom_point(size = 0.6) +
  facet_grid(family_lbl ~ condition, scales = "free_x", switch = "y") +
  scale_x_log10(breaks = c(100, 1000),
                labels = label_number(suffix = "M", accuracy = 1),
                expand = expansion(mult = c(0.05, 0.1))) +
  scale_color_manual(values = rank_colors, name = NULL) +
  scale_linetype_manual(values = split_lt, name = NULL) +
  labs(x = NULL, y = expression(R^2)) +
  theme_paper(base_size = 9) +
  theme(strip.placement = "outside",
        strip.text.y.left = element_text(angle = 90, size = 7),
        axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  guides(color = guide_legend(nrow = 1, order = 1),
         linetype = guide_legend(nrow = 1, order = 2))

p_r2 +
  theme(legend.position = "bottom", legend.box = "vertical",
        legend.margin = margin(t = -2), legend.spacing.y = unit(0, "pt"),
        legend.key.width = unit(14, "pt"),
        legend.text = element_text(size = 6.5),
        axis.text.x = element_text(size = 7), axis.ticks.x = element_line()) +
  labs(x = "Tokens (M, log scale)")
```

```{r}
#| label: fig-ckpt-sources
#| fig-cap: "All four terms of @eq-relfreq across training, with 95\\% credible intervals. Rows are the two sources and their moderations by frequency, on separate y-scales because they occupy very different ranges; color marks the attention condition and linetype the model size. All variables are standardized within each fit, so values are comparable across checkpoints and models. The abstract component is substantial from the earliest checkpoint while the attested order starts near zero and accumulates. The two conditions track each other on the main effects until the final checkpoints; they separate most clearly on the attested order's moderation (bottom row), where attention-zeroing roughly halves the coefficient by the end of training in every model. The abstract component's moderation stays near zero throughout and shows no trend common to the two corpora. Neither the y-axes nor the x-axes are on a common footing across panels: BabyLM's checkpoints span most of its training run, Pythia's only the first 0.7\\% of its own."
#| fig-width: 3.3
#| fig-height: 5.8
#| eval: !expr nrow(ckpt_coefs) > 0

# Faceting by TERM rather than by condition, with free y-scales. The terms
# occupy very different ranges, so a shared axis sets its scale by y_pred and
# compresses everything else -- including the condition gap, which is the
# quantity of interest -- into the bottom of the panel. Condition moves inside
# the panel as linetype so the comparison is read within a panel rather than
# across two columns.
#
# Both moderation terms get their own row. The two conditions separate only
# modestly on the main effects, which track each other until the last checkpoint
# or two; it is on rel_freq x log_freq that the intervention bites, roughly
# halving the coefficient. A figure showing only main effects understates the
# result the section argues for.
#
# No fig-pos here: the LaTeX default is [tbp], and restricting a float this tall
# to 't' leaves LaTeX no legal slot, so it defers the float to the end of the
# document -- which is how this figure ended up printing after the references.
row_lab <- c(pred  = "Abstract preference",
             predx = "Abstract preference × frequency",
             rel   = "Relative frequency",
             relx  = "Relative frequency × frequency")

purrr::map_dfr(names(row_lab), function(s) {
  stem <- c(pred = "pred", predx = "pred_x", rel = "rel", relx = "rel_x")[[s]]
  ckpt_coefs |> transmute(
    label, family_lbl, condition_lbl, tokens_M,
    term = factor(row_lab[[s]], levels = unname(row_lab)),
    est  = .data[[paste0(stem, "_mean")]],
    lo   = .data[[paste0(stem, "_lo")]],
    hi   = .data[[paste0(stem, "_hi")]])
}) |>
  mutate(size_rank = mk_rank(label)) |>
  ggplot(aes(tokens_M, est, color = condition_lbl, linetype = size_rank,
             group = interaction(label, condition_lbl))) +
  geom_hline(yintercept = 0, color = "gray50", linewidth = 0.3) +
  geom_linerange(aes(ymin = lo, ymax = hi), linewidth = 0.3, alpha = 0.55,
                 linetype = "solid") +
  geom_line(linewidth = 0.5) + geom_point(size = 0.6) +
  facet_grid(term ~ family_lbl, scales = "free", switch = "y",
             labeller = labeller(term = label_wrap_gen(16))) +
  scale_x_log10(breaks = c(100, 1000),
                labels = label_number(suffix = "M", accuracy = 1)) +
  # Colour carries the attention condition, which is the comparison this figure
  # exists to make; linetype carries model size, which is secondary here.
  scale_color_manual(values = condition_colors, name = NULL) +
  scale_linetype_manual(values = c("solid", "42", "12"), name = NULL) +
  labs(x = "Tokens (M, log scale)", y = "Standardized slope on ordering preference") +
  theme_paper(base_size = 9) +
  theme(legend.position = "bottom", legend.box = "vertical",
        legend.key.width = unit(14, "pt"),
        legend.spacing.y = unit(0, "pt"),
        legend.text = element_text(size = 6.5),
        legend.margin = margin(t = -4),
        strip.placement = "outside",
        strip.text.y.left = element_text(angle = 90, size = 7))
```

# Discussion and Conclusion {#sec-discussion}

The present study examined whether language models' binomial ordering preferences are driven by abstract structure, and if so, how that structure emerges throughout training. Across 16 models from five families, we found that ordering preferences are recoverable and generalize to unseen words, that a model's preference for an attested binomial is governed jointly by what generalizes from binomials it never saw and by the order its own corpus attests for that particular pair, that the balance between the two shifts toward the attested order as a binomial becomes more frequent, and that zeroing out cross-word attention removes roughly half of the attested-order contribution in the large-scale models while the abstract one survives, growing rather than shrinking in every one of them.

Ordering preferences are therefore clearly driven by abstract preferences. The probe recovers a model's preference for binomials built entirely from words it was never trained on, which is not something a lookup of stored items can do, and a ridge probe with a single tuned hyperparameter recovers the same pattern, so the result does not depend on the MLP's capacity (Appendix\ \ref{sec-appendix-linear}), and the answer to the question left open by @houghton2025role is that models do have abstract preferences, but ones a test for human constraints failed to detect. Further evidence that these preferences are systematic rather than an arbitrary consequence of the weights is established without the probe at all: independently trained models converge on the same preferences for the same novel binomials ($r = 0.45$, against a shuffled baseline of $r \approx 0.00$), including the three BabyLM models, which share a training corpus but differ in capacity. Agreement is also higher within a family than between families, which rules out the shared sentence frame as its source: the frame is identical for all sixteen models, so it could not produce such a gap. At the same time, a binomial's own attested order contributes over and above anything the abstract mapping supplies, and contributes more the more often the model saw it. That contribution is item-specific by construction: it is a fact about one pair of words, not a constraint that generalizes. This is the same trade-off between abstract preferences and holistic storage that @morgan2016 found in humans, emerging in a system that does not have them innately encoded in them.

The developmental trajectory separates the two components in time. Abstract preferences are present very early: probe R² is already high at the earliest BabyLM checkpoint, at roughly 20M tokens, and the abstract component of @eq-relfreq is large from the first checkpoint in every model. The stored contribution is not. It is negligible through the first three checkpoints of every model, becomes credibly positive everywhere from the fourth, and climbs monotonically thereafter under full context, without levelling off. Item-specific ordering information is therefore acquired rather than present from the outset, and it is still accumulating at the last point we can observe. What the intervention removes grows too, and it lags the contribution itself. At the first checkpoint where a binomial's attested order predicts the model's preference at all, attention-zeroing removes none of that prediction; the share it removes then climbs to roughly half by the end of training. Item-specific ordering knowledge therefore arrives in the lexical representations before it arrives as holistic storage.

Usage-based accounts have long held that frequently encountered expressions become stored and retrieved as units rather than assembled from their parts, and behavioral work on binomials has shown that human ordering preferences reflect both abstract constraints and direct experience with particular items. The present study corroborates these accounts in a system with no innate linguistic preferences at all: the same frequency-dependent trade-off emerges in a distributed network trained only to predict text, suggesting that it need not be built in, and can emerge from exposure alone. Zeroing out cross-word attention then shows where the stored component lives. Because the intervention makes the representation additive in the two words, it can carry nothing about them jointly, and what it removes is drawn specifically from the contribution of the binomial's own attested order. What it does *not* remove is equally informative. Roughly half of that contribution survives an additive representation, and it is not, at least, the part the probe's abstract prediction already explains, since that prediction is a covariate in the same fit and is computed from the same additive representation. The probe is not a complete account of abstract structure, so some of what survives may be generalizable knowledge it failed to recover. The more interesting possibility is that repeated exposure to a binomial in its own frame shifts the representations of its individual words, so that a word encountered overwhelmingly in first position comes to encode something of that tendency on its own. That is still item-specific knowledge, a fact about particular words rather than a constraint that generalizes, but it is not holistic storage: it has been distilled into the lexical representations and no longer needs the pair. On this reading, entrenchment yields item-specific knowledge in two forms rather than one, and the developmental data orders them: the lexical form is present as soon as the attested order predicts anything at all, and holistic storage accumulates afterwards. The trade-off @morgan2016 describe in humans is then not a single dial between abstract preferences and holistic storage, but a sequence, with the lexicon carrying the earliest item-specific knowledge and the pair representation taking over later. The abstract component survives the same intervention: it is *larger* under attention-zeroing in all `r rf_tot("default", large_fams)` large-scale models, by `r sprintf("%.0f", -max(sapply(large_models, rf_drop, term="pred")))`--`r sprintf("%.0f", -min(sapply(large_models, rf_drop, term="pred")))`%, and smaller by only `r sprintf("%.0f", min(sapply(babylm_labels, rf_drop, term="pred")))`--`r sprintf("%.0f", max(sapply(babylm_labels, rf_drop, term="pred")))`% in the three BabyLM models. A representation that had merely been degraded would have lost both. Repeated exposure does not simply add a stored preference alongside an abstract one: it builds something that requires the two words to be encoded together, and that can be switched off without disturbing what generalizes.

# Limitations {.unnumbered}

The novel binomials were selected to be absent from BabyLM's training corpus, so they are genuinely novel only for the three BabyLM models; indexed against Dolma [@dolma], a trillion-token corpus, roughly `r dolma_pct_nonzero`% occur at non-zero frequency. For the thirteen large-scale models the probe is therefore trained on infrequent attested binomials rather than unseen ones. This is why Experiment 1 anchors the interpretation: the inference that a recovered preference must have been generalized rather than retrieved rests on BabyLM, and the larger models extend that pattern rather than establish it. Corpus frequency is likewise an exact measure of what the model saw only for BabyLM and Pythia; for GPT-2, OLMo, and Llama, Pile counts stand in as a proxy for web-scale exposure.

The stored-ordering measure is estimated from counts, so it is least reliable exactly where exposure is lowest. A binomial seen once takes the same boundary value as one seen a thousand times in a single order, which makes the predictor coarse for rare items and attenuates its coefficient there. The effect is strongest for BabyLM, where most items sit at a boundary and the measure is close to binary, and mild for the large-scale models, where it varies continuously through its range. That the two return moderation estimates of similar size argues against the coefficient being mostly an artifact of saturation, but attenuation and growing storage do push it in the same direction, so part of any positive estimate reflects improving measurement. This is why we read the comparison between attention conditions rather than the moderation on its own, since the counts, and therefore the attenuation, are identical in both conditions and cancel from the difference.

Separately, BabyLM and Pythia deliver cumulative exposure to individual binomials at nearly the same rate per training token, so total text read and per-binomial exposure cannot be told apart here; distinguishing them would require matched models trained on repeated versus varied text.

Probe performance establishes that ordering information is present in a model's representations and recoverable by a mapping that generalizes to unseen words, but not that the model uses that information in the way the probe reads it. Our central claims are accordingly framed as differences between attention conditions and across frequency strata rather than as inferences from absolute decodability, and the attention-zeroed quantity is guaranteed by the architecture rather than estimated by the probe. This distinction does real work: a representation that merely carried the outcome of the computation would be read equally well at every frequency, predicting neither a shift in which source governs the ordering nor any change in that shift when cross-word attention is blocked, and both are observed in every large-scale model. Overall R² does also differ between the two attention conditions in a direction that is not stable across families or across training, and we draw no conclusions from it.

Finally, the results are limited to English and to a single construction, and checkpoints were available for two of the five families, so the developmental claims rest on BabyLM and Pythia. Whether attention-zeroing separates per-word from relational information as cleanly in constructions with a different structure is untested, though the argument that it should is architectural rather than empirical and does not depend on properties peculiar to binomials.

```{=latex}
\bibliography{references}
\clearpage
```

\onecolumn

\appendix

```{=latex}
% Appendix layout. Two separate things are going on here.
%
% \raggedbottom is the important one. acl.sty sets \flushbottom, which forces
% every page to exact height by stretching the flexible glue between paragraphs
% and around headings. On an appendix page that cannot fill naturally, that
% stretch becomes a gap in the middle of the page with text stranded above and
% below it. \raggedbottom instead lets the slack collect at the foot of the
% page, which is the conventional choice for appendix matter. Main-text pages
% keep \flushbottom and the two-column look.
%
% Tables are not floats here at all: every appendix table is emitted as a
% longtable, which typesets inline and breaks across pages. That is what keeps
% them exactly where they are declared. A float cannot do this: Table 9 alone
% is about 640pt against a 706pt text block, so as a float it could never share
% a page with the heading and prose that introduce it, and [H] simply pushed it
% to the next page and left the previous one blank.
%
% Figures are still floats and are pinned with [H], which is fine because none
% of them needs to break across a page.
\raggedbottom
\floatplacement{figure}{H}
```

# Model Details {#sec-appendix-models}

Table \ref{tbl-models} lists all 16 models used across experiments, organized by family, with approximate parameter counts and training token counts.

```{r}
#| label: tbl-models
#| tbl-cap: "Models used across experiments. Training token counts for full-scale models are approximate; GPT-2 counts are estimated from the reported 40GB WebText corpus. BabyLM models train for 20 epochs over 150M tokens (3B effective token exposures)."

model_table <- tribble(
  ~Family,  ~Model,          ~Parameters, ~`Tokens`,
  "BabyLM", "BabyLM-125M",  "125M",      "150M",
  "BabyLM", "BabyLM-350M",  "350M",      "150M",
  "BabyLM", "BabyLM-1.3B",  "1.3B",      "150M",
  "GPT-2",  "GPT-2",        "117M",      "~8B",
  "GPT-2",  "GPT-2-medium", "345M",      "~8B",
  "GPT-2",  "GPT-2-large",  "762M",      "~8B",
  "GPT-2",  "GPT-2-XL",     "1.5B",      "~8B",
  "Pythia", "Pythia-160M",  "160M",      "~300B",
  "Pythia", "Pythia-410M",  "410M",      "~300B",
  "Pythia", "Pythia-1B",    "1B",        "~300B",
  "OLMo",   "OLMo-1B",      "1B",        "3T",
  "OLMo",   "OLMo-7B",      "7B",        "2.5T",
  "OLMo",   "OLMo-2-1B",    "1B",        "~4T",
  "OLMo",   "OLMo-2-7B",    "7B",        "~4T",
  "Llama",  "Llama-1.3B",   "1.2B",      "~9T",
  "Llama",  "Llama-3-8B",   "8B",        "~15T"
)

model_table |>
  kbl(longtable = TRUE, booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 9, latex_options = c("repeat_header")) |>
  collapse_rows(columns = 1, latex_hline = "major") |>
  allow_group_breaks()
```

\clearpage

# Binomial Extraction Criteria {#sec-appendix-extraction}

Candidate "$W_1$ and $W_2$" sequences were first selected using spaCy [@spacy2] part-of-speech tags, retaining only cases where both $W_1$ and $W_2$ are open-class words (nouns, verbs, adjectives, or adverbs). Each candidate was then parsed with the Berkeley Neural Parser [@kitaev-klein-2018-constituency; @kitaev-etal-2019-multilingual] to confirm that the two words form a minimal coordinate phrase. The inclusion criterion was that the two words be co-constituents under the same phrasal node (NP, VP, ADJP, ADVP, PP, or equivalent), with that node spanning exactly the three tokens \[$W_1$, and, $W_2$\] and nothing else. This ensures the extracted items are genuine coordinate binomials rather than incidental sequences where "and" links clauses or longer phrases. Novel binomials were extracted from Wikitext by the identical procedure, with the added requirements that both words appear in BabyLM's vocabulary and that the binomial is absent from the attested set.

Because cross-category coordination is syntactically unacceptable, violating the Coordination of Likes constraint that benepar enforces through its Penn Treebank training, genuine mixed-category binomials are excluded by the constituency criterion rather than by any explicit POS filter. We deliberately avoid filtering by spaCy POS tag agreement because spaCy frequently misclassifies participial forms in coordinate contexts, for example tagging "mortified" as VERB rather than ADJ in "disgusted and mortified." Explicit POS filtering would therefore discard many valid same-category binomials without catching any genuine cross-category ones.

\clearpage

# Implementation Details {#sec-appendix-implementation}

**Probe training.** The probe is optimized with Adam (learning rate $10^{-3}$, weight decay $10^{-4}$) under mean-squared-error loss, with early stopping (patience 20) against a 10% held-out validation split. Antisymmetric augmentation is applied with probability 0.5: for each binomial the flipped version, with swapped ordering and negated label, is added to the training batch. The input dimensionality follows from the representation type as defined in @sec-methods.

**Attention-zeroing.** The mask is supplied to the model as a 4D attention mask rather than a 2D padding mask, so it applies at every transformer layer rather than only the first. This matters because a mask applied at the first layer alone would leave $W_2$ able to recover information about $W_1$ indirectly at greater depth, through positions that did attend to it.

**Recomputing the target.** Because $y_{true}$ is the model's own log-probability ratio, it is recomputed from whichever forward pass produced the representations being probed. The two full-context representation types therefore share a target, while the attention-zeroed condition has its own: blocking cross-word attention changes the model's output probabilities as well as its hidden states. Each attention condition is thus a self-consistent probe problem rather than two input encodings evaluated against a fixed target.

\clearpage

# Linear Probe Replication {#sec-appendix-linear}

The MLP probe has many free choices: hidden dimension, depth, activation, learning rate, weight decay, patience, validation fraction, augmentation probability, batch size, and initialization. A skeptical reading is that some of them were tuned toward the reported result. We therefore repeat the analysis with a probe that has one.

The linear probe is a ridge regression on the same antisymmetric features, fit on the same cross-validation folds. It has no intercept, since the antisymmetric feature is odd by construction and the target has zero mean at the origin, and no optimization schedule, since the solution is closed-form. Its only free quantity is the ridge penalty, selected per fold from $\{10^0, \ldots, 10^5\}$ on an inner split of that fold's training data. That inner split is the same size as the MLP's early-stopping split, so both probes receive an identical budget of held-out data for hyperparameter selection.

```{r}
#| label: linear-probe-load
#| include: false

sup_dir <- file.path(RES, "..", "Data", "supplementary")
lin_raw <- if (dir.exists(sup_dir)) {
  cells <- setdiff(basename(list.dirs(sup_dir, recursive = FALSE)), "logs")
  purrr::map_dfr(cells, function(s) {
    f <- file.path(sup_dir, s, "linear_probe.csv")
    if (!file.exists(f) || length(readLines(f, warn = FALSE)) < 2) return(NULL)
    readr::read_csv(f, show_col_types = FALSE,
                    col_types = readr::cols(layer = readr::col_character(),
                                            mean_r2 = readr::col_double())) |>
      mutate(slug = s)
  })
} else tibble()

lin_tbl <- if (nrow(lin_raw)) {
  lin <- lin_raw |>
    filter(mode == "mean_pooled") |>
    mutate(L = suppressWarnings(as.integer(layer))) |>
    group_by(slug, condition, split) |>
    slice_max(L, n = 1, with_ties = FALSE) |> ungroup() |>
    select(slug, condition, split, lin = mean_r2)
  mlp <- purrr::map_dfr(unique(lin$slug), function(s) {
    f <- file.path(RES, s, "by_layer_mlp.csv")
    if (!file.exists(f)) return(NULL)
    readr::read_csv(f, show_col_types = FALSE,
                    col_types = readr::cols(layer = readr::col_character(),
                                            mean_r2 = readr::col_double())) |>
      filter(mode == "mean_pooled") |>
      mutate(L = suppressWarnings(as.integer(layer))) |>
      group_by(condition, split) |> slice_max(L, n = 1, with_ties = FALSE) |>
      ungroup() |> transmute(slug = s, condition, split, mlp = mean_r2)
  })
  inner_join(lin, mlp, by = c("slug", "condition", "split")) |>
    mutate(gap = mlp - lin, is_ckpt = grepl("_step", slug))
} else tibble()
```

```{r}
#| label: linear-probe-report
#| echo: false
#| results: asis

if (!nrow(lin_tbl)) {
  cat("_Linear probe results not yet available; re-run",
      "`Scripts/supplementary/supplementary_analyses.sh`._\n")
} else {
  n_cells <- dplyr::n_distinct(lin_tbl$slug)
  r_all   <- cor(lin_tbl$lin, lin_tbl$mlp)
  wn      <- dplyr::filter(lin_tbl, split == "word_novel")
  cat(sprintf(paste0("Across %d model-checkpoint cells and %d comparisons, ",
      "the two probes agree closely (*r* = %.3f between their R² values). ",
      "On the word-held-out split the linear probe recovers %.0f%% of the ",
      "MLP's R² on average (%.3f vs. %.3f).\n\n"),
      n_cells, nrow(lin_tbl), r_all,
      100 * mean(wn$lin) / mean(wn$mlp), mean(wn$lin), mean(wn$mlp)))
}
```

```{r}
#| label: tbl-linear
#| tbl-cap: "Linear (ridge) versus MLP probe R² at the final hidden layer, binomial-level input mode. Both probes use identical cross-validation folds. B = binomial-held-out; W = word-held-out. Gap = MLP minus linear."

if (nrow(lin_tbl)) {
  lin_tbl |>
    filter(!is_ckpt) |>
    mutate(Model = slug |>
             stringr::str_remove("EleutherAI_|allenai_|meta-llama_") |>
             stringr::str_replace("znhoughton_opt-babylm-", "BabyLM-") |>
             stringr::str_remove("-20eps-seed964"),
           Condition = recode(condition, !!!condition_labels),
           Split = if_else(split == "pair_novel", "B", "W"),
           Linear = sprintf("%.3f", lin), MLP = sprintf("%.3f", mlp),
           Gap = sprintf("%+.3f", gap)) |>
    arrange(Model, Condition, Split) |>
    select(Model, Condition, Split, Linear, MLP, Gap) |>
    kbl(longtable = TRUE, booktabs = TRUE, linesep = "") |>
    kable_styling(font_size = 8, latex_options = c("repeat_header")) |>
    collapse_rows(columns = 1:2, latex_hline = "major") |>
  allow_group_breaks()
} else {
  kbl(longtable = TRUE, tibble(Note = "pending"), booktabs = TRUE)
}
```

The gap between the probes is consistently larger on the binomial-held-out split than on the word-held-out split. This is expected: binomial-held-out performance is partly recoverable from word identity alone, which a flexible model can exploit and a linear map on the same features cannot. Word-held-out performance, on which the paper's claims rest, is where the two agree most closely.

\clearpage

# Cross-Validation Split Sizes {#sec-appendix-cv}

Cross-validation is carried out on the novel set only. The attested set is never cross-validated: it is used solely as the transfer target described in @sec-methods-analyses, where a probe trained on the full novel set predicts every attested binomial and those predictions enter the ordering-source regression as $\hat{y}$. Both splits use ten folds with seed 964, and the same folds are reused across models, attention conditions, and probe architectures so that all reported R² values are directly comparable.

Under the binomial-held-out split, binomials are assigned to folds directly, so each fold trains on nine tenths of the set and tests on the remaining tenth. Under the word-held-out split it is the *vocabulary* that is partitioned into ten folds. A binomial enters the test set for fold $k$ only if both of its words are in fold $k$, and enters the training set only if neither word is. Binomials with one word in fold $k$ and one word outside it are used in neither, and are discarded for that fold. Since a binomial needs both words drawn from the same tenth of the vocabulary to be testable, the word-held-out test set is roughly a hundredth of the set rather than a tenth. @tbl-cv-sizes gives the resulting sizes.

```{r}
#| label: tbl-cv-sizes
#| tbl-cap: "Cross-validation fold sizes for the novel set (340,042 binomials over 50,137 unique words). Ranges are across the 10 folds. Discarded items under word-held-out are binomials with exactly one word in the held-out vocabulary fold."

tibble::tribble(
  ~Split,                ~`Train / fold`,     ~`Test / fold`,   ~`Discarded / fold`,
  "Binomial-held-out",   "306,038",           "34,004",         "0",
  "Word-held-out",       "271,560--279,306",  "3,086--3,839",   "61,101 (18.0\\%)"
) |>
  kbl(longtable = TRUE, booktabs = TRUE, linesep = "", escape = FALSE) |>
  kable_styling(font_size = 9, latex_options = c("repeat_header"))
```

No fold under either split fell below the minimum test-set size of 10 binomials required for a fold to be scored, so all ten folds contribute to every reported mean. For reference, the attested set to which the trained probe is transferred contains 48,965 binomials over 21,721 unique words, all of which receive a prediction.

\clearpage

# Experiment 1: Supplementary Results {#sec-appendix-exp1}

This appendix provides supplementary materials for Experiment 1 (§\ref{sec-exp1}), which examined final-layer probe performance for the three BabyLM models. Subsections report full R² tables, control probe results, and a word-level replication. Coefficients for the ordering-source regression are reported together for all models in Appendix\ \ref{sec-appendix-relfreq}.

## Full Probe R² {#sec-appendix-full-r2-babylm}

Probe R² at the final hidden layer for BabyLM models, at the binomial level by attention condition. B = binomial-held-out; W = word-held-out. SD across 10 folds in parentheses.

```{r}
#| label: tbl-r2-babylm
#| tbl-cap: "Probe R² at the final hidden layer for BabyLM models, at the binomial level by attention condition. B = binomial-held-out split; W = word-held-out split. SD across 10 folds in parentheses."
#| eval: !expr nrow(all_r2) > 0

all_r2 |>
  filter(family == "BabyLM", mode == "mean_pooled") |>
  select(model, condition, split, mean_r2, sd_r2, is_control) |>
  pivot_wider(names_from  = is_control,
              values_from = c(mean_r2, sd_r2),
              names_sep   = "_") |>
  rename(r2_main = mean_r2_FALSE, sd_main = sd_r2_FALSE,
         r2_ctrl = mean_r2_TRUE,  sd_ctrl = sd_r2_TRUE) |>
  mutate(
    condition = recode_condition(condition),
    model     = factor(model, levels = model_order),
    Split     = if_else(split == "pair_novel", "B", "W"),
    `R²`      = sprintf("%.3f (%.3f)", r2_main, sd_main),
    Ctrl      = sprintf("%.3f (%.3f)", r2_ctrl,  sd_ctrl),
  ) |>
  arrange(model, condition, split) |>
  select(Model = model, Condition = condition, Split, `R²`, Ctrl) |>
  kbl(longtable = TRUE, booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 8, latex_options = c("repeat_header")) |>
  collapse_rows(columns = 1:2, latex_hline = "major") |>
  allow_group_breaks()
```

## Control Probe R² {#sec-appendix-ctrl-exp1}

Control probe R² for BabyLM models at the final hidden layer, at the binomial level by attention condition. Labels are randomly permuted; values near zero confirm the main probe recovers genuine linguistic structure rather than label artifacts or probe capacity.

```{r}
#| label: tbl-ctrl-r2-babylm
#| tbl-cap: "Control probe R² for BabyLM models (final layer), at the binomial level by attention condition. SD across 10 folds in parentheses. B = binomial-held-out; W = word-held-out."

all_r2 |>
  filter(family == "BabyLM", mode == "mean_pooled", is_control) |>
  mutate(
    condition    = recode(condition, !!!condition_labels),
    model        = factor(model, levels = model_order),
    Split        = if_else(split == "pair_novel", "B", "W"),
    `Control R²` = sprintf("%.4f (%.4f)", mean_r2, sd_r2)
  ) |>
  arrange(model, condition, split) |>
  select(Model = model, Condition = condition, Split, `Control R²`) |>
  kbl(longtable = TRUE, booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 8, latex_options = c("repeat_header")) |>
  collapse_rows(columns = 1:2, latex_hline = "major") |>
  allow_group_breaks()
```

## Cross-Model Agreement on Novel Binomials {#sec-appendix-agreement}

A binomial absent from the training corpus cannot have a retrieved preference, but the model still assigns one, and that value is some deterministic function of the two words' representations. Such a function could reflect systematic knowledge, or it could be an arbitrary consequence of a particular set of weights. Probe generalization cannot distinguish these, because both are systematic functions of the input and both transfer to held-out words. Agreement between independently trained models can: an arbitrary function of one model's weights has no reason to match another's.

We therefore correlate $y_{true}$ across models on identical binomials, separately for the attested and novel sets and for both attention conditions. As a baseline we shuffle one model's values, which removes any pairing between models while preserving both distributions.

The three BabyLM models are the decisive comparison. They share a training corpus and its ordering but differ in capacity, and they are the only models for which the novel binomials are genuinely absent from training (@sec-exp2, Methods). On novel binomials they agree at $r = 0.45$ under full context and $r = 0.41$ under attention-zeroing, against a shuffled baseline of $r \approx 0.00$. Preferences for binomials the models never encountered are therefore reproducible across models of different sizes trained on the same data, which rules out their being arbitrary.

Because a Pearson correlation can in principle be carried by a minority of high-magnitude items, we checked whether the agreement is concentrated in strong preferences while the bulk of binomials agree at chance. It is not. Preferences on novel binomials are not weak to begin with: median $|y_{true}|$ ranges from 1.79 to 3.21 across the three models, so the preferred ordering is between roughly six and twenty-five times more likely, and only 9% to 17% of binomials fall below 0.5. Rank correlations track the Pearson values closely (0.371–0.406 for BabyLM on novel binomials), and the models order the same binomial the same way on 60% to 70% of items against a chance rate of 50%, a statistic that ignores magnitude and weights every binomial equally. Stratifying by preference strength, using the third model's $|y_{true}|$ so that the bins are independent of the two series being correlated, agreement is graded but present throughout: in the weakest quintile $r = 0.30$ under full context and $0.28$ under attention-zeroing, with directional agreement near 59% in both, rising to $r = 0.61$ and 73% in the strongest. That gradient is itself informative, since agreement increasing with preference strength is the signature of a signal measured with noise rather than of an arbitrary function of a particular set of weights.

Two further patterns appear in @tbl-agreement. Agreement within a family is roughly constant across all four cells, near $r = 0.68$, whereas agreement between families falls from 0.56 on attested binomials under full context to 0.28 on novel binomials under attention-zeroing. What the models share thus survives novelty when the training corpus is shared and degrades when it is not, as expected if the preference is driven by distributional structure in the training data. This also rules out the shared sentence frame as the source of the agreement: the frame is identical for all sixteen models, so it cannot produce a gap between within-family and between-family agreement.

```{r}
#| label: tbl-agreement
#| tbl-cap: "Cross-model agreement (Pearson $r$ on $y_{true}$) for identical binomials, by item set and attention condition. Shuffled = one model's values permuted. Within-family averages over model pairs from the same family; BabyLM is broken out because its novel binomials are genuinely unseen."

read_csv(file.path(RES, "cross_model_agreement_summary.csv"),
         show_col_types = FALSE) |>
  filter(scope %in% c("all pairs of models", "within family",
                      "between family", "within BabyLM")) |>
  mutate(
    Set       = if_else(item_set == "novel", "Novel", "Attested"),
    Condition = recode(condition, !!!condition_labels),
    Scope     = recode(scope,
                       "all pairs of models" = "All",
                       "within family"       = "Within family",
                       "between family"      = "Between family",
                       "within BabyLM"       = "Within BabyLM"),
    Scope     = factor(Scope, levels = c("All", "Within family",
                                         "Between family", "Within BabyLM")),
    Set       = factor(Set, levels = c("Attested", "Novel")),
    `$r$`     = sprintf("%.3f", mean_r),
    Shuffled  = sprintf("%.3f", shuffled_null_r)
  ) |>
  arrange(Set, Condition, Scope) |>
  select(Set, Condition, Scope, `$r$`, Shuffled) |>
  kbl(longtable = TRUE, booktabs = TRUE, linesep = "", escape = FALSE) |>
  kable_styling(font_size = 8, latex_options = c("repeat_header")) |>
  collapse_rows(columns = 1:2, latex_hline = "major") |>
  allow_group_breaks()
```

## Word-Level Replication {#sec-appendix-words-only-exp1}

We trained the MLP probe using word-level representations for the three BabyLM models to verify that the preference signal is not an artifact of mean pooling across the full binomial phrase. Under this condition the probe receives a concatenation of two word vectors extracted from isolated sentence contexts ("*the bread*", "*the butter*"), so it cannot rely on any co-occurrence signal from the two words' shared sentence context.

Binomial-held-out R² ranges from 0.31 to 0.33, within a few percentage points of the corresponding binomial-level values (0.29–0.34). The generalization pattern is preserved: word-held-out R² is consistently lower than binomial-held-out R² for all three models (word-held-out 0.18–0.20; binomial-minus-word gap 0.127–0.129), confirming that the lexical generalization cost seen at the binomial level is a property of the preference encoding itself rather than an artifact of the mean-pooled representation. The ordering-source regression was fit at the binomial level only, so it has no word-level counterpart. Full results are in @tbl-r2-babylm-words-only.

```{r}
#| label: tbl-r2-babylm-words-only
#| tbl-cap: "Word-level probe R² and selectivity (main $-$ control R²) for BabyLM models (final layer). B = binomial-held-out split; W = word-held-out split. SD across 10 folds in parentheses."

all_r2 |>
  filter(family == "BabyLM", mode == "words_only", condition == "default") |>
  select(model, condition, mode, split, mean_r2, sd_r2, is_control) |>
  pivot_wider(names_from  = is_control,
              values_from = c(mean_r2, sd_r2),
              names_sep   = "_") |>
  rename(r2_main = mean_r2_FALSE, sd_main = sd_r2_FALSE,
         r2_ctrl = mean_r2_TRUE,  sd_ctrl = sd_r2_TRUE) |>
  mutate(
    selectivity = r2_main - r2_ctrl,
    mode        = recode(mode, !!!mode_labels),
    model       = factor(model, levels = model_order),
    Split       = if_else(split == "pair_novel", "B", "W"),
    `R²`        = sprintf("%.3f (%.3f)", r2_main, sd_main),
    Ctrl        = sprintf("%.3f (%.3f)", r2_ctrl,  sd_ctrl),
    Sel.        = sprintf("%.3f", selectivity)
  ) |>
  arrange(model, split) |>
  select(Model = model, Mode = mode, Split, `R²`, Ctrl, Sel.) |>
  kbl(longtable = TRUE, booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 8, latex_options = c("repeat_header")) |>
  collapse_rows(columns = 1:2, latex_hline = "major") |>
  allow_group_breaks()
```

# Experiment 2: Supplementary Results {#sec-appendix-exp2}

This appendix provides supplementary materials for Experiment 2 (§\ref{sec-exp2}), which examined final-layer representations across all 16 models from five families. Subsections report full R² tables, control probe results, and a word-level replication. Coefficients for the ordering-source regression are in Appendix\ \ref{sec-appendix-relfreq}.

## Full Probe R² {#sec-appendix-full-r2}

Probe R² for all 16 models at the final hidden layer, binomial-level input mode, by attention condition and cross-validation split. B = binomial-held-out; W = word-held-out. SD across 10 folds in parentheses.

```{r}
#| label: tbl-r2
#| tbl-cap: "Probe R² at the final hidden layer for all 16 models, at the binomial level by attention condition. B = binomial-held-out split; W = word-held-out split. SD across 10 folds in parentheses."
#| eval: !expr nrow(all_r2) > 0

all_r2 |>
  filter(mode == "mean_pooled") |>
  select(model, family, condition, mode, split, mean_r2, sd_r2, is_control) |>
  pivot_wider(names_from  = is_control,
              values_from = c(mean_r2, sd_r2),
              names_sep   = "_") |>
  rename(r2_main = mean_r2_FALSE, sd_main = sd_r2_FALSE,
         r2_ctrl = mean_r2_TRUE,  sd_ctrl = sd_r2_TRUE) |>
  mutate(
    condition = recode_condition(condition),
    model     = factor(model, levels = model_order),
    Split     = if_else(split == "pair_novel", "B", "W"),
    `R²`      = sprintf("%.3f (%.3f)", r2_main, sd_main),
    Ctrl      = sprintf("%.3f (%.3f)", r2_ctrl,  sd_ctrl),
  ) |>
  arrange(model, condition, split) |>
  select(Model = model, Condition = condition, Split, `R²`, Ctrl) |>
  kbl(longtable = TRUE, booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 7, latex_options = c("scale_down", "repeat_header")) |>
  collapse_rows(columns = 1:2, latex_hline = "major") |>
  allow_group_breaks()
```

## Control Probe R² {#sec-appendix-ctrl-exp2}

Control probe R² for all 16 models at the final hidden layer, at the binomial level by attention condition.

```{r}
#| label: tbl-ctrl-r2
#| tbl-cap: "Control probe R² for all 16 models (final layer), at the binomial level by attention condition. SD across 10 folds in parentheses. B = binomial-held-out; W = word-held-out."

all_r2 |>
  filter(mode == "mean_pooled", is_control) |>
  mutate(
    condition    = recode(condition, !!!condition_labels),
    model        = factor(model, levels = model_order),
    Split        = if_else(split == "pair_novel", "B", "W"),
    `Control R²` = sprintf("%.4f (%.4f)", mean_r2, sd_r2)
  ) |>
  arrange(family, model, condition, split) |>
  select(Family = family, Model = model, Condition = condition, Split, `Control R²`) |>
  kbl(longtable = TRUE, booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 8, latex_options = c("repeat_header")) |>
  collapse_rows(columns = 1:3, latex_hline = "major") |>
  allow_group_breaks()
```

## Word-Level Replication {#sec-appendix-words-only-exp2}

Across all 16 models, word-level probing reproduces the main findings from Experiment 2. Binomial-held-out R² is slightly lower than the corresponding binomial-level values (by roughly 5–10 percentage points for most large-scale models), but rank order across families and the generalization pattern (binomial-held-out \> word-held-out) are preserved in every case. For GPT-2, binomial-held-out R² under word-level probing ranges from 0.46 to 0.62 (vs. 0.49–0.65 at the binomial level); for Pythia, 0.51–0.57 (vs. 0.53–0.59); for OLMo, 0.52–0.57 (vs. 0.56–0.62); for Llama, 0.36–0.45 (vs. 0.43–0.55). The generalization gaps (binomial-held-out minus word-held-out) range from 0.044 to 0.073 across large-scale models, closely matching the corresponding binomial-level gaps (0.035–0.080). The ordering-source regression was fit at the binomial level only, so it has no word-level counterpart. Full results are in @tbl-r2-words-only.

```{r}
#| label: tbl-r2-words-only
#| tbl-cap: "Word-level probe R² and selectivity (main $-$ control R²) for all 16 models (final layer). B = binomial-held-out split; W = word-held-out split. SD across 10 folds in parentheses."

all_r2 |>
  filter(mode == "words_only", condition == "default") |>
  select(family, model, condition, mode, split, mean_r2, sd_r2, is_control) |>
  pivot_wider(names_from  = is_control,
              values_from = c(mean_r2, sd_r2),
              names_sep   = "_") |>
  rename(r2_main = mean_r2_FALSE, sd_main = sd_r2_FALSE,
         r2_ctrl = mean_r2_TRUE,  sd_ctrl = sd_r2_TRUE) |>
  mutate(
    selectivity = r2_main - r2_ctrl,
    mode        = recode(mode, !!!mode_labels),
    model       = factor(model, levels = model_order),
    Split       = if_else(split == "pair_novel", "B", "W"),
    `R²`        = sprintf("%.3f (%.3f)", r2_main, sd_main),
    Ctrl        = sprintf("%.3f (%.3f)", r2_ctrl,  sd_ctrl),
    Sel.        = sprintf("%.3f", selectivity)
  ) |>
  arrange(family, model, split) |>
  select(Family = family, Model = model, Mode = mode, Split, `R²`, Ctrl, Sel.) |>
  kbl(longtable = TRUE, booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 8, latex_options = c("repeat_header")) |>
  collapse_rows(columns = 1:3, latex_hline = "major") |>
  allow_group_breaks()
```

\clearpage

# Experiment 3: Supplementary Results {#sec-appendix-exp3}

This appendix provides supplementary materials for Experiment 3 (§\ref{sec-exp3}), which tracked probe R² and the ordering-source regression across training checkpoints for BabyLM and Pythia. Subsections report full numerical checkpoint tables and a word-level replication of the probe results.

## Checkpoint Probe Tables {#sec-appendix-ckpt-tables}

Full numerical results for the Experiment 3 checkpoint analysis. Tables report probe R² (binomial-held-out and word-held-out) and the coefficients of @eq-relfreq for BabyLM and Pythia across training checkpoints, for the binomial-level (mean-pooled) representation under full-context and attention-zeroed conditions.

```{r}
#| echo: false

make_ckpt_r2_tbl <- function(df, cond_val, mode_val) {
  df |>
    filter(condition == cond_val, mode == mode_val,
           split %in% c("pair_novel", "word_novel")) |>
    select(model, step, tokens_M, split, mean_r2, sd_r2, is_control) |>
    pivot_wider(names_from = is_control, values_from = c(mean_r2, sd_r2),
                names_sep = "_") |>
    rename(r2_main = mean_r2_FALSE, sd_main = sd_r2_FALSE,
           r2_ctrl = mean_r2_TRUE,  sd_ctrl = sd_r2_TRUE) |>
    mutate(
      model  = factor(model, levels = model_order),
      Split  = if_else(split == "pair_novel", "B", "W"),
      Tokens = sprintf("%.0f M", tokens_M),
      `R²`   = sprintf("%.3f (%.3f)", r2_main, sd_main),
      Ctrl   = sprintf("%.3f (%.3f)", r2_ctrl,  sd_ctrl),
      Sel.   = sprintf("%.3f", r2_main - r2_ctrl)
    ) |>
    arrange(model, step, split) |>
    select(Model = model, Step = step, Tokens, Split, `R²`, Ctrl, Sel.) |>
    kbl(longtable = TRUE, booktabs = TRUE, linesep = "") |>
    kable_styling(font_size = 8, latex_options = c("repeat_header")) |>
    collapse_rows(columns = 1:2, latex_hline = "major") |>
  allow_group_breaks()
}

```

```{r}
#| label: tbl-r2-checkpoints-babylm-default-mean
#| tbl-cap: "BabyLM checkpoint probe R²:full context, binomial-level. B = binomial-held-out; W = word-held-out. SD in parentheses."

if (nrow(babylm_ckpt_r2) > 0)
  make_ckpt_r2_tbl(babylm_ckpt_r2, "default", "mean_pooled")
```

```{r}
#| label: tbl-r2-checkpoints-babylm-attn-mean
#| tbl-cap: "BabyLM checkpoint probe R²:attention-zeroed, binomial-level."

if (nrow(babylm_ckpt_r2) > 0)
  make_ckpt_r2_tbl(babylm_ckpt_r2, "attn_zeroed", "mean_pooled")
```

```{r}
#| label: tbl-r2-checkpoints-pythia-default-mean
#| tbl-cap: "Pythia checkpoint probe R²:full context, binomial-level."

if (nrow(pythia_ckpt_r2) > 0)
  make_ckpt_r2_tbl(pythia_ckpt_r2, "default", "mean_pooled")
```

```{r}
#| label: tbl-r2-checkpoints-pythia-attn-mean
#| tbl-cap: "Pythia checkpoint probe R²:attention-zeroed, binomial-level."

if (nrow(pythia_ckpt_r2) > 0)
  make_ckpt_r2_tbl(pythia_ckpt_r2, "attn_zeroed", "mean_pooled")
```

```{r}
#| label: tbl-relfreq-checkpoints
#| tbl-cap: "Ordering-source regression (@eq-relfreq) at every training checkpoint, both attention conditions. All four variables are z-scored within each fit, so coefficients are standardized betas comparable across checkpoints and models. 95\\% credible intervals in brackets."
#| eval: !expr nrow(relfreq_ckpt) > 0

tt2 <- function(x) paste0("\\texttt{", gsub("_", "\\\\_", x), "}")

relfreq_ckpt |>
  arrange(model, condition != "default", step) |>
  mutate(Condition = recode_condition(condition),
         a = sprintf("%.3f [%.3f, %.3f]", pred_mean, pred_lo, pred_hi),
         b = sprintf("%.3f [%.3f, %.3f]", rel_mean,  rel_lo,  rel_hi),
         c = sprintf("%.3f [%.3f, %.3f]", pred_x_mean, pred_x_lo, pred_x_hi),
         d = sprintf("%.3f [%.3f, %.3f]", rel_x_mean,  rel_x_lo,  rel_x_hi)) |>
  select(Model = model, Step = step, Condition, a, b, c, d) |>
  kbl(longtable = TRUE, booktabs = TRUE, linesep = "", escape = FALSE,
      col.names = c("Model", "Step", "Condition", tt2("y_pred"), tt2("rel_freq"),
                    paste(tt2("y_pred"), "$\\times$", tt2("log_freq")),
                    paste(tt2("rel_freq"), "$\\times$", tt2("log_freq")))) |>
  kable_styling(font_size = 6, latex_options = c("scale_down", "repeat_header")) |>
  collapse_rows(columns = 1, latex_hline = "major") |>
  allow_group_breaks()
```

## Word-Level Replication {#sec-appendix-words-only-exp3}

Word-level probe results across BabyLM and Pythia checkpoints replicate the key patterns from the binomial-level analysis. The generalization gap (binomial-held-out minus word-held-out) is largest at the earliest checkpoints and shrinks over the course of training. For BabyLM, the gap at the first checkpoint ranges from 0.15 to 0.36 across model sizes, narrowing to 0.13–0.14 by the final checkpoint, the same qualitative trajectory observed at the binomial level. Full R² results are in @tbl-r2-checkpoints-babylm-default-words and @tbl-r2-checkpoints-pythia-default-words. The ordering-source regression was fit at the binomial level only, so it has no word-level counterpart here; its checkpoint coefficients are in @tbl-relfreq-checkpoints.

```{r}
#| label: tbl-r2-checkpoints-babylm-default-words
#| tbl-cap: "BabyLM checkpoint probe R²:full context, word-level. B = binomial-held-out; W = word-held-out. SD in parentheses."

if (nrow(babylm_ckpt_r2) > 0)
  make_ckpt_r2_tbl(babylm_ckpt_r2, "default", "words_only")
```

```{r}
#| label: tbl-r2-checkpoints-pythia-default-words
#| tbl-cap: "Pythia checkpoint probe R²:full context, word-level. B = binomial-held-out; W = word-held-out. SD in parentheses."

if (nrow(pythia_ckpt_r2) > 0)
  make_ckpt_r2_tbl(pythia_ckpt_r2, "default", "words_only")
```

\clearpage

# Corpus Frequency Distributions {#sec-appendix-corpus-freq}

The figure below compares the frequency distributions of the approximately 47,000 corpus binomials in the BabyLM training corpus (150M tokens, 20 epochs) and the Pile, illustrating the stark difference in lexical coverage between the two training regimes. BabyLM's distribution is dominated by singleton binomials (70.7% appear exactly once), while Pile frequencies span several orders of magnitude.

```{r}
#| label: fig-corpus-freq
#| fig-pos: "H"
#| fig-cap: "Frequency distributions for the ~47K corpus binomials in the BabyLM training corpus (150M tokens, 20 epochs) and the Pile. Left: histogram of log$_{10}$ raw counts. Right: rank-frequency plot treating counts as a probability distribution (probability = count / $\\sum$ counts); both axes log-scaled. BabyLM's distribution is heavily concentrated at a single occurrence per pair (70.7% of pairs appear exactly once), whereas Pile frequencies span several orders of magnitude with a median of 365 occurrences per pair."
#| fig-width: 7
#| fig-height: 3.2

corpus_freq_df <- local({
  babylm_raw <- read_csv("../Data/corpus_binomials.csv", show_col_types = FALSE) |>
    mutate(freq = freq_w1_w2 + freq_w2_w1) |>
    filter(freq > 0) |>
    select(word1, word2, freq) |>
    mutate(corpus = "BabyLM (150M tok, 20 ep)")

  pile_raw <- read_csv("../Results/corpus_binomials_infinigram_piletrain.csv",
                       show_col_types = FALSE) |>
    rename(freq = freq_total) |>
    filter(freq > 0) |>
    select(word1, word2, freq) |>
    mutate(corpus = "Pile (~300B tok, 1 ep)")

  bind_rows(babylm_raw, pile_raw) |>
    group_by(corpus) |>
    mutate(prob = freq / sum(freq),
           rank = rank(-freq, ties.method = "first")) |>
    ungroup()
})

corpus_colors <- c("BabyLM (150M tok, 20 ep)" = "#154733",
                   "Pile (~300B tok, 1 ep)"    = "#F0B323")

p_hist <- corpus_freq_df |>
  ggplot(aes(x = log10(freq), fill = corpus, color = corpus)) +
  geom_histogram(aes(y = after_stat(count)),
                 binwidth = 0.25, position = "identity", alpha = 0.55) +
  scale_fill_manual(values  = corpus_colors, name = NULL) +
  scale_color_manual(values = corpus_colors, name = NULL) +
  scale_y_continuous(labels = label_comma()) +
  labs(x = expression(log[10]~"(count)"), y = "Number of pairs") +
  theme_paper() +
  theme(legend.position = "none")

p_rank <- corpus_freq_df |>
  ggplot(aes(x = rank, y = prob, color = corpus)) +
  geom_line(linewidth = 0.7, alpha = 0.85) +
  scale_x_log10(labels = label_comma()) +
  scale_y_log10(labels = label_scientific()) +
  scale_color_manual(values = corpus_colors, name = NULL) +
  labs(x = "Rank (log scale)", y = "Probability (log scale)") +
  theme_paper() +
  theme(legend.position = "bottom")

p_hist + p_rank + plot_layout(ncol = 2, widths = c(1, 1))
```

The BabyLM training corpus is a 150M-token collection of child-directed and informal text (CHILDES, BNC Spoken, children's books, OpenSubtitles, Wikipedia), trained for 20 epochs. The Pile is a \~300B-token web-scraped corpus used to train the Pythia family. For each of the \~47K corpus binomials, we extracted 3-gram counts (i.e., the frequency of "word$_1$ and word$_2$" and its reverse) from each corpus via infini-gram [@liu2024infinigram]. As @fig-corpus-freq shows, the BabyLM corpus produces a highly degenerate frequency distribution: 70.7% of binomials appear exactly once, 87.4% appear at most twice, and 98.2% appear at most ten times, with a maximum of 1,095 occurrences. The Pile distribution is far richer: the median binomial appears 365 times, the 25th percentile is 62 occurrences, and the 99th percentile exceeds 77,000.

In absolute training exposure terms (accounting for BabyLM's 20 epochs), the most frequent BabyLM binomials are seen at most \~21,900 times total across all training steps, while the equivalent Pile binomials are seen over 1.5 million times in a single pass. Even the median Pile binomial (365 occurrences per pass) substantially exceeds BabyLM's total exposure for a typical singleton binomial (1 × 20 = 20 exposures). This gap in absolute exposure is why the two corpora cannot be placed on a common developmental axis, and why the frequency range available for the moderation analysis differs so sharply between them.

# Ordering-Source Regression {#sec-appendix-relfreq}

Full results for @eq-relfreq: the abstract component $\hat{y}$ (`y_pred`), the corpus's own attested order for that binomial (`rel_freq`), total binomial frequency (`log_freq`), the two moderations, and the residual-scale term. All four variables are z-scored within each fit, so coefficients are standardized betas: a one-standard-deviation change in the predictor moves the model's ordering preference by that many standard deviations. Standardizing the outcome as well is what makes the coefficients comparable across models, attention conditions, and training checkpoints, since the spread of `y_true` differs between fits.

`r if (nrow(relfreq)) format(min(relfreq$n), big.mark=",") else "--"` binomials enter the smallest fit and `r if (nrow(relfreq)) format(max(relfreq$n), big.mark=",") else "--"` the largest.

Two limits on how far the moderation terms should be read. `rel_freq` is estimated from counts, so it is less reliable for rare binomials, and a less reliable predictor carries an attenuated coefficient: part of every positive `rel_freq`$\times$`log_freq` estimate is that attenuation rather than storage. Admitting singly-attested binomials trades one form of this problem for another. Nothing is discarded, but a binomial seen once now sits at the boundary $\pm 0.5$, indistinguishable from one seen a thousand times in a single order. The effect is strongest for BabyLM, where 92% of items take a boundary value and the first seven frequency deciles are entirely singletons, and mild for the large-scale models, where roughly 8% sit at a boundary and `rel_freq` varies continuously through its range. The two nevertheless return moderation estimates of the same magnitude, which is the reassuring outcome: if the coefficient were mostly a record of how often the predictor saturates, the corpus at 92% saturation and the corpora at 8% would not agree. The comparison between attention conditions is unaffected in either case, since `rel_freq` is the identical predictor carrying identical noise in both conditions, so whatever attenuation contributes cancels from the difference. Where a single number is wanted, the condition contrast is the defensible one.

```{r}
#| label: tbl-relfreq
#| tbl-cap: "Ordering-source regression (@eq-relfreq) for all 16 models at their final checkpoint, by attention condition. All four variables are z-scored within each fit, so these are standardized betas, comparable across models and conditions. 95\\% credible intervals in brackets. Holistic storage predicts a negative `y_pred` $\\times$ `log_freq` and a positive `rel_freq` $\\times$ `log_freq`, both attenuating under attention-zeroing."
#| eval: !expr nrow(relfreq) > 0

# Column names are set via col.names rather than as backtick-quoted variable
# names: R processes backslash escapes inside backticks (so "\times" would parse
# as a tab), and with escape = FALSE an unescaped underscore reaches LaTeX as a
# subscript. Both go through cleanly when written as plain LaTeX here.
tt <- function(x) paste0("\\texttt{", gsub("_", "\\\\_", x), "}")

relfreq |>
  mutate(Condition = recode_condition(condition),
         a = sprintf("%.3f [%.3f, %.3f]", pred_mean, pred_lo, pred_hi),
         b = sprintf("%.3f [%.3f, %.3f]", rel_mean,  rel_lo,  rel_hi),
         c = sprintf("%.3f [%.3f, %.3f]", pred_x_mean, pred_x_lo, pred_x_hi),
         d = sprintf("%.3f [%.3f, %.3f]", rel_x_mean,  rel_x_lo,  rel_x_hi)) |>
  arrange(model, Condition) |>
  select(Model = model, Condition, a, b, c, d) |>
  kbl(longtable = TRUE, booktabs = TRUE, linesep = "", escape = FALSE,
      col.names = c("Model", "Condition", tt("y_pred"), tt("rel_freq"),
                    paste(tt("y_pred"), "$\\times$", tt("log_freq")),
                    paste(tt("rel_freq"), "$\\times$", tt("log_freq")))) |>
  kable_styling(font_size = 7, latex_options = c("scale_down", "repeat_header")) |>
  collapse_rows(columns = 1, latex_hline = "major") |>
  allow_group_breaks()
```

@fig-relfreq-coefs shows the two moderation terms across models and conditions. The pattern is uniform across all sixteen: `rel_freq`$\times$`log_freq` is credibly positive in every model under full context and much smaller under attention-zeroing, while `y_pred`$\times$`log_freq` is credibly negative in every large-scale model under both conditions. The BabyLM models match the large-scale ones on the first term and sit nearer zero on the second, which is what the restricted frequency range would predict: `y_pred`$\times$`log_freq` is the term that needs exposure to vary over orders of magnitude before it can be estimated apart from the main effect.

```{r}
#| label: fig-relfreq-coefs
#| fig-cap: "Moderation terms of @eq-relfreq by model and attention condition. Circles = full context; triangles = attention-zeroed. Points = posterior mean; bars = 95\\% credible intervals. Left: the corpus's attested order matters more for frequent binomials, and attention-zeroing removes most of that. Right: the abstract component matters less for frequent binomials, and attention-zeroing does not remove it."
#| fig-width: 6.5
#| fig-height: 4.2
#| fig-env: "figure*"
#| eval: !expr nrow(relfreq) > 0
#| fig-pos: "t"

relfreq |>
  transmute(model_fct = factor(label, levels = rev(dumbbell_order)),
            family = factor(family, levels = c("BabyLM","GPT-2","Pythia","OLMo","Llama")),
            condition = recode_condition(condition),
            `rel_freq × log_freq` = rel_x_mean, rel_lo = rel_x_lo, rel_hi = rel_x_hi,
            `y_pred × log_freq`   = pred_x_mean, pred_lo = pred_x_lo, pred_hi = pred_x_hi) |>
  tidyr::pivot_longer(c(`rel_freq × log_freq`, `y_pred × log_freq`),
                      names_to = "term", values_to = "est") |>
  mutate(lo = ifelse(term == "rel_freq × log_freq", rel_lo, pred_lo),
         hi = ifelse(term == "rel_freq × log_freq", rel_hi, pred_hi),
         term = factor(term, levels = c("rel_freq × log_freq", "y_pred × log_freq"))) |>
  ggplot(aes(x = est, y = model_fct, color = family, shape = condition)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.4) +
  geom_linerange(aes(xmin = lo, xmax = hi), linewidth = 0.5,
                 position = position_dodge(width = 0.6)) +
  geom_point(size = 1.6, position = position_dodge(width = 0.6)) +
  facet_wrap(~term, nrow = 1, scales = "free_x") +
  scale_color_manual(values = family_colors, guide = "none") +
  scale_shape_manual(values = c("Full context" = 19, "Attention-zeroed" = 17), name = NULL) +
  labs(x = "Standardized coefficient", y = NULL) +
  theme_paper(base_size = 8) +
  theme(legend.position = "bottom", panel.spacing = unit(0.6, "lines"))
```
