---
title: "Stored or Computed? Abstract and Memorized Binomial Preferences in LLM Representations"
acl-mode: review
format:
  acl-pdf:
    keep-tex: true
    cite-method: natbib
  html: default
editor: visual
author:
  - name:
      literal: Zachary Nicholas Houghton
    email: znh@uoregon.edu
    affiliations:
      - name: University of Oregon
      - name: Vail Systems, Inc.

abstract: |
  TODO

bibliography: references.bib
tbl-cap-location: bottom
header-includes:
  - \usepackage{float}
  - \floatplacement{table}{H}
  - \usepackage[section]{placeins}
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)
library(kableExtra)
library(scales)

RES       <- "../Results"
MODEL_DIR <- "../Data/brms_models"

# ── Slug → label map ──────────────────────────────────────────────────────────
all_slugs <- c(
  "znhoughton_opt-babylm-125m-20eps-seed964" = "BabyLM-125M",
  "znhoughton_opt-babylm-350m-20eps-seed964" = "BabyLM-350M",
  "znhoughton_opt-babylm-1_3b-20eps-seed964" = "BabyLM-1.3B",
  "EleutherAI_pythia-160m"                   = "Pythia-160M",
  "EleutherAI_pythia-410m"                   = "Pythia-410M",
  "EleutherAI_pythia-1b"                     = "Pythia-1B",
  "gpt2"                                     = "GPT-2",
  "gpt2-medium"                              = "GPT-2-M",
  "gpt2-large"                               = "GPT-2-L",
  "gpt2-xl"                                  = "GPT-2-XL",
  "allenai_OLMo-1B-hf"                       = "OLMo-1B",
  "allenai_OLMo-7B-hf"                       = "OLMo-7B",
  "allenai_OLMo-2-0425-1B"                   = "OLMo-2-1B",
  "allenai_OLMo-2-1124-7B"                   = "OLMo-2-7B",
  "meta-llama_Llama-3.2-1B"                  = "Llama-1.3B",
  "meta-llama_Meta-Llama-3-8B"               = "Llama-3-8B"
)

family_map <- c(
  "BabyLM-125M" = "BabyLM", "BabyLM-350M" = "BabyLM", "BabyLM-1.3B" = "BabyLM",
  "Pythia-160M" = "Pythia", "Pythia-410M"  = "Pythia", "Pythia-1B"   = "Pythia",
  "GPT-2"       = "GPT-2",  "GPT-2-M"      = "GPT-2",  "GPT-2-L"     = "GPT-2",
  "GPT-2-XL"   = "GPT-2",
  "OLMo-1B"    = "OLMo",   "OLMo-7B"      = "OLMo",   "OLMo-2-1B"   = "OLMo",
  "OLMo-2-7B"  = "OLMo",
  "Llama-1.3B" = "Llama",  "Llama-3-8B"   = "Llama"
)

# ── Training-scale metadata ───────────────────────────────────────────────────
model_meta <- tribble(
  ~label,         ~params_M,  ~training_tokens_B,
  "BabyLM-125M",  125,         0.15,
  "BabyLM-350M",  350,         0.15,
  "BabyLM-1.3B",  1300,        0.15,
  "GPT-2",        117,         8.0,
  "GPT-2-M",      345,         8.0,
  "GPT-2-L",      762,         8.0,
  "GPT-2-XL",     1500,        8.0,
  "Pythia-160M",  160,         300.0,
  "Pythia-410M",  410,         300.0,
  "Pythia-1B",    1000,        300.0,
  "OLMo-1B",      1000,        3000.0,
  "OLMo-7B",      7000,        2500.0,
  "OLMo-2-1B",    1000,        4000.0,
  "OLMo-2-7B",    7000,        4000.0,
  "Llama-1.3B",   1200,        9000.0,
  "Llama-3-8B",   8000,        15000.0
)

# ── Load MLP R² from per-model CSVs ──────────────────────────────────────────
load_mlp <- function(slug, label, is_control = FALSE) {
  fname <- if (is_control) "by_layer_mlp_control.csv" else "by_layer_mlp.csv"
  path  <- file.path(RES, slug, fname)
  if (!file.exists(path)) return(NULL)
  read_csv(path, show_col_types = FALSE) |>
    mutate(model = label, family = family_map[label], is_control = is_control)
}

all_r2 <- map_dfr(names(all_slugs), function(slug) {
  bind_rows(
    load_mlp(slug, all_slugs[[slug]], FALSE),
    load_mlp(slug, all_slugs[[slug]], TRUE)
  )
}) |>
  filter(split %in% c("pair_novel", "word_novel")) |>
  group_by(model, family, condition, mode, split, is_control) |>
  filter(layer == max(layer)) |>
  ungroup()

# ── Precomputed brms summaries (from Scripts/prepare_results.R) ───────────────
brms_summary <- readRDS(file.path(MODEL_DIR, "..", "brms_summary.rds"))
brms_curves  <- readRDS(file.path(MODEL_DIR, "..", "brms_curves.rds"))

# ── Checkpoint slug registries ────────────────────────────────────────────────
babylm_ckpt_registry <- bind_rows(
  tibble(slug  = paste0("znhoughton_opt-babylm-125m-20eps-seed964_step",
                        c(24, 48, 144, 384, 912, 2280)),
         label = "BabyLM-125M",
         step  = c(24, 48, 144, 384, 912, 2280),
         tokens_M = c(24, 48, 144, 384, 912, 2280) * 0.8192),  # 1024 * 400 * 2 / 1e6
  tibble(slug  = paste0("znhoughton_opt-babylm-350m-20eps-seed964_step",
                        c(48, 96, 288, 768, 1824, 4560)),
         label = "BabyLM-350M",
         step  = c(48, 96, 288, 768, 1824, 4560),
         tokens_M = c(48, 96, 288, 768, 1824, 4560) * 0.4096),  # 1024 * 200 * 2 / 1e6
  tibble(slug  = paste0("znhoughton_opt-babylm-1.3b-20eps-seed964_step",
                        c(97, 194, 582, 1455, 3686, 9021)),
         label = "BabyLM-1.3B",
         step  = c(97, 194, 582, 1455, 3686, 9021),
         tokens_M = c(97, 194, 582, 1455, 3686, 9021) * 0.2048)  # 1024 * 100 * 2 / 1e6
)

pythia_ckpt_registry <- bind_rows(
  tibble(slug  = paste0("EleutherAI_pythia-160m_step", c(16, 32, 64, 256, 512, 1000)),
         label = "Pythia-160M",
         step  = c(16, 32, 64, 256, 512, 1000),
         tokens_M = c(16, 32, 64, 256, 512, 1000) * 2.097152),
  tibble(slug  = paste0("EleutherAI_pythia-410m_step", c(16, 32, 64, 256, 512, 1000)),
         label = "Pythia-410M",
         step  = c(16, 32, 64, 256, 512, 1000),
         tokens_M = c(16, 32, 64, 256, 512, 1000) * 2.097152),
  tibble(slug  = paste0("EleutherAI_pythia-1b_step",   c(16, 32, 64, 256, 512, 1000)),
         label = "Pythia-1B",
         step  = c(16, 32, 64, 256, 512, 1000),
         tokens_M = c(16, 32, 64, 256, 512, 1000) * 2.097152),
)

# ── UO-brand colour palettes ──────────────────────────────────────────────────
# Oregon Green: #154733 / #007030  UO Gold: #F0B323  Amber: #875E00  UO Blue: #005581
family_colors <- c(
  "BabyLM" = "#154733",
  "GPT-2"  = "#F0B323",
  "Pythia" = "#007030",
  "OLMo"   = "#875E00",
  "Llama"  = "#005581"
)

babylm_size_colors <- c(
  "BabyLM-125M" = "#F0B323",
  "BabyLM-350M" = "#007030",
  "BabyLM-1.3B" = "#154733"
)

# ── Factor order for model labels ──────────────────────────────────────────────
model_order <- c(
  "BabyLM-125M", "BabyLM-350M", "BabyLM-1.3B",
  "GPT-2", "GPT-2-M", "GPT-2-L", "GPT-2-XL",
  "Pythia-160M", "Pythia-410M", "Pythia-1B",
  "OLMo-1B", "OLMo-7B", "OLMo-2-1B", "OLMo-2-7B",
  "Llama-1.3B", "Llama-3-8B"
)

# ascending params within family (for dot-plot row order)
dumbbell_order <- c(
  "BabyLM-125M", "BabyLM-350M", "BabyLM-1.3B",
  "GPT-2", "GPT-2-M", "GPT-2-L", "GPT-2-XL",
  "Pythia-160M", "Pythia-410M", "Pythia-1B",
  "OLMo-1B", "OLMo-2-1B", "OLMo-7B", "OLMo-2-7B",
  "Llama-1.3B", "Llama-3-8B"
)

condition_labels <- c(
  "default"     = "Full context",
  "attn_zeroed" = "Attention-zeroed"
)

recode_condition <- function(x) {
  factor(recode(x, !!!condition_labels),
         levels = c("Full context", "Attention-zeroed"))
}

mode_labels <- c(
  "mean_pooled" = "Binomial-level",
  "words_only"  = "Word-level"
)

theme_paper <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      text             = element_text(color = "black"),
      axis.text        = element_text(color = "black", size = rel(0.9)),
      axis.title       = element_text(color = "black", size = rel(1.0)),
      legend.text      = element_text(color = "black", size = rel(0.9)),
      legend.title     = element_text(color = "black", size = rel(1.0)),
      strip.text       = element_text(color = "black", face = "bold", size = rel(1.0)),
      panel.grid.minor = element_blank(),
      axis.ticks       = element_line(color = "black"),
      axis.ticks.length = unit(3, "pt")
    )
}
```

# Introduction

# Experiment 1: BabyLM Representations {#sec-exp1}

In this first experiment, we probe three BabyLM models (125M, 350M, and 1.3B parameters)[^1] at their final training checkpoint. These three models were trained on an amount of data comparable to humans and thus are more directly comparable to the human behavioral literature. Because the binomials used in this study were selected to be absent from BabyLM's training corpus, this experiment provides the cleanest test of whether a language model encodes abstract binomial ordering preferences: any preference signal the probe recovers must have been generalized from the training distribution rather than read off from direct exposure to the test pairs.

[^1]: The BabyLM models used here are OPT-based models trained for 20 epochs on a 150M-token subset of the BabyLM benchmark corpus; @houghton2026holisticstorageverbupphrases; @warstadt-etal-2023-findings\].

<!--# Figure 2 "Word-held-out" is being cut off -->

## Methods

Below we describe the methods used in Experiment 1.

### Stimuli

We use two complementary datasets of binomial expressions (*X and Y*). Binomials are a strong test case because they are well-studied in the linguistics literature [e.g., @morgan2016; @benor2006] and human ordering preferences for them are driven by abstract preferences [@morgan2015; @morgan2016]. The **novel set** comprises approximately 340,000 binomial pairs from the Wikitext corpus [@merity2016pointer] selected to be absent from BabyLM's training corpus. The novel pairs are therefore genuinely novel for the BabyLM models used in Experiment 1.

The **attested corpus set** comprises approximately 49,000 binomial pairs from the BabyLM corpus [@warstadt-etal-2023-findings]. For each corpus pair and each model, the ordering preference is operationalized as the model's own log-probability ratio: $y_{\text{true}} = \log P(W_1\text{ and }W_2) - \log P(W_2\text{ and }W_1)$.

Binomial pairs were identified automatically using a two-stage procedure. First, candidate "$W_1$ and $W_2$" sequences were selected using spaCy [@spacy2] part-of-speech tags, retaining only cases where both $W_1$ and $W_2$ are open-class words (nouns, verbs, adjectives, or adverbs). Second, each candidate was parsed with the Berkeley Neural Parser [benepar; @kitaev-klein-2018-constituency; @kitaev-etal-2019-multilingual] to confirm that $W_1$ and $W_2$ form a minimal coordinate phrase: the two words must be co-constituents under the same phrasal node (NP, VP, ADJP, ADVP, PP, or equivalent), with that node spanning exactly the three tokens \[$W_1$, and, $W_2$\] and nothing else. This ensures that the extracted pairs are genuine coordinate binomials rather than incidental sequences where "and" links clauses or longer phrases. Because cross-category coordination is syntactically unacceptable, violating the Coordination of Likes constraint that benepar enforces through its Penn Treebank training, genuine mixed-category pairs are excluded by the constituency criterion rather than by any explicit POS filter.[^2] Novel binomials from the Wikitext corpus were extracted using the identical procedure, with the additional requirement that both words appear in BabyLM's vocabulary and the pair is absent from the attested corpus set.

[^2]: We deliberately avoid filtering by spaCy POS tag agreement because spaCy frequently misclassifies participial forms in coordinate contexts, for example tagging "mortified" or "battered" as VERB rather than ADJ in "disgusted and mortified" or "battered and broken." Pairs with apparent cross-category spaCy tags in the final corpus reflect these tagging errors: the words function as the same syntactic category in context, which is why they pass the constituency criterion in the first place. Explicit POS filtering would therefore discard many valid same-category binomials without catching any genuine cross-category ones.

### Representation Extraction

For each model, we extract hidden-state representations from the final hidden-layer. Each binomial pair is presented to the model in both orderings (*W*~1~ and *W*~2~; *W*~2~ and *W*~1~) within a natural sentence frame. We extract representations under two input modes:

-   **Binomial-level** (`mean_pooled`): the input to the probe is $[\bar{h}(W_1\text{ and }W_2); \bar{h}(W_2\text{ and }W_1)]$, where $\bar{h}$ denotes the mean of token representations across all tokens in the binomial in each ordering.
-   **Word-level** (`words_only`): the input is $[h(W_1); h(W_2)]$, where $h(W_k)$ is the hidden state of word $W_k$ extracted from the sentence in which it appears in first position, excluding the conjunction token.

### Attention Conditions

We probe representations under two attention conditions:

-   **Full context** (`default`): standard forward pass with full cross-token attention. Each word's representation reflects contextual influences from all other tokens in the sentence.
-   **Attention-zeroed** (`attn_zeroed`): attention from the second word and the conjunction to the first word is blocked by setting the corresponding attention logits to $-\infty$ before the softmax. The mask is supplied as a 4D attention mask and therefore applies at every transformer layer, not only the first, so no cross-word information reaches $W_2$ or the conjunction at any depth. This condition tests whether the second word's representation is specifically shaped by direct cross-word attention to the first word. Ablating attention edges in this way follows a line of interpretability work that uses such interventions to localize where information is transmitted within a model [@geva-etal-2023-dissecting; @wang2023interpretability; @meng2022locating]; what differs here is that we read the ablation against each pair's frequency in the model's training corpus, which turns it into a test of whether an ordering preference is retrieved or derived.

Because the word-level mode extracts each word from the sentence in which it occupies first position, its representation depends only on the preceding sentence context and is not influenced by any later token's attention. We therefore report three representation types per model: full-context binomial-level, attention-zeroed binomial-level, and full-context word-level.

Note that $y_{true}$ is the model's own log-probability ratio (Stimuli, above), so it is recomputed from whichever forward pass produced the representations. The two full-context representation types therefore share a target, whereas the attention-zeroed condition has its own: blocking cross-word attention changes the model's output probabilities as well as its hidden states. The attention-zeroed comparison is thus a comparison of two self-consistent probe problems, not a comparison of two input encodings against a fixed target.

### MLP Probe

For each representation type, we train a two-layer MLP classifier (hidden dimension 128, ReLU activation) to predict binomial preference from the extracted representations. The probe takes as its input the concatenated hidden-layer representations for a given binomial (dimensionality and construction determined by the representation type, as operationalized above) and outputs a single continuous value $\hat{y}$, trained via mean-squared-error loss to approximate the true preference score, $y_{true}$, defined in the Stimuli section. The model is trained with Adam (learning rate $10^{-3}$, weight decay $10^{-4}$) with early stopping (patience = 20) and a 10% held-out validation set. Training uses antisymmetric augmentation: for each pair, we add a flipped version (swapped ordering, negated label) with probability 0.5. We report cross-validated R² (the squared Pearson correlation between $\hat{y}$ and $y_{true}$ on held-out folds) as our measure of probe performance, averaged across 10 folds.

We evaluate generalization under two cross-validation splits:

-   **Pair-held-out** (`pair_novel`): pairs are held out as units; both orderings of a pair appear only in training or only in test.
-   **Word-held-out** (`word_novel`): all pairs containing a test word are excluded from training, testing generalization to completely novel lexical items.

### Control Probes

Following @hewitt-liang-2019-designing, we train a matched control probe on randomly permuted preference labels under the same architecture and hyperparameters. Near-zero control R² provides evidence that the probe is extracting genuine linguistic structure rather than exploiting label artifacts or probe capacity. Full control R² results are reported in the supplementary appendix for each experiment.

### Frequency-Residual Regression {#sec-methods-freq-regression}

To characterize how corpus frequency relates to probe accuracy, we regress the absolute residual error of the MLP probe on log corpus frequency using a Bayesian quadratic regression model implemented in *brms* [@burkner2017brms] with default priors. Frequency was estimated using exact counts of the binomial in the BabyLM training corpus [@warstadt-etal-2023-findings]. Our statistical model equation is presented below:

$$|y_{\text{true}} - \hat{y}| \sim \beta_0 + \beta_1 \log(\text{freq}) + \beta_2 \log(\text{freq})^2$$ {#eq-freq-regression}

```{r}
#| label: load-brms-babylm
#| include: false

babylm_labels <- c("BabyLM-125M", "BabyLM-350M", "BabyLM-1.3B")

babylm_curves <- brms_curves |>
  filter(label %in% babylm_labels, is.na(step))

babylm_coefs <- brms_summary |>
  filter(label %in% babylm_labels, is.na(step)) |>
  left_join(model_meta, by = "label") |>
  mutate(family = family_map[label])
```

```{r}
#| label: ctrl-stats
#| include: false

ctrl_r2_exp1 <- all_r2 |>
  filter(family == "BabyLM", is_control, mode == "mean_pooled") |>
  pull(mean_r2)

ctrl_r2_exp2 <- all_r2 |>
  filter(is_control, mode == "mean_pooled") |>
  pull(mean_r2)

ctrl_mean_exp1 <- mean(ctrl_r2_exp1)
ctrl_sd_exp1   <- sd(ctrl_r2_exp1)
ctrl_mean_exp2 <- mean(ctrl_r2_exp2)
ctrl_sd_exp2   <- sd(ctrl_r2_exp2)
```

## Results

### Probe Performance

Results are visualized in @fig-r2-babylm; full numerical results are in Appendix B (Table @tbl-r2-babylm). The control probe R² values were consistently near zero across all conditions and all three models (mean ± SD = `r sprintf("%.4f ± %.4f", ctrl_mean_exp1, ctrl_sd_exp1)`; full control results are in Appendix B (Table @tbl-ctrl-r2-babylm)), confirming that the probe doesn't recover spurious signal from label noise.

For the binomial-level conditions, pair-held-out R² ranges from 0.24 to 0.34; word-held-out R² is consistently lower, ranging from 0.12 to 0.23. The drop from pair-held-out to word-held-out reflects a lexical generalization cost: the probe is more accurate when the words in a test pair appeared in other pairs during probe training than when both words are entirely novel, suggesting that some preference signal is tied to word identity rather than being fully abstract. Probe accuracy is non-monotonic with model size: BabyLM-350M achieves the highest pair-held-out R² across most conditions, with BabyLM-125M and BabyLM-1.3B lower by similar margins. The full-context condition yields higher pair-held-out R² than attention-zeroed, indicating that cross-word attention contributes positively to preference encoding for these novel binomials. We return to possible explanations for this in the discussion section.

```{r}
#| label: fig-r2-babylm
#| fig-cap: "Cross-validated R² at the final hidden layer for BabyLM models, at the binomial level by attention condition. Circles = pair-held-out; triangles = word-held-out. Bars = ±1 SD (minimum width enforced for visibility)."
#| fig-width: 3.3
#| fig-height: 2.8
#| eval: !expr nrow(all_r2) > 0

all_r2 |>
  filter(family == "BabyLM", mode == "mean_pooled", !is_control) |>
  mutate(
    model      = factor(model, levels = model_order),
    size_lbl   = factor(sub("BabyLM-", "", as.character(model)),
                        levels = c("125M", "350M", "1.3B")),
    display_sd = pmax(sd_r2, 0.010),
    condition  = recode_condition(condition),
    panel      = factor(recode(condition,
                               "Full context"     = "Full context",
                               "Attention-zeroed" = "Attention-zeroed"),
                        levels = c("Full context", "Attention-zeroed"))
  ) |>
  ggplot(aes(x = size_lbl, y = mean_r2, color = model, shape = split)) +
  geom_pointrange(aes(ymin = mean_r2 - display_sd, ymax = mean_r2 + display_sd),
                  position = position_dodge(0.5), linewidth = 0.5, size = 0.35) +
  facet_wrap(~panel, nrow = 1) +
  scale_color_manual(values = babylm_size_colors, guide = "none") +
  scale_shape_manual(values = c("pair_novel" = 19, "word_novel" = 17),
                     labels = c("pair_novel" = "Pair-held-out",
                                "word_novel" = "Word-held-out"),
                     name = NULL) +
  labs(x = "Size", y = expression(R^2)) +
  theme_paper(base_size = 10) +
  theme(legend.position = "bottom",
        axis.text.x     = element_text(angle = 30, hjust = 1, size = 8))
```

### Memorization vs. Abstraction

We report the results of the frequency-residual regression results here. Full numerical results are in Appendix B (Table @tbl-beta-babylm). Under the full-context condition, the quadratic coefficient $\hat\beta_2$ is positive for all three models, with 95% credible intervals excluding zero. This indicates a U-shaped relationship between corpus frequency and probe error: the probe makes the smallest errors for pairs at intermediate log-frequency values, with errors increasing toward both the low- and high-frequency tails. Under the attention-zeroed condition, the curvature is attenuated relative to full context and remains reliably positive for BabyLM-125M and BabyLM-350M; though the effect falls below statistical significance levels for BabyLM-1.3B. The attenuation under the attention-zeroed condition indicates that the high-frequency penalty on probe accuracy is driven in part by cross-word attention: blocking the second word's attention to the first flattens the frequency-residual relationship.

## Discussion

Experiment 1 establishes that BabyLM models encode binomial ordering preferences that are both generalizable and partially lexical. Positive pair-held-out R² shows the probe generalizes beyond training binomials, and positive word-held-out R² extends this to novel words, contrary to @houghton2025role. The consistently lower word-held-out R² reveals a lexical component in the learned representations: part of the probe's accuracy reflects word-specific ordering tendencies rather than abstract principles that hold regardless of which words are used. This gap is reliably present across all three model sizes, indicating that even at small training scale, preference representations blend abstract structure with word-specific information.

The U-shaped frequency-residual relationship ($\hat{\beta}_2 > 0$ for all three models) reveals where pair-specific information is concentrated. At low frequencies, probe accuracy suffers because the model hasn't seen the pair often enough to develop reliable representations. At high frequencies, accuracy suffers for the opposite reason: repeated co-occurrence has made the representations context-sensitive to the specific pair, so they no longer reflect ordering principles that generalize across novel contexts. The attenuation of $\hat{\beta}_2$ under the attention-zeroed condition shows that cross-word attention is the mechanism driving the high-frequency penalty: blocking it partially removes the pair-specific context-sensitivity that makes high-frequency representations harder to generalize from.

Despite contributing to memorization at high frequencies, cross-word attention's overall effect in BabyLM is abstractive: full-context outperforms attention-zeroed across the frequency range. We treat cross-word attention as a context-integration mechanism whose outcome depends on training exposure. In BabyLM, where most pairs appear only once or twice, cross-word attention hasn't been exposed to pair-specific co-occurrence statistics at scale; it instead computes more general relational features that predict ordering across many pairs. Most pairs fall below the frequency threshold at which memorization dominates, so the net effect of cross-word attention is abstractive. Experiment 2 shows how this balance shifts when models are trained on far more data.

```{r}
#| label: exp2-novel-pile-stats
#| include: false

# Counts are from DOLMA, not the Pile. Dolma is OLMo's training corpus; it stands
# in here as a large-web-corpus proxy for the other families, whose exact training
# data is either different (Pile, WebText) or undisclosed (Llama).
novel_dolma_freqs <- read_csv(
  file.path(RES, "..", "Results", "novel_binomials_infinigram_dolma.csv"),
  show_col_types = FALSE
)
dolma_nonzero     <- novel_dolma_freqs$freq_total[novel_dolma_freqs$freq_total > 0]
dolma_pct_nonzero <- round(mean(novel_dolma_freqs$freq_total > 0) * 100, 1)
dolma_q025 <- round(quantile(dolma_nonzero, 0.025))
dolma_q975 <- round(quantile(dolma_nonzero, 0.975))
```

# Experiment 2: Cross-Family Final-Layer Representations {#sec-exp2}

Experiment 1 demonstrates that BabyLMs are relying on both abstract preferences and memorization for binomial ordering preferences. However, those are models that are both smaller in terms of number of parameters as well as in terms of training data. In Experiment 2, we extend this to a larger variety of models.

## Methods

In Experiment 2, we extend the probe to 13 models across four large-scale families: Pythia (160M, 410M, 1B), GPT-2 (117M–1.5B), OLMo (1B–7B), and Llama (1.3B, 8B), and include the BabyLM results from Experiment 1 for comparison. One design consideration differs: the novel binomials were selected to be absent from BabyLM's training corpus, so they are genuinely novel only for the BabyLM models of Experiment 1. For the large-scale models they are not. Indexing them against Dolma [@dolma], approximately `r dolma_pct_nonzero`% occur at non-zero frequency, most at low-to-medium exposure (counts spanning `r dolma_q025` to `r dolma_q975` occurrences for the middle 95% of items). Dolma is OLMo's training corpus and serves here as a proxy for large-scale web-crawled text generally; the other families were trained on the Pile, WebText, or undisclosed data, so this is an approximation rather than a direct exposure measure for them. The probe R² results for the large-scale models therefore reflect representations of low-to-mid-frequency attested pairs rather than truly novel ones, and any claim that a preference must have been generalized rather than retrieved applies only to the BabyLM models. Probe architecture, attention conditions, and frequency regression are identical to Experiment 1.

## Results

### Probe Performance

Full numerical results are in Appendix C (Table @tbl-r2); @fig-r2 visualizes the pair-held-out results. Control probe R² was consistently near zero across all models and conditions (mean ± SD = `r sprintf("%.4f ± %.4f", ctrl_mean_exp2, ctrl_sd_exp2)`; full control results in Appendix C (Table @tbl-ctrl-r2)). As in Experiment 1, the binomial-level conditions (full-context and attention-zeroed) are directly comparable; word-level results are provided in Appendix C (Table @tbl-r2-words-only). All Pile-trained models substantially outperform BabyLM: pair-held-out R² under the full-context, binomial-level condition ranges from 0.43 to 0.65 across the four Pile-trained families, compared to 0.29–0.34 for BabyLM. Word-held-out R² is consistently lower than pair-held-out R² for every model and condition, matching the pattern in Experiment 1. GPT-2 shows the most striking result: under the attention-zeroed condition, pair-held-out R² reaches 0.79 for GPT-2 (117M; substantially higher than its full-context value of 0.65), with attention-condition advantages across all four GPT-2 models: GPT-2-M (0.76 vs. 0.53), GPT-2-L (0.67 vs. 0.52), and GPT-2-XL (0.73 vs. 0.49). Pythia, OLMo, and Llama-1.3B show the same direction: attention-zeroed outperforms full context at the binomial level. Llama-3-8B is an exception, matching the BabyLM pattern with full context slightly outperforming attention-zeroed (0.54 vs. 0.52). Scale effects within families are non-monotonic: in GPT-2, R² decreases with model size under full context (117M highest at 0.65, XL lowest at 0.49); in Pythia, the pattern is non-monotonic (160M highest at 0.59, 410M lowest at 0.53, 1B at 0.55); OLMo's 7B models slightly outperform 1B counterparts; Llama's 8B model outperforms 1.3B.

```{r}
#| label: fig-r2
#| fig-cap: "Cross-validated R² at the final hidden layer for all 16 models (binomial-level), by attention condition (columns) and cross-validation split (shape). Circles = pair-held-out; triangles = word-held-out. Bars = 95\\% CI across 10 folds ($t$-distribution, $df = 9$). Models ordered by parameter count within each family and colored by family. Control R² $\\approx$ 0 (Appendix C)."
#| fig-width: 6.5
#| fig-height: 3.2
#| fig-env: "figure*"
#| eval: !expr nrow(all_r2) > 0

size_labels <- c(
  "BabyLM-125M" = "125M", "BabyLM-350M" = "350M", "BabyLM-1.3B" = "1.3B",
  "GPT-2" = "S", "GPT-2-M" = "M", "GPT-2-L" = "L", "GPT-2-XL" = "XL",
  "Pythia-160M" = "160M", "Pythia-410M" = "410M", "Pythia-1B" = "1B",
  "OLMo-1B" = "1B", "OLMo-7B" = "7B", "OLMo-2-1B" = "2-1B", "OLMo-2-7B" = "2-7B",
  "Llama-1.3B" = "1.3B", "Llama-3-8B" = "8B"
)

all_r2 |>
  filter(mode == "mean_pooled", !is_control) |>
  mutate(
    condition = recode_condition(condition),
    split_lbl = factor(recode(split,
                              "pair_novel" = "Pair-held-out",
                              "word_novel" = "Word-held-out"),
                       levels = c("Pair-held-out", "Word-held-out")),
    family    = factor(family, levels = c("BabyLM", "GPT-2", "Pythia", "OLMo", "Llama")),
    model_fct = factor(model, levels = dumbbell_order),
    ci_half   = qt(0.975, df = 9) * sd_r2 / sqrt(n_folds),
    r2_lo     = mean_r2 - pmax(ci_half, 0.02),
    r2_hi     = mean_r2 + pmax(ci_half, 0.02)
  ) |>
  ggplot(aes(x = model_fct, y = mean_r2, color = family, shape = split_lbl)) +
  geom_linerange(aes(ymin = r2_lo, ymax = r2_hi), linewidth = 0.7,
                 position = position_dodge(width = 0.6)) +
  geom_point(size = 2.0, position = position_dodge(width = 0.6)) +
  facet_wrap(~condition, nrow = 1) +
  scale_color_manual(values = family_colors, name = "Family",
                     guide = guide_legend(nrow = 1, order = 1)) +
  scale_shape_manual(values = c("Pair-held-out" = 19, "Word-held-out" = 17), name = NULL,
                     guide = guide_legend(nrow = 1, order = 2)) +
  scale_x_discrete(labels = size_labels) +
  scale_y_continuous(breaks = c(0.25, 0.5, 0.75)) +
  labs(y = expression(R^2), x = NULL) +
  theme_paper(base_size = 9) +
  theme(panel.spacing   = unit(0.8, "lines"),
        legend.position = "bottom",
        legend.box      = "vertical",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 7))
```

```{r}
#| label: load-brms-all
#| include: false

all_coefs <- brms_summary |>
  filter(is.na(step), label %in% names(family_map)) |>
  left_join(model_meta, by = "label") |>
  mutate(family = family_map[label])
```

### Memorization vs. Abstraction

Full numerical results are in Appendix C (Table @tbl-beta) and visualized in @fig-beta. $\hat{\beta}_2$ is positive for all 16 models under both attention conditions, with 95% credible intervals excluding zero in all cases except BabyLM-1.3B under the attention-zeroed condition ($\hat{\beta}_2 = 0.008$; 75% of posterior above zero). The U-shaped frequency-residual relationship is therefore a robust finding across model families, training corpora, and parameter scales. Two patterns stand out. First, BabyLM models show larger $\hat{\beta}_2$ magnitudes under full context (0.064–0.073) than any Pile-trained family (0.007–0.018), despite far less training, an effect likely driven by BabyLM's extreme frequency distribution (70% singleton pairs), which creates a sharper contrast between low- and high-frequency items in the regression. Second, under the attention-zeroed condition, $\hat{\beta}_2$ is consistently attenuated relative to full context across all families, confirming that cross-word attention contributes to the high-frequency memorization effect regardless of model family or scale.

<!--# need to back up the statement that $\hat{\beta}_2 is consistently attentuated under attention-zeroed condition with a statistic -->

It is worth noting that the frequency-residual regression results do not reflect an underlying correlation between preference strength and frequency: all $|r| < 0.04$ between $|y_{true}|$ and $\log(\text{frequency})$ across all 16 models. Appendix C (Figure @fig-scatter) corroborates this visually: mid-frequency pairs cluster tightly around the prediction diagonal, while low- and high-frequency pairs scatter in opposite directions, precisely the pattern implied by positive $\hat{\beta}_2$.

```{r}
#| label: fig-beta
#| fig-cap: "Quadratic curvature ($\\hat{\\beta}_2$) of the frequency-residual relationship at the final hidden layer, binomial-level by attention condition. Circles = full context; triangles = attention-zeroed. Positive values indicate that probe error is U-shaped: lowest at intermediate corpus frequencies, highest at extremes. Points = posterior mean; bars = 95\\% credible intervals (minimum bar enforced for visibility)."
#| fig-width: 3.3
#| fig-height: 4.8
#| eval: !expr nrow(all_coefs) > 0

all_coefs |>
  filter(mode == "mean_pooled") |>
  mutate(
    condition = recode_condition(condition),
    family    = factor(family, levels = c("BabyLM", "GPT-2", "Pythia", "OLMo", "Llama")),
    model_fct = factor(label, levels = dumbbell_order),
    b_lo      = pmin(b_quad_sq_lo, b_quad_sq - 0.006),
    b_hi      = pmax(b_quad_sq_hi, b_quad_sq + 0.006)
  ) |>
  ggplot(aes(y = model_fct, x = b_quad_sq, color = family, shape = condition)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.4) +
  geom_linerange(aes(xmin = b_lo, xmax = b_hi), linewidth = 0.6,
                 position = position_dodge(width = 0.6)) +
  geom_point(size = 2.0, position = position_dodge(width = 0.6)) +
  facet_grid(family ~ ., scales = "free_y", space = "free_y") +
  scale_color_manual(values = family_colors, guide = "none") +
  scale_shape_manual(values = c("Full context" = 19, "Attention-zeroed" = 17), name = NULL) +
  scale_x_continuous(breaks = c(0, 0.05, 0.10),
                     labels = c("0", "0.05", "0.10"),
                     expand = expansion(mult = c(0.05, 0.12))) +
  labs(x = expression(hat(beta)[2]~"(quadratic curvature)"), y = NULL) +
  theme_paper(base_size = 9) +
  theme(strip.text.y  = element_blank(),
        panel.spacing = unit(0.2, "lines"),
        legend.position = "bottom")
```

## Discussion

Experiment 2 shows that both abstraction and lexical memorization increase with training scale, though not uniformly. Pile-trained models achieve substantially higher word-held-out R² than BabyLM models of comparable size, indicating that larger-scale training on naturally varied text improves abstract preference representations beyond what BabyLM's data regime allows. The pair-held-out vs. word-held-out gap persists across all Pile-trained families, confirming that the lexical component identified in Experiment 1 is not eliminated by scale. Within families, word-held-out R² does not scale monotonically with parameter count, suggesting that corpus composition and diversity, rather than model size alone, drive the quality of abstract preference representations.

The U-shaped frequency-residual relationship holds across all 16 models ($\hat{\beta}_2 > 0$ in all cases), replicating the key finding from Experiment 1. BabyLM's larger $\hat{\beta}_2$ magnitudes relative to Pile-trained families reflect BabyLM's more extreme frequency distribution, in which the contrast between singleton and high-frequency pairs is sharper. The attenuation of $\hat{\beta}_2$ under the attention-zeroed condition is consistent across all five families, confirming that cross-word attention mediates the high-frequency memorization effect regardless of training scale or architecture. The same mechanism identified in Experiment 1 operates universally.

The most critical difference from Experiment 1 is in the attention condition comparison. Unlike BabyLM, where full-context outperforms attention-zeroed, all Pile-trained families show the reverse: blocking cross-word attention improves probe generalizability. Under the context-integration framing introduced in Experiment 1, this flip indicates that cross-word attention has crossed the threshold from computing general relational features to encoding pair-specific co-occurrence statistics. With sufficient and diverse training data, the same mechanism that abstracts in a data-limited regime instead becomes a vehicle for pair-specific memorization. Experiment 3 traces the developmental timecourse of this transition.

# Experiment 3: Developmental Trajectories {#sec-exp3}

Experiments 1 and 2 establish that language models encode binomial preferences at the end of training, but they leave open the timecourse in which these representations emerge, and whether the memorization-abstraction tradeoff is present from early on or crystallizes with additional exposure. Experiment 3 addresses this by examining the same probe at multiple training checkpoints. We ask two questions: (1) Does probe accuracy increase monotonically with training, or does it plateau early, suggesting that preference representations are rapidly acquired? (2) Does the U-shaped frequency-residual relationship hold from early checkpoints, or does it only emerge as training progresses and memorization of high-frequency items accumulates?

## Methods

We probe six checkpoints across two different model families (BabyLM and Pythia). The same BabyLMs from the previous Experiments were used for this one(125M, 350M, and 1.3B) and the comparable Pythia models were chosen for rough comparison (160M, 410M, and 1B). Checkpoints were sampled at approximately even intervals across the BabyLM training. For Pythia, however, rather than sampling even intervals across training, we instead opted to choose comparable checkpoints such that the Pythia models had seen roughly the same number of tokens as the BabyLMs at each checkpoint. These checkpoints span roughly 20M to 2B effective training tokens for each model.

The probe, input modes, attention conditions, and cross-validation splits are identical to Experiments 1 and 2. For the frequency-residual analysis, we fit the quadratic Bayesian regression described in Experiment 1 at each checkpoint and report $\hat\beta_2$ (the curvature of the U-shape) as a function of training step.

<!--# Maybe should go back to reporting the vertex? -->

## Results

```{r}
#| label: load-r2-checkpoints
#| cache: false
#| include: false

load_ckpt_r2 <- function(registry) {
  map_dfr(seq_len(nrow(registry)), function(i) {
    slug  <- registry$slug[i]
    label <- registry$label[i]
    step  <- registry$step[i]
    tok   <- registry$tokens_M[i]
    bind_rows(
      load_mlp(slug, label, FALSE),
      load_mlp(slug, label, TRUE)
    ) |>
      mutate(step = step, tokens_M = tok)
  }) |>
    filter(split %in% c("pair_novel", "word_novel")) |>
    group_by(model, condition, mode, split, is_control, step) |>
    filter(layer == max(layer)) |>
    ungroup()
}

babylm_ckpt_r2  <- load_ckpt_r2(babylm_ckpt_registry)
pythia_ckpt_r2  <- load_ckpt_r2(pythia_ckpt_registry)

ckpt_r2 <- bind_rows(
  babylm_ckpt_r2 |> mutate(family_lbl = "BabyLM"),
  pythia_ckpt_r2 |> mutate(family_lbl = "Pythia")
) |> mutate(family_lbl = factor(family_lbl, levels = c("BabyLM", "Pythia")))
```

```{r}
#| label: load-brms-checkpoints
#| include: false

babylm_ckpt_coefs <- brms_summary |>
  filter(label %in% c("BabyLM-125M", "BabyLM-350M", "BabyLM-1.3B"), !is.na(step)) |>
  left_join(babylm_ckpt_registry |> select(label, step, tokens_M),
            by = c("label", "step"))

pythia_ckpt_coefs <- brms_summary |>
  filter(label %in% c("Pythia-160M", "Pythia-410M", "Pythia-1B"),
         !is.na(step)) |>
  left_join(pythia_ckpt_registry |> select(label, step, tokens_M),
            by = c("label", "step"))

ckpt_coefs <- bind_rows(
  babylm_ckpt_coefs |> mutate(family_lbl = "BabyLM"),
  pythia_ckpt_coefs |> mutate(family_lbl = "Pythia")
) |> mutate(family_lbl = factor(family_lbl, levels = c("BabyLM", "Pythia")))

ckpt_colors <- c(
  "BabyLM-125M" = "#4E9D69",
  "BabyLM-350M" = "#007030",
  "BabyLM-1.3B" = "#154733",
  "Pythia-160M"  = "#F0B323",
  "Pythia-410M"  = "#B8860B",
  "Pythia-1B"    = "#875E00"
)
```

BabyLM pair-held-out R² is already high at the earliest checkpoint (\~20M tokens: 0.67–0.72) and broadly declines to 0.32–0.38 by the end of training. At the earliest checkpoint, the attention-zeroed condition outperforms full context (0.78–0.80 vs. 0.67–0.72); this ordering reverses between \~120M and \~750M tokens, so that full context exceeds attention-zeroed at the final checkpoint for all three model sizes. Pythia shows an inverted-U trajectory instead: R² rises from 0.68–0.74 at 33.6M tokens to a peak of 0.80–0.86 at 537M tokens, then declines to 0.71–0.76 at 1074M–2097M tokens, consistent with Pile-trained models still consolidating preference representations during this early window (the first 0.7% of 300B tokens). The attention condition flip for Pythia follows the same qualitative pattern as BabyLM. Full numerical results are in Appendix D; word-level results are in Appendix D.

For $\hat{\beta}_2$, BabyLM starts negative at all sizes (-0.013 to -0.029), remains negative until approximately 750M tokens, and reaches 0.045–0.093 at the final checkpoint, matching the positive curvature from Experiment 1. Pythia's $\hat{\beta}_2$ is positive from its earliest checkpoint (33.6M tokens), with credible intervals confidently excluding zero by 537M tokens. As discussed in Appendix E, this contrast reflects the far greater absolute token exposure per pair in the Pile relative to BabyLM. Full numerical results are in Appendix D.

```{r}
#| label: fig-checkpoints
#| fig-cap: "Probe R² (top) and quadratic frequency-residual curvature ($\\hat{\\beta}_2$, bottom) across BabyLM and Pythia training checkpoints, binomial-level by attention condition. BabyLM spans ~20M–1.9B effective training tokens; Pythia spans ~34M–2.1B tokens (~0.7\\% of the full 300B training run). Ribbons show 95\\% credible intervals. $\\hat{\\beta}_2$ y-axes are free per family row (BabyLM and Pythia have very different scales). Control R² $\\approx$ 0 (Appendix D)."
#| fig-width: 3.3
#| fig-height: 4.5
#| eval: !expr nrow(ckpt_r2) > 0

p_r2 <- ckpt_r2 |>
  filter(mode == "mean_pooled", !is_control) |>
  mutate(
    model     = factor(model, levels = model_order),
    condition = recode_condition(condition),
    panel     = factor(recode(condition,
                              "Full context"     = "Full context",
                              "Attention-zeroed" = "Attention-zeroed"),
                       levels = c("Full context", "Attention-zeroed")),
    split_lbl = factor(recode(split,
                              "pair_novel" = "Pair-held-out",
                              "word_novel" = "Word-held-out"),
                       levels = c("Pair-held-out", "Word-held-out"))
  ) |>
  ggplot(aes(x = tokens_M, y = mean_r2, color = model,
             linetype = split_lbl, group = interaction(model, split_lbl))) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 0.7) +
  facet_grid(family_lbl ~ panel, scales = "free_x") +
  scale_x_log10(breaks = c(100, 1000),
               labels = label_number(suffix = "M", accuracy = 1, big.mark = ""),
               expand = expansion(mult = c(0.05, 0.1))) +
  scale_color_manual(values = ckpt_colors, name = "Model") +
  scale_linetype_manual(values = c("Pair-held-out" = "solid", "Word-held-out" = "dashed"),
                        guide = "none") +
  labs(x = NULL, y = expression(R^2)) +
  theme_paper(base_size = 9) +
  theme(legend.position = "none")

p_beta <- ckpt_coefs |>
  filter(mode == "mean_pooled") |>
  mutate(
    model     = factor(label, levels = model_order),
    condition = recode_condition(condition),
    panel     = factor(recode(condition,
                              "Full context"     = "Full context",
                              "Attention-zeroed" = "Attention-zeroed"),
                       levels = c("Full context", "Attention-zeroed"))
  ) |>
  ggplot(aes(x = tokens_M, y = beta2_mean, color = model, fill = model, group = model)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.3) +
  geom_ribbon(aes(ymin = beta2_ci025, ymax = beta2_ci975), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 0.7) +
  facet_grid(family_lbl ~ panel, scales = "free") +
  scale_x_log10(breaks = c(100, 1000),
               labels = label_number(suffix = "M", accuracy = 1, big.mark = ""),
               expand = expansion(mult = c(0.05, 0.1))) +
  scale_y_continuous(n.breaks = 3) +
  scale_color_manual(values = ckpt_colors, name = "Model") +
  scale_fill_manual(values  = ckpt_colors, guide = "none") +
  labs(x = "Tokens (M, log scale)", y = expression(hat(beta)[2])) +
  theme_paper(base_size = 9) +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 3, title = "Model"))

p_r2 / p_beta + plot_layout(heights = c(1, 1.4))
```

## Discussion

The checkpoint trajectories connect the endpoints observed in Experiments 1 and 2 to a developmental timecourse. BabyLM pair-held-out R² starts high at the earliest checkpoints and declines over training, while the pair-held-out vs. word-held-out gap narrows, indicating that the lexical component of the representation weakens faster than the generalizable component. The final BabyLM checkpoint examined in Experiment 1 represents the endpoint of this trajectory. Pythia's pair-held-out R² follows an inverted-U, consistent with Pile-trained models still consolidating preference representations in the first 0.7% of training examined here. The Pile-trained final-layer results from Experiment 2 represent where these trajectories eventually lead.

The $\hat{\beta}_2$ trajectories show that the onset of pair-specific memorization follows a different timecourse across the two families. BabyLM's $\hat{\beta}_2$ starts negative, reflecting conventional frequency effects in which more frequent pairs are easier to decode, and only turns positive after approximately 750M effective training tokens, when pair-specific exposure has accumulated enough to introduce the high-frequency penalty. Pythia's $\hat{\beta}_2$ is positive at every checkpoint, including the earliest (33.6M tokens), although the credible intervals include zero until roughly 1B tokens and the estimates only become reliably positive thereafter. The direction is nonetheless consistent from the outset, in line with the Pile's far greater absolute token exposure per pair: even at the earliest stage of Pile training, many binomials have been seen often enough for pair-specific co-occurrence statistics to begin shaping the representation. The difference is not in the mechanism but in when the threshold for pair-specific memorization is crossed.

The attention condition comparison provides the most direct window into the developmental arc of cross-word attention’s role. Both BabyLM and Pythia begin in a regime where attention-zeroed outperforms full-context. This pattern admits two interpretations: cross-word attention may be adding noise from random initialization, or it may already be encoding nascent pair-specific co-occurrence statistics that hurt generalization to novel words — the same interpretation we adopt for fully-trained Pile models in Experiment 2. Our behavioral R² data cannot distinguish these possibilities. Crucially, however, the pair-word R² gap is present from the earliest checkpoints alongside positive word-held-out R², establishing that lexical memorization and abstraction co-occur from the onset of training rather than one preceding the other. As training continues, full-context overtakes attention-zeroed in both families, indicating that cross-word attention shifts toward computing more generalizable relational features. The checkpoints here span only the first 0.7% of Pythia’s full 300B-token training, so the further shift back toward pair-specific memorization (observed in Experiment 2’s fully-trained Pile models) falls outside this window. Together, Experiments 1, 2, and 3 trace a coherent arc: memorization and abstraction are not staged processes but co-occur throughout training, with cross-word attention oscillating between pair-specific and generalizable encoding as a function of cumulative exposure per pair.

# Conclusion {#sec-discussion}

Experiment 1 showed that BabyLM models encode binomial ordering preferences recoverable from their final-layer representations, generalizing both to held-out pairs and to held-out words, and that probe error is U-shaped in corpus frequency. Experiment 2 extended this to sixteen models across five families and 117M to 8B parameters. Probe R² exceeded near-zero control values in every case and was higher for the Pile-trained models (0.43–0.79) than for BabyLM (0.24–0.34), with non-monotonic scale effects within families. The U-shaped frequency-error relationship held in every model and both attention conditions, and the conditions differed in overall R², with attention-zeroed exceeding full context in the Pile-trained families and the reverse in BabyLM. Experiment 3 traced the timecourse across BabyLM and Pythia checkpoints: the curvature is negative or near zero at the earliest checkpoints and becomes reliably positive later, while probe R² rises to a peak within the first few hundred million effective training tokens and declines thereafter in both families.

The frequency dependence of that recovery is the central result. A probe trained only on pairs the model never saw, which estimates what generalizable ordering knowledge alone predicts, is least accurate at the extremes of the frequency distribution, a pattern ($\hat{\beta}_2 > 0$) that holds in every model and both attention conditions. The failure at the high-frequency end is the informative one: representations of frequently encountered binomials carry ordering information that a word-general mapping does not reproduce. Blocking attention between the two conjuncts does not merely weaken this effect but reverses it over most of the frequency range, so that the most frequent binomials become the best predicted. The individual words therefore carry a regular ordering signal, and carry it most strongly for precisely the pairs the model has encountered most often. It is within the phrase that this regularity is altered: what makes a frequent binomial hard to derive from general structure is not what its words encode, but what cross-word attention does to it. Across BabyLM training the same relationship turns over, with frequent pairs beginning as the easiest to predict from general structure and ending as the hardest.

The broader point is that repeated exposure changes how a phrase is represented, from something the model derives to something it retrieves. Usage-based accounts have long held that frequently encountered expressions become entrenched, retrieved as units rather than assembled from their parts, and behavioral work on binomials has shown that human ordering preferences reflect both abstract constraints and direct experience with particular pairs. Those accounts rest on behavioral evidence, which can establish that frequent expressions behave as units but not what makes them do so. Our results identify a mechanism: cross-word attention, acting selectively on the most frequent pairs and strengthening over training. Methodologically, the ablation is what makes this readable. Interventions on attention have been used to trace factual recall and to isolate circuits [@geva-etal-2023-dissecting; @wang2023interpretability; @meng2022locating], but pairing such an intervention with a frequency-stratified probe turns it into a test of whether a given piece of knowledge is retrieved or derived, a question that arises for any construction with both regular and idiosyncratic instances. The result also complicates an assumption common in interpretability work, that a model's grasp of a general pattern and its knowledge of particular items simply add together. They do not: the phrases a model knows best are the ones its general knowledge explains worst.

```{=latex}
\bibliography{references}
```

\clearpage
\onecolumn
```{=latex}
\appendix
```

\clearpage

# Model Details {#sec-appendix-models}

Table \ref{tbl-models} lists all 16 models used across experiments, organized by family, with approximate parameter counts and training token counts.

```{r}
#| label: tbl-models
#| tbl-cap: "Models used across experiments. Training token counts for full-scale models are approximate; GPT-2 counts are estimated from the reported 40GB WebText corpus. BabyLM models train for 20 epochs over 150M tokens (3B effective token exposures)."

model_table <- tribble(
  ~Family,  ~Model,          ~Parameters, ~`Tokens`,
  "BabyLM", "BabyLM-125M",  "125M",      "150M",
  "BabyLM", "BabyLM-350M",  "350M",      "150M",
  "BabyLM", "BabyLM-1.3B",  "1.3B",      "150M",
  "GPT-2",  "GPT-2",        "117M",      "~8B",
  "GPT-2",  "GPT-2-medium", "345M",      "~8B",
  "GPT-2",  "GPT-2-large",  "762M",      "~8B",
  "GPT-2",  "GPT-2-XL",     "1.5B",      "~8B",
  "Pythia", "Pythia-160M",  "160M",      "~300B",
  "Pythia", "Pythia-410M",  "410M",      "~300B",
  "Pythia", "Pythia-1B",    "1B",        "~300B",
  "OLMo",   "OLMo-1B",      "1B",        "3T",
  "OLMo",   "OLMo-7B",      "7B",        "2.5T",
  "OLMo",   "OLMo-2-1B",    "1B",        "~4T",
  "OLMo",   "OLMo-2-7B",    "7B",        "~4T",
  "Llama",  "Llama-1.3B",   "1.2B",      "~9T",
  "Llama",  "Llama-3-8B",   "8B",        "~15T"
)

model_table |>
  kbl(booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 9, latex_options = "HOLD_position") |>
  collapse_rows(columns = 1, latex_hline = "major")
```

\clearpage

# Experiment 1: Supplementary Results {#sec-appendix-exp1}

This appendix provides supplementary materials for Experiment 1 (§\ref{sec-exp1}), which examined final-layer probe performance and frequency-residual curvature for the three BabyLM models. Subsections report full R² tables, control probe results, a word-level replication, and the frequency-residual regression coefficients.

## Full Probe R² {#sec-appendix-full-r2-babylm}

Probe R² at the final hidden layer for BabyLM models, at the binomial level by attention condition. P = pair-held-out; W = word-held-out. SD across 10 folds in parentheses.

```{r}
#| label: tbl-r2-babylm
#| tbl-cap: "Probe R² at the final hidden layer for BabyLM models, at the binomial level by attention condition. P = pair-held-out split; W = word-held-out split. SD across 10 folds in parentheses."
#| eval: !expr nrow(all_r2) > 0

all_r2 |>
  filter(family == "BabyLM", mode == "mean_pooled") |>
  select(model, condition, split, mean_r2, sd_r2, is_control) |>
  pivot_wider(names_from  = is_control,
              values_from = c(mean_r2, sd_r2),
              names_sep   = "_") |>
  rename(r2_main = mean_r2_FALSE, sd_main = sd_r2_FALSE,
         r2_ctrl = mean_r2_TRUE,  sd_ctrl = sd_r2_TRUE) |>
  mutate(
    condition = recode_condition(condition),
    model     = factor(model, levels = model_order),
    Split     = if_else(split == "pair_novel", "P", "W"),
    `R²`      = sprintf("%.3f (%.3f)", r2_main, sd_main),
    Ctrl      = sprintf("%.3f (%.3f)", r2_ctrl,  sd_ctrl),
  ) |>
  arrange(model, condition, split) |>
  select(Model = model, Condition = condition, Split, `R²`, Ctrl) |>
  kbl(booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 8, latex_options = "HOLD_position") |>
  collapse_rows(columns = 1:2, latex_hline = "major")
```

## Control Probe R² {#sec-appendix-ctrl-exp1}

Control probe R² for BabyLM models at the final hidden layer, at the binomial level by attention condition. Labels are randomly permuted; values near zero confirm the main probe recovers genuine linguistic structure rather than label artifacts or probe capacity.

```{r}
#| label: tbl-ctrl-r2-babylm
#| tbl-cap: "Control probe R² for BabyLM models (final layer), at the binomial level by attention condition. SD across 10 folds in parentheses. P = pair-held-out; W = word-held-out."

all_r2 |>
  filter(family == "BabyLM", mode == "mean_pooled", is_control) |>
  mutate(
    condition    = recode(condition, !!!condition_labels),
    model        = factor(model, levels = model_order),
    Split        = if_else(split == "pair_novel", "P", "W"),
    `Control R²` = sprintf("%.4f (%.4f)", mean_r2, sd_r2)
  ) |>
  arrange(model, condition, split) |>
  select(Model = model, Condition = condition, Split, `Control R²`) |>
  kbl(booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 8, latex_options = "HOLD_position") |>
  collapse_rows(columns = 1:2, latex_hline = "major")
```

## Word-Level Replication {#sec-appendix-words-only-exp1}

We trained the MLP probe using word-level representations (`words_only`) for the three BabyLM models to verify that the preference signal is not an artifact of mean pooling across the full binomial phrase. Under this condition the probe receives a concatenation of two word vectors extracted from isolated sentence contexts ("*the bread*", "*the butter*"), so it cannot rely on any co-occurrence signal from the two words' shared sentence context.

Pair-held-out R² ranges from 0.31 to 0.33, within a few percentage points of the corresponding binomial-level values (0.29–0.34). The generalization pattern is preserved: word-held-out R² is consistently lower than pair-held-out R² for all three models (word-held-out 0.18–0.20; pair-minus-word gap 0.127–0.129), confirming that the lexical generalization cost seen at the binomial level is a property of the preference encoding itself rather than an artifact of the mean-pooled representation. $\hat{\beta}_2$ is positive and credible intervals exclude zero for all three models, matching the curvature found at the binomial level (words-only $\hat{\beta}_2$: 0.039–0.079; full-context binomial-level: 0.064–0.073). Full results are in @tbl-r2-babylm-words-only.

```{r}
#| label: tbl-r2-babylm-words-only
#| tbl-cap: "Word-level probe R² and selectivity (main $-$ control R²) for BabyLM models (final layer). P = pair-held-out split; W = word-held-out split. SD across 10 folds in parentheses."

all_r2 |>
  filter(family == "BabyLM", mode == "words_only", condition == "default") |>
  select(model, condition, mode, split, mean_r2, sd_r2, is_control) |>
  pivot_wider(names_from  = is_control,
              values_from = c(mean_r2, sd_r2),
              names_sep   = "_") |>
  rename(r2_main = mean_r2_FALSE, sd_main = sd_r2_FALSE,
         r2_ctrl = mean_r2_TRUE,  sd_ctrl = sd_r2_TRUE) |>
  mutate(
    selectivity = r2_main - r2_ctrl,
    mode        = recode(mode, !!!mode_labels),
    model       = factor(model, levels = model_order),
    Split       = if_else(split == "pair_novel", "P", "W"),
    `R²`        = sprintf("%.3f (%.3f)", r2_main, sd_main),
    Ctrl        = sprintf("%.3f (%.3f)", r2_ctrl,  sd_ctrl),
    Sel.        = sprintf("%.3f", selectivity)
  ) |>
  arrange(model, split) |>
  select(Model = model, Mode = mode, Split, `R²`, Ctrl, Sel.) |>
  kbl(booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 8, latex_options = "HOLD_position") |>
  collapse_rows(columns = 1:2, latex_hline = "major")
```

## Frequency-Residual Regression: BabyLM Models {#sec-appendix-beta-babylm}

Bayesian frequency-residual regression results for BabyLM models (final layer), at the binomial level by attention condition. 95% CIs in brackets. Vertex = $-\hat\beta_1 / (2\hat\beta_2)$, the log-frequency at which probe error is minimized.

```{r}
#| label: tbl-beta-babylm
#| tbl-cap: "Bayesian frequency-residual regression results for BabyLM models (final layer), at the binomial level by attention condition. 95\\% CIs in brackets. Vertex = $-\\hat\\beta_1 / (2\\hat\\beta_2)$, the log-frequency at which probe error is minimized."
#| eval: !expr nrow(babylm_coefs) > 0

babylm_coefs |>
  filter(mode == "mean_pooled") |>
  mutate(
    model     = factor(label, levels = model_order),
    condition = recode_condition(condition),
    `$\\hat{\\beta}_1$` = sprintf("%.4f [%.4f, %.4f]",
                        b_quad_linear, b_quad_linear_lo, b_quad_linear_hi),
    `$\\hat{\\beta}_2$` = sprintf("%.5f [%.5f, %.5f]",
                        b_quad_sq, b_quad_sq_lo, b_quad_sq_hi),
    Vertex    = sprintf("%.2f", vertex)
  ) |>
  arrange(model, condition) |>
  select(Model = model, Condition = condition,
         `$\\hat{\\beta}_1$`, `$\\hat{\\beta}_2$`, Vertex) |>
  kbl(booktabs = TRUE, linesep = "", escape = FALSE) |>
  kable_styling(font_size = 9, latex_options = "HOLD_position") |>
  collapse_rows(columns = 1, latex_hline = "major")
```

\clearpage

# Experiment 2: Supplementary Results {#sec-appendix-exp2}

This appendix provides supplementary materials for Experiment 2 (§\ref{sec-exp2}), which examined final-layer representations across all 16 models from five families. Subsections report full R² tables, control probe results, a word-level replication, frequency-residual regression coefficients for all models, and a predicted-vs.-observed scatter plot.

## Full Probe R² {#sec-appendix-full-r2}

Probe R² for all 16 models at the final hidden layer, binomial-level input mode, by attention condition and cross-validation split. P = pair-held-out; W = word-held-out. SD across 10 folds in parentheses.

```{r}
#| label: tbl-r2
#| tbl-cap: "Probe R² at the final hidden layer for all 16 models, at the binomial level by attention condition. P = pair-held-out split; W = word-held-out split. SD across 10 folds in parentheses."
#| eval: !expr nrow(all_r2) > 0

all_r2 |>
  filter(mode == "mean_pooled") |>
  select(model, family, condition, mode, split, mean_r2, sd_r2, is_control) |>
  pivot_wider(names_from  = is_control,
              values_from = c(mean_r2, sd_r2),
              names_sep   = "_") |>
  rename(r2_main = mean_r2_FALSE, sd_main = sd_r2_FALSE,
         r2_ctrl = mean_r2_TRUE,  sd_ctrl = sd_r2_TRUE) |>
  mutate(
    condition = recode_condition(condition),
    model     = factor(model, levels = model_order),
    Split     = if_else(split == "pair_novel", "P", "W"),
    `R²`      = sprintf("%.3f (%.3f)", r2_main, sd_main),
    Ctrl      = sprintf("%.3f (%.3f)", r2_ctrl,  sd_ctrl),
  ) |>
  arrange(model, condition, split) |>
  select(Model = model, Condition = condition, Split, `R²`, Ctrl) |>
  kbl(booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 7, latex_options = c("scale_down", "HOLD_position")) |>
  collapse_rows(columns = 1:2, latex_hline = "major")
```

## Control Probe R² {#sec-appendix-ctrl-exp2}

Control probe R² for all 16 models at the final hidden layer, at the binomial level by attention condition.

```{r}
#| label: tbl-ctrl-r2
#| tbl-cap: "Control probe R² for all 16 models (final layer), at the binomial level by attention condition. SD across 10 folds in parentheses. P = pair-held-out; W = word-held-out."

all_r2 |>
  filter(mode == "mean_pooled", is_control) |>
  mutate(
    condition    = recode(condition, !!!condition_labels),
    model        = factor(model, levels = model_order),
    Split        = if_else(split == "pair_novel", "P", "W"),
    `Control R²` = sprintf("%.4f (%.4f)", mean_r2, sd_r2)
  ) |>
  arrange(family, model, condition, split) |>
  select(Family = family, Model = model, Condition = condition, Split, `Control R²`) |>
  kbl(booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 8, latex_options = "HOLD_position") |>
  collapse_rows(columns = 1:3, latex_hline = "major")
```

## Word-Level Replication {#sec-appendix-words-only-exp2}

Across all 16 models, word-level probing reproduces the main findings from Experiment 2. Pair-held-out R² is slightly lower than the corresponding binomial-level values (by roughly 5–10 percentage points for most Pile-trained models), but rank order across families and the generalization pattern (pair-held-out \> word-held-out) are preserved in every case. For GPT-2, pair-held-out R² under word-level probing ranges from 0.46 to 0.62 (vs. 0.49–0.65 at the binomial level); for Pythia, 0.51–0.57 (vs. 0.53–0.59); for OLMo, 0.52–0.57 (vs. 0.56–0.62); for Llama, 0.36–0.45 (vs. 0.43–0.55). The generalization gaps (pair-held-out minus word-held-out) range from 0.044 to 0.073 across Pile-trained models, closely matching the corresponding binomial-level gaps (0.035–0.080). $\hat{\beta}_2$ is positive for all 16 models, with magnitudes similar to the full-context binomial-level condition. Full results are in @tbl-r2-words-only.

```{r}
#| label: tbl-r2-words-only
#| tbl-cap: "Word-level probe R² and selectivity (main $-$ control R²) for all 16 models (final layer). P = pair-held-out split; W = word-held-out split. SD across 10 folds in parentheses."

all_r2 |>
  filter(mode == "words_only", condition == "default") |>
  select(family, model, condition, mode, split, mean_r2, sd_r2, is_control) |>
  pivot_wider(names_from  = is_control,
              values_from = c(mean_r2, sd_r2),
              names_sep   = "_") |>
  rename(r2_main = mean_r2_FALSE, sd_main = sd_r2_FALSE,
         r2_ctrl = mean_r2_TRUE,  sd_ctrl = sd_r2_TRUE) |>
  mutate(
    selectivity = r2_main - r2_ctrl,
    mode        = recode(mode, !!!mode_labels),
    model       = factor(model, levels = model_order),
    Split       = if_else(split == "pair_novel", "P", "W"),
    `R²`        = sprintf("%.3f (%.3f)", r2_main, sd_main),
    Ctrl        = sprintf("%.3f (%.3f)", r2_ctrl,  sd_ctrl),
    Sel.        = sprintf("%.3f", selectivity)
  ) |>
  arrange(family, model, split) |>
  select(Family = family, Model = model, Mode = mode, Split, `R²`, Ctrl, Sel.) |>
  kbl(booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 8, latex_options = "HOLD_position") |>
  collapse_rows(columns = 1:3, latex_hline = "major")
```

## Frequency-Residual Regression: All Models {#sec-appendix-full-beta}

Bayesian frequency-residual regression results for all 16 final-layer models, binomial-level input mode. 95% CIs in brackets. Vertex = $-\hat\beta_1 / (2\hat\beta_2)$, the log-frequency at which probe error is minimized.

```{r}
#| label: tbl-beta
#| tbl-cap: "Bayesian frequency-residual regression results for all 16 final-layer models, at the binomial level by attention condition. 95\\% CIs in brackets. Vertex = $-\\hat\\beta_1 / (2\\hat\\beta_2)$, the log-frequency at minimum probe error."
#| eval: !expr nrow(all_coefs) > 0

all_coefs |>
  filter(mode == "mean_pooled") |>
  mutate(
    model     = factor(label, levels = model_order),
    condition = recode_condition(condition),
    `$\\hat{\\beta}_1$` = sprintf("%.4f [%.4f, %.4f]",
                        b_quad_linear, b_quad_linear_lo, b_quad_linear_hi),
    `$\\hat{\\beta}_2$` = sprintf("%.5f [%.5f, %.5f]",
                        b_quad_sq, b_quad_sq_lo, b_quad_sq_hi),
    Vertex    = sprintf("%.2f", vertex)
  ) |>
  arrange(model, condition) |>
  select(Model = model, Condition = condition,
         `$\\hat{\\beta}_1$`, `$\\hat{\\beta}_2$`, Vertex) |>
  kbl(booktabs = TRUE, linesep = "", escape = FALSE) |>
  kable_styling(font_size = 7, latex_options = c("scale_down", "HOLD_position")) |>
  collapse_rows(columns = 1, latex_hline = "major")
```

## Predicted vs. Observed {#sec-appendix-scatter}

Predicted vs. observed binomial preference for all 16 models at the final hidden layer, full-context condition, binomial-level input mode. Each point is a corpus binomial pair colored by log-frequency tertile (low/medium/high). Perfect prediction falls on the diagonal. Models are ordered by family (BabyLM, GPT-2, Pythia, OLMo, Llama) and then by size within family.

```{r}
#| label: fig-scatter
#| fig-cap: "Predicted vs. observed binomial preference for all 16 models at the final hidden layer, full-context condition, binomial-level input mode. Each point is a corpus binomial pair colored by log-frequency tertile (low/medium/high). Perfect prediction falls on the diagonal."
#| fig-width: 6.5
#| fig-height: 9.0

scatter_slugs <- names(all_slugs)
scatter_labels <- unname(all_slugs)

freq_pile <- read_csv(
  file.path(RES, "corpus_binomials_infinigram_piletrain.csv"),
  show_col_types = FALSE
) |> select(word1, word2, freq_total)

freq_babylm <- read_csv(
  "../Data/corpus_binomials.csv",
  show_col_types = FALSE
) |>
  mutate(freq_total = freq_w1_w2 + freq_w2_w1) |>
  select(word1, word2, freq_total)

load_scatter <- function(slug, label) {
  xz  <- file.path(RES, slug, "by_layer_corpus_pred.csv.xz")
  csv <- file.path(RES, slug, "by_layer_corpus_pred.csv")
  if (!file.exists(xz) && !file.exists(csv)) return(NULL)
  pred <- if (file.exists(xz)) {
    con <- xzfile(xz, "rb"); d <- read_csv(con, show_col_types = FALSE); close(con); d
  } else {
    read_csv(csv, show_col_types = FALSE)
  }
  last_layer <- as.character(max(suppressWarnings(as.integer(unique(pred$layer))), na.rm = TRUE))
  freq_df <- if (grepl("babylm", slug)) freq_babylm else freq_pile
  pred |>
    filter(as.character(layer) == last_layer,
           condition == "default", mode == "mean_pooled") |>
    left_join(freq_df, by = c("word1", "word2")) |>
    filter(!is.na(freq_total), freq_total > 0) |>
    mutate(
      label    = label,
      log_freq = log(freq_total),
      freq_bin = cut(log_freq, breaks = 3,
                     labels = c("Low frequency", "Medium frequency", "High frequency"))
    )
}

scatter_data <- map_dfr(seq_along(scatter_slugs), function(i) {
  load_scatter(scatter_slugs[i], scatter_labels[i])
})

if (nrow(scatter_data) > 0) {
  scatter_data |>
    mutate(label = factor(label, levels = scatter_labels)) |>
    ggplot(aes(x = y_true, y = y_pred, color = freq_bin)) +
    geom_point(size = 0.2, alpha = 0.12) +
    geom_abline(slope = 1, intercept = 0,
                linetype = "dashed", color = "gray40", linewidth = 0.4) +
    facet_wrap(~label, nrow = 4) +
    scale_color_manual(
      values = c(
        "Low frequency"    = "#4477AA",
        "Medium frequency" = "#CCBB44",
        "High frequency"   = "#EE6677"
      )
    ) +
    labs(x = "Observed preference (log-odds)",
         y = "Predicted preference",
         color = "Corpus frequency") +
    theme_paper() +
    theme(legend.position = "bottom",
          axis.text = element_text(size = 6),
          strip.text = element_text(size = 7))
}
```

\clearpage

# Experiment 3: Supplementary Results {#sec-appendix-exp3}

This appendix provides supplementary materials for Experiment 3 (§\ref{sec-exp3}), which tracked probe R² and frequency-residual curvature across training checkpoints for BabyLM and Pythia. Subsections report full numerical checkpoint tables and a word-level replication.

## Checkpoint Probe Tables {#sec-appendix-ckpt-tables}

Full numerical results for the Experiment 3 checkpoint analysis. Tables report probe R² (pair-held-out and word-held-out) and frequency-residual regression coefficients ($\hat{\beta}_2$) for BabyLM and Pythia across training checkpoints, for the binomial-level (mean-pooled) representation under full-context and attention-zeroed conditions.

```{r}
#| echo: false

make_ckpt_r2_tbl <- function(df, cond_val, mode_val) {
  df |>
    filter(condition == cond_val, mode == mode_val,
           split %in% c("pair_novel", "word_novel")) |>
    select(model, step, tokens_M, split, mean_r2, sd_r2, is_control) |>
    pivot_wider(names_from = is_control, values_from = c(mean_r2, sd_r2),
                names_sep = "_") |>
    rename(r2_main = mean_r2_FALSE, sd_main = sd_r2_FALSE,
           r2_ctrl = mean_r2_TRUE,  sd_ctrl = sd_r2_TRUE) |>
    mutate(
      model  = factor(model, levels = model_order),
      Split  = if_else(split == "pair_novel", "P", "W"),
      Tokens = sprintf("%.0f M", tokens_M),
      `R²`   = sprintf("%.3f (%.3f)", r2_main, sd_main),
      Ctrl   = sprintf("%.3f (%.3f)", r2_ctrl,  sd_ctrl),
      Sel.   = sprintf("%.3f", r2_main - r2_ctrl)
    ) |>
    arrange(model, step, split) |>
    select(Model = model, Step = step, Tokens, Split, `R²`, Ctrl, Sel.) |>
    kbl(booktabs = TRUE, linesep = "") |>
    kable_styling(font_size = 8, latex_options = "HOLD_position") |>
    collapse_rows(columns = 1:2, latex_hline = "major")
}

make_ckpt_beta_tbl <- function(df, cond_val, mode_val) {
  df |>
    filter(condition == cond_val, mode == mode_val) |>
    mutate(
      model = factor(label, levels = model_order),
      `$\\hat{\\beta}_2$` = sprintf("%.4f", beta2_mean),
      SE      = sprintf("%.4f", beta2_se),
      CI      = sprintf("[%.4f, %.4f]", beta2_ci025, beta2_ci975),
      `P(>0)` = sprintf("%.1f\\%%", beta2_pct_gt0)
    ) |>
    arrange(model, tokens_M) |>
    mutate(Tokens = sprintf("%.0f M", tokens_M)) |>
    select(Model = model, Step = step, Tokens,
           `$\\hat{\\beta}_2$`, SE, CI, `P(>0)`) |>
    kbl(booktabs = TRUE, linesep = "", escape = FALSE) |>
    kable_styling(font_size = 8, latex_options = "HOLD_position") |>
    collapse_rows(columns = 1, latex_hline = "major")
}
```

```{r}
#| label: tbl-r2-checkpoints-babylm-default-mean
#| tbl-cap: "BabyLM checkpoint probe R²:full context, binomial-level. P = pair-held-out; W = word-held-out. SD in parentheses."

if (nrow(babylm_ckpt_r2) > 0)
  make_ckpt_r2_tbl(babylm_ckpt_r2, "default", "mean_pooled")
```

```{r}
#| label: tbl-r2-checkpoints-babylm-attn-mean
#| tbl-cap: "BabyLM checkpoint probe R²:attention-zeroed, binomial-level."

if (nrow(babylm_ckpt_r2) > 0)
  make_ckpt_r2_tbl(babylm_ckpt_r2, "attn_zeroed", "mean_pooled")
```

```{r}
#| label: tbl-r2-checkpoints-pythia-default-mean
#| tbl-cap: "Pythia checkpoint probe R²:full context, binomial-level."

if (nrow(pythia_ckpt_r2) > 0)
  make_ckpt_r2_tbl(pythia_ckpt_r2, "default", "mean_pooled")
```

```{r}
#| label: tbl-r2-checkpoints-pythia-attn-mean
#| tbl-cap: "Pythia checkpoint probe R²:attention-zeroed, binomial-level."

if (nrow(pythia_ckpt_r2) > 0)
  make_ckpt_r2_tbl(pythia_ckpt_r2, "attn_zeroed", "mean_pooled")
```

```{r}
#| label: tbl-beta-checkpoints-babylm-default-mean
#| tbl-cap: "BabyLM checkpoint frequency-residual regression ($\\hat{\\beta}_2$):full context, binomial-level. SE = posterior SD; CI = 95\\% credible interval; P(>0) = percentage of posterior samples with $\\hat{\\beta}_2 > 0$."

if (nrow(babylm_ckpt_coefs) > 0)
  make_ckpt_beta_tbl(babylm_ckpt_coefs, "default", "mean_pooled")
```

```{r}
#| label: tbl-beta-checkpoints-babylm-attn-mean
#| tbl-cap: "BabyLM checkpoint frequency-residual regression ($\\hat{\\beta}_2$):attention-zeroed, binomial-level."

if (nrow(babylm_ckpt_coefs) > 0)
  make_ckpt_beta_tbl(babylm_ckpt_coefs, "attn_zeroed", "mean_pooled")
```

```{r}
#| label: tbl-beta-checkpoints-pythia-default-mean
#| tbl-cap: "Pythia checkpoint frequency-residual regression ($\\hat{\\beta}_2$):full context, binomial-level."

if (nrow(pythia_ckpt_coefs) > 0)
  make_ckpt_beta_tbl(pythia_ckpt_coefs, "default", "mean_pooled")
```

```{r}
#| label: tbl-beta-checkpoints-pythia-attn-mean
#| tbl-cap: "Pythia checkpoint frequency-residual regression ($\\hat{\\beta}_2$):attention-zeroed, binomial-level."

if (nrow(pythia_ckpt_coefs) > 0)
  make_ckpt_beta_tbl(pythia_ckpt_coefs, "attn_zeroed", "mean_pooled")
```

## Word-Level Replication {#sec-appendix-words-only-exp3}

Word-level probe results across BabyLM and Pythia checkpoints replicate the key patterns from the binomial-level analysis. The generalization gap (pair-held-out minus word-held-out) is largest at the earliest checkpoints and shrinks over the course of training. For BabyLM, the gap at the first checkpoint ranges from 0.15 to 0.36 across model sizes, narrowing to 0.13–0.14 by the final checkpoint, the same qualitative trajectory observed at the binomial level. $\hat{\beta}_2$ shows the same sign trajectory as in the main analysis: negative early in BabyLM training, transitioning to positive after approximately 750M tokens. Full R² results are in @tbl-r2-checkpoints-babylm-default-words and @tbl-r2-checkpoints-pythia-default-words; frequency-residual regression results are in @tbl-beta-checkpoints-babylm-default-words and @tbl-beta-checkpoints-pythia-default-words.

```{r}
#| label: tbl-r2-checkpoints-babylm-default-words
#| tbl-cap: "BabyLM checkpoint probe R²:full context, word-level. P = pair-held-out; W = word-held-out. SD in parentheses."

if (nrow(babylm_ckpt_r2) > 0)
  make_ckpt_r2_tbl(babylm_ckpt_r2, "default", "words_only")
```

```{r}
#| label: tbl-r2-checkpoints-pythia-default-words
#| tbl-cap: "Pythia checkpoint probe R²:full context, word-level. P = pair-held-out; W = word-held-out. SD in parentheses."

if (nrow(pythia_ckpt_r2) > 0)
  make_ckpt_r2_tbl(pythia_ckpt_r2, "default", "words_only")
```

```{r}
#| label: tbl-beta-checkpoints-babylm-default-words
#| tbl-cap: "BabyLM checkpoint frequency-residual regression ($\\hat{\\beta}_2$):full context, word-level. SE = posterior SD; CI = 95\\% credible interval; P(>0) = percentage of posterior samples with $\\hat{\\beta}_2 > 0$."

if (nrow(babylm_ckpt_coefs) > 0)
  make_ckpt_beta_tbl(babylm_ckpt_coefs, "default", "words_only")
```

```{r}
#| label: tbl-beta-checkpoints-pythia-default-words
#| tbl-cap: "Pythia checkpoint frequency-residual regression ($\\hat{\\beta}_2$):full context, word-level. SE = posterior SD; CI = 95\\% credible interval; P(>0) = percentage of posterior samples with $\\hat{\\beta}_2 > 0$."

if (nrow(pythia_ckpt_coefs) > 0)
  make_ckpt_beta_tbl(pythia_ckpt_coefs, "default", "words_only")
```

\clearpage

# Corpus Frequency Distributions {#sec-appendix-corpus-freq}

The figure below compares the frequency distributions of the approximately 47,000 corpus binomials in the BabyLM training corpus (150M tokens, 20 epochs) and the Pile, illustrating the stark difference in lexical coverage between the two training regimes. BabyLM's distribution is dominated by singleton pairs (70.7% appear exactly once), while Pile frequencies span several orders of magnitude.

```{r}
#| label: fig-corpus-freq
#| fig-cap: "Frequency distributions for the ~47K corpus binomials in the BabyLM training corpus (150M tokens, 20 epochs) and the Pile. Left: histogram of log$_{10}$ raw counts. Right: rank-frequency plot treating counts as a probability distribution (probability = count / $\\sum$ counts); both axes log-scaled. BabyLM's distribution is heavily concentrated at a single occurrence per pair (70.7% of pairs appear exactly once), whereas Pile frequencies span several orders of magnitude with a median of 365 occurrences per pair."
#| fig-width: 7
#| fig-height: 3.2

corpus_freq_df <- local({
  babylm_raw <- read_csv("../Data/corpus_binomials.csv", show_col_types = FALSE) |>
    mutate(freq = freq_w1_w2 + freq_w2_w1) |>
    filter(freq > 0) |>
    select(word1, word2, freq) |>
    mutate(corpus = "BabyLM (150M tok, 20 ep)")

  pile_raw <- read_csv("../Results/corpus_binomials_infinigram_piletrain.csv",
                       show_col_types = FALSE) |>
    rename(freq = freq_total) |>
    filter(freq > 0) |>
    select(word1, word2, freq) |>
    mutate(corpus = "Pile (~300B tok, 1 ep)")

  bind_rows(babylm_raw, pile_raw) |>
    group_by(corpus) |>
    mutate(prob = freq / sum(freq),
           rank = rank(-freq, ties.method = "first")) |>
    ungroup()
})

corpus_colors <- c("BabyLM (150M tok, 20 ep)" = "#154733",
                   "Pile (~300B tok, 1 ep)"    = "#F0B323")

p_hist <- corpus_freq_df |>
  ggplot(aes(x = log10(freq), fill = corpus, color = corpus)) +
  geom_histogram(aes(y = after_stat(count)),
                 binwidth = 0.25, position = "identity", alpha = 0.55) +
  scale_fill_manual(values  = corpus_colors, name = NULL) +
  scale_color_manual(values = corpus_colors, name = NULL) +
  scale_y_continuous(labels = label_comma()) +
  labs(x = expression(log[10]~"(count)"), y = "Number of pairs") +
  theme_paper() +
  theme(legend.position = "none")

p_rank <- corpus_freq_df |>
  ggplot(aes(x = rank, y = prob, color = corpus)) +
  geom_line(linewidth = 0.7, alpha = 0.85) +
  scale_x_log10(labels = label_comma()) +
  scale_y_log10(labels = label_scientific()) +
  scale_color_manual(values = corpus_colors, name = NULL) +
  labs(x = "Rank (log scale)", y = "Probability (log scale)") +
  theme_paper() +
  theme(legend.position = "bottom")

p_hist + p_rank + plot_layout(ncol = 2, widths = c(1, 1))
```

The BabyLM training corpus is a 150M-token collection of child-directed and informal text (CHILDES, BNC Spoken, children's books, OpenSubtitles, Wikipedia), trained for 20 epochs. The Pile is a \~300B-token web-scraped corpus used to train the Pythia family. For each of the \~47K corpus binomials, we extracted 3-gram counts (i.e., the frequency of "word$_1$ and word$_2$" and its reverse) from each corpus via infinigram. As @fig-corpus-freq shows, the BabyLM corpus produces a highly degenerate frequency distribution: 70.7% of pairs appear exactly once, 87.4% appear at most twice, and 98.2% appear at most ten times, with a maximum of 1,095 occurrences. The Pile distribution is far richer: the median pair appears 365 times, the 25th percentile is 62 occurrences, and the 99th percentile exceeds 77,000.

In absolute training exposure terms (accounting for BabyLM's 20 epochs), the most frequent BabyLM binomials are seen at most \~21,900 times total across all training steps, while the equivalent Pile binomials are seen over 1.5 million times in a single pass. Even the median Pile pair (365 occurrences per pass) substantially exceeds BabyLM's total exposure for a typical singleton pair (1 × 20 = 20 exposures). This gap in absolute exposure is the proximate explanation for why idiosyncratic memorization (reflected in positive $\hat{\beta}_2$) emerges immediately in Pythia but requires hundreds of millions of tokens to appear in BabyLM.
