---
title: "Stored or Computed? Abstract and Memorized Binomial Preferences in LLM Representations"
acl-mode: review
format:
  acl-pdf:
    keep-tex: true
    cite-method: natbib
  html: default
editor: visual
author:
  - name:
      literal: Zachary Nicholas Houghton
    email: znh@uoregon.edu
    affiliations:
      - name: University of Oregon
      - name: Vail Systems, Inc.

abstract: |
  TODO

bibliography: references.bib
tbl-cap-location: bottom
header-includes:
  - \usepackage{float}
  - \floatplacement{table}{H}
  - \usepackage[section]{placeins}
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)
library(kableExtra)
library(scales)

RES       <- "../Results"
MODEL_DIR <- "../Data/brms_models"

# ── Slug → label map ──────────────────────────────────────────────────────────
all_slugs <- c(
  "znhoughton_opt-babylm-125m-20eps-seed964" = "BabyLM-125M",
  "znhoughton_opt-babylm-350m-20eps-seed964" = "BabyLM-350M",
  "znhoughton_opt-babylm-1_3b-20eps-seed964" = "BabyLM-1.3B",
  "EleutherAI_pythia-160m"                   = "Pythia-160M",
  "EleutherAI_pythia-410m"                   = "Pythia-410M",
  "EleutherAI_pythia-1b"                     = "Pythia-1B",
  "gpt2"                                     = "GPT-2",
  "gpt2-medium"                              = "GPT-2-M",
  "gpt2-large"                               = "GPT-2-L",
  "gpt2-xl"                                  = "GPT-2-XL",
  "allenai_OLMo-1B-hf"                       = "OLMo-1B",
  "allenai_OLMo-7B-hf"                       = "OLMo-7B",
  "allenai_OLMo-2-0425-1B"                   = "OLMo-2-1B",
  "allenai_OLMo-2-1124-7B"                   = "OLMo-2-7B",
  "meta-llama_Llama-3.2-1B"                  = "Llama-1.3B",
  "meta-llama_Meta-Llama-3-8B"               = "Llama-3-8B"
)

family_map <- c(
  "BabyLM-125M" = "BabyLM", "BabyLM-350M" = "BabyLM", "BabyLM-1.3B" = "BabyLM",
  "Pythia-160M" = "Pythia", "Pythia-410M"  = "Pythia", "Pythia-1B"   = "Pythia",
  "GPT-2"       = "GPT-2",  "GPT-2-M"      = "GPT-2",  "GPT-2-L"     = "GPT-2",
  "GPT-2-XL"   = "GPT-2",
  "OLMo-1B"    = "OLMo",   "OLMo-7B"      = "OLMo",   "OLMo-2-1B"   = "OLMo",
  "OLMo-2-7B"  = "OLMo",
  "Llama-1.3B" = "Llama",  "Llama-3-8B"   = "Llama"
)

# ── Training-scale metadata ───────────────────────────────────────────────────
model_meta <- tribble(
  ~label,         ~params_M,  ~training_tokens_B,
  "BabyLM-125M",  125,         0.15,
  "BabyLM-350M",  350,         0.15,
  "BabyLM-1.3B",  1300,        0.15,
  "GPT-2",        117,         8.0,
  "GPT-2-M",      345,         8.0,
  "GPT-2-L",      762,         8.0,
  "GPT-2-XL",     1500,        8.0,
  "Pythia-160M",  160,         300.0,
  "Pythia-410M",  410,         300.0,
  "Pythia-1B",    1000,        300.0,
  "OLMo-1B",      1000,        3000.0,
  "OLMo-7B",      7000,        2500.0,
  "OLMo-2-1B",    1000,        4000.0,
  "OLMo-2-7B",    7000,        4000.0,
  "Llama-1.3B",   1200,        9000.0,
  "Llama-3-8B",   8000,        15000.0
)

# ── Load MLP R² from per-model CSVs ──────────────────────────────────────────
load_mlp <- function(slug, label, is_control = FALSE) {
  fname <- if (is_control) "by_layer_mlp_control.csv" else "by_layer_mlp.csv"
  path  <- file.path(RES, slug, fname)
  if (!file.exists(path)) return(NULL)
  read_csv(path, show_col_types = FALSE) |>
    mutate(model = label, family = family_map[label], is_control = is_control)
}

all_r2 <- map_dfr(names(all_slugs), function(slug) {
  bind_rows(
    load_mlp(slug, all_slugs[[slug]], FALSE),
    load_mlp(slug, all_slugs[[slug]], TRUE)
  )
}) |>
  filter(split %in% c("pair_novel", "word_novel")) |>
  group_by(model, family, condition, mode, split, is_control) |>
  filter(layer == max(layer)) |>
  ungroup()

# ── Precomputed brms summaries (from Scripts/prepare_results.R) ───────────────
brms_summary <- readRDS(file.path(MODEL_DIR, "..", "brms_summary.rds"))
brms_curves  <- readRDS(file.path(MODEL_DIR, "..", "brms_curves.rds"))

# ── Checkpoint slug registries ────────────────────────────────────────────────
babylm_ckpt_registry <- bind_rows(
  tibble(slug  = paste0("znhoughton_opt-babylm-125m-20eps-seed964_step",
                        c(24, 48, 144, 384, 912, 2280)),
         label = "BabyLM-125M",
         step  = c(24, 48, 144, 384, 912, 2280),
         tokens_M = c(24, 48, 144, 384, 912, 2280) * 0.8192),  # 1024 * 400 * 2 / 1e6
  tibble(slug  = paste0("znhoughton_opt-babylm-350m-20eps-seed964_step",
                        c(48, 96, 288, 768, 1824, 4560)),
         label = "BabyLM-350M",
         step  = c(48, 96, 288, 768, 1824, 4560),
         tokens_M = c(48, 96, 288, 768, 1824, 4560) * 0.4096),  # 1024 * 200 * 2 / 1e6
  tibble(slug  = paste0("znhoughton_opt-babylm-1.3b-20eps-seed964_step",
                        c(97, 194, 582, 1455, 3686, 9021)),
         label = "BabyLM-1.3B",
         step  = c(97, 194, 582, 1455, 3686, 9021),
         tokens_M = c(97, 194, 582, 1455, 3686, 9021) * 0.2048)  # 1024 * 100 * 2 / 1e6
)

pythia_ckpt_registry <- bind_rows(
  tibble(slug  = paste0("EleutherAI_pythia-160m_step", c(16, 32, 64, 256, 512, 1000)),
         label = "Pythia-160M",
         step  = c(16, 32, 64, 256, 512, 1000),
         tokens_M = c(16, 32, 64, 256, 512, 1000) * 2.097152),
  tibble(slug  = paste0("EleutherAI_pythia-410m_step", c(16, 32, 64, 256, 512, 1000)),
         label = "Pythia-410M",
         step  = c(16, 32, 64, 256, 512, 1000),
         tokens_M = c(16, 32, 64, 256, 512, 1000) * 2.097152),
  tibble(slug  = paste0("EleutherAI_pythia-1b_step",   c(16, 32, 64, 256, 512, 1000)),
         label = "Pythia-1B",
         step  = c(16, 32, 64, 256, 512, 1000),
         tokens_M = c(16, 32, 64, 256, 512, 1000) * 2.097152),
)

# ── UO-brand colour palettes ──────────────────────────────────────────────────
# Oregon Green: #154733 / #007030  UO Gold: #F0B323  Amber: #875E00  UO Blue: #005581
family_colors <- c(
  "BabyLM" = "#154733",
  "GPT-2"  = "#F0B323",
  "Pythia" = "#007030",
  "OLMo"   = "#875E00",
  "Llama"  = "#005581"
)

babylm_size_colors <- c(
  "BabyLM-125M" = "#F0B323",
  "BabyLM-350M" = "#007030",
  "BabyLM-1.3B" = "#154733"
)

# ── Factor order for model labels ──────────────────────────────────────────────
model_order <- c(
  "BabyLM-125M", "BabyLM-350M", "BabyLM-1.3B",
  "GPT-2", "GPT-2-M", "GPT-2-L", "GPT-2-XL",
  "Pythia-160M", "Pythia-410M", "Pythia-1B",
  "OLMo-1B", "OLMo-7B", "OLMo-2-1B", "OLMo-2-7B",
  "Llama-1.3B", "Llama-3-8B"
)

# ascending params within family (for dot-plot row order)
dumbbell_order <- c(
  "BabyLM-125M", "BabyLM-350M", "BabyLM-1.3B",
  "GPT-2", "GPT-2-M", "GPT-2-L", "GPT-2-XL",
  "Pythia-160M", "Pythia-410M", "Pythia-1B",
  "OLMo-1B", "OLMo-2-1B", "OLMo-7B", "OLMo-2-7B",
  "Llama-1.3B", "Llama-3-8B"
)

condition_labels <- c(
  "default"     = "Full context",
  "attn_zeroed" = "Attention-zeroed"
)

recode_condition <- function(x) {
  factor(recode(x, !!!condition_labels),
         levels = c("Full context", "Attention-zeroed"))
}

mode_labels <- c(
  "mean_pooled" = "Binomial-level",
  "words_only"  = "Word-level"
)

theme_paper <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      text             = element_text(color = "black"),
      axis.text        = element_text(color = "black", size = rel(0.9)),
      axis.title       = element_text(color = "black", size = rel(1.0)),
      legend.text      = element_text(color = "black", size = rel(0.9)),
      legend.title     = element_text(color = "black", size = rel(1.0)),
      strip.text       = element_text(color = "black", face = "bold", size = rel(1.0)),
      panel.grid.minor = element_blank(),
      axis.ticks       = element_line(color = "black"),
      axis.ticks.length = unit(3, "pt")
    )
}
```

# Introduction

# Experiment 1: BabyLM Representations {#sec-exp1}

In this first experiment, we probe three BabyLM models (125M, 350M, and 1.3B parameters)[^1] at their final training checkpoint. These three models were trained on an amount of data comparable to humans and thus are more directly comparable to the human behavioral literature. Because the binomials used in this study were selected to be absent from BabyLM's training corpus, this experiment provides the cleanest test of whether a language model encodes abstract binomial ordering preferences: any preference signal the probe recovers must have been generalized from the training distribution rather than read off from direct exposure to the test pairs.

[^1]: The BabyLM models used here are OPT-based models trained for 20 epochs on a 150M-token subset of the BabyLM benchmark corpus; @houghton2026holisticstorageverbupphrases; @warstadt-etal-2023-findings\].

<!--# Figure 2 "Word-held-out" is being cut off -->

## Methods

Below we describe the methods used in Experiment 1.

### Stimuli

We use two complementary datasets of binomial expressions (*X and Y*). Binomials are a strong test case because they are well-studied in the linguistics literature [e.g., @morgan2016; @benor2006] and human ordering preferences for them are driven by abstract preferences [@morgan2015; @morgan2016]. The **novel set** comprises approximately 340,000 binomial pairs from the Wikitext corpus [@merity2016pointer] selected to be absent from BabyLM's training corpus. The novel pairs are therefore genuinely novel for the BabyLM models used in Experiment 1.

The **attested corpus set** comprises approximately 49,000 binomial pairs from the BabyLM corpus [@warstadt-etal-2023-findings]. For each corpus pair and each model, the ordering preference is operationalized as the model's own log-probability ratio: $y_{\text{true}} = \log P(W_1\text{ and }W_2) - \log P(W_2\text{ and }W_1)$.

### Representation Extraction

For each model, we extract hidden-state representations from the final hidden-layer. Each binomial pair is presented to the model in both orderings (*W*~1~ and *W*~2~; *W*~2~ and *W*~1~) within a natural sentence frame. We extract representations under two input modes:

-   **Binomial-level** (`mean_pooled`): the input to the probe is $[\bar{h}(W_1\text{ and }W_2); \bar{h}(W_2\text{ and }W_1)]$, where $\bar{h}$ denotes the mean of token representations across all tokens in the binomial in each ordering.
-   **Word-level** (`words_only`): the input is $[h(W_1); h(W_2)]$, where $h(W_k)$ is the hidden state of word $W_k$ extracted from the sentence in which it appears in first position, excluding the conjunction token.

### Attention Conditions

We probe representations under two attention conditions:

-   **Full context** (`default`): standard forward pass with full cross-token attention. Each word's representation reflects contextual influences from all other tokens in the sentence.
-   **Attention-zeroed** (`attn_zeroed`): attention from the second word and the conjunction to the first word is blocked by setting the corresponding attention logits to $-\infty$ before the softmax. This condition tests whether the second word's representation is specifically shaped by direct cross-word attention to the first word.

Because the word-level mode extracts each word from the sentence in which it occupies first position, its representation depends only on the preceding sentence context and is not influenced by any later token's attention. We therefore report three representation types per model: full-context binomial-level, attention-zeroed binomial-level, and full-context word-level.

### MLP Probe

For each representation type, we train a two-layer MLP classifier (hidden dimension 128, ReLU activation) to predict binomial preference from the extracted representations. The probe takes as its input the concatenated hidden-layer representations for a given binomial (dimensionality and construction determined by the representation type, as operationalized above) and outputs a single continuous value $\hat{y}$, trained via mean-squared-error loss to approximate the true preference score, $y_{true}$, defined in the Stimuli section. The model is trained with Adam (learning rate $10^{-3}$, weight decay $10^{-4}$) with early stopping (patience = 20) and a 10% held-out validation set. Training uses antisymmetric augmentation: for each pair, we add a flipped version (swapped ordering, negated label) with probability 0.5. We report cross-validated R² (the squared Pearson correlation between $\hat{y}$ and $y_{true}$ on held-out folds) as our measure of probe performance, averaged across 10 folds.

We evaluate generalization under two cross-validation splits:

-   **Pair-held-out** (`pair_novel`): pairs are held out as units; both orderings of a pair appear only in training or only in test.
-   **Word-held-out** (`word_novel`): all pairs containing a test word are excluded from training, testing generalization to completely novel lexical items.

### Control Probes

Following @hewitt-liang-2019-designing, we train a matched control probe on randomly permuted preference labels under the same architecture and hyperparameters. Near-zero control R² provides evidence that the probe is extracting genuine linguistic structure rather than exploiting label artifacts or probe capacity. Full control R² results are reported in appendix sections alongside each experiment.

### Frequency-Residual Regression {#sec-methods-freq-regression}

To characterize how corpus frequency relates to probe accuracy, we regress the absolute residual error of the MLP probe on log corpus frequency using a Bayesian quadratic regression model implemented in *brms* [@burkner2017brms] with default priors. Frequency was estimated using exact counts of the binomial in the BabyLM training corpus [@warstadt-etal-2023-findings]. Our statistical model equation is presented below:

$$|y_{\text{true}} - \hat{y}| \sim \beta_0 + \beta_1 \log(\text{freq}) + \beta_2 \log(\text{freq})^2$$ {#eq-freq-regression}

```{r}
#| label: load-brms-babylm
#| include: false

babylm_labels <- c("BabyLM-125M", "BabyLM-350M", "BabyLM-1.3B")

babylm_curves <- brms_curves |>
  filter(label %in% babylm_labels, is.na(step))

babylm_coefs <- brms_summary |>
  filter(label %in% babylm_labels, is.na(step)) |>
  left_join(model_meta, by = "label") |>
  mutate(family = family_map[label])
```

```{r}
#| label: ctrl-stats
#| include: false

ctrl_r2_exp1 <- all_r2 |>
  filter(family == "BabyLM", is_control, mode == "mean_pooled") |>
  pull(mean_r2)

ctrl_r2_exp2 <- all_r2 |>
  filter(is_control, mode == "mean_pooled") |>
  pull(mean_r2)

ctrl_mean_exp1 <- mean(ctrl_r2_exp1)
ctrl_sd_exp1   <- sd(ctrl_r2_exp1)
ctrl_mean_exp2 <- mean(ctrl_r2_exp2)
ctrl_sd_exp2   <- sd(ctrl_r2_exp2)
```

## Results

### Probe Performance

Results are visualized in @fig-r2-babylm; full numerical results are in Appendix [-@sec-appendix-exp1]. The control probe R² values were consistently near zero across all conditions and all three models (mean ± SD = `r sprintf("%.4f ± %.4f", ctrl_mean_exp1, ctrl_sd_exp1)`; full control results are in Appendix [-@sec-appendix-exp1]), confirming that the probe doesn't recover spurious signal from label noise.

For the binomial-level conditions, pair-held-out R² ranges from 0.24 to 0.34; word-held-out R² is consistently lower, ranging from 0.12 to 0.23. The drop from pair-held-out to word-held-out reflects a lexical generalization cost: the probe is more accurate when the words in a test pair appeared in other pairs during probe training than when both words are entirely novel, suggesting that some preference signal is tied to word identity rather than being fully abstract. Probe accuracy is non-monotonic with model size: BabyLM-350M achieves the highest pair-held-out R² across most conditions, with BabyLM-125M and BabyLM-1.3B lower by similar margins. The full-context condition yields higher pair-held-out R² than attention-zeroed, indicating that cross-word attention contributes positively to preference encoding for these novel binomials. We return to possible explanations for this in the discussion section.

```{r}
#| label: fig-r2-babylm
#| fig-cap: "Cross-validated R² at the final hidden layer for BabyLM models, at the binomial level by attention condition. Circles = pair-held-out; triangles = word-held-out. Bars = ±1 SD (minimum width enforced for visibility)."
#| fig-width: 3.3
#| fig-height: 2.8
#| eval: !expr nrow(all_r2) > 0

all_r2 |>
  filter(family == "BabyLM", mode == "mean_pooled", !is_control) |>
  mutate(
    model      = factor(model, levels = model_order),
    size_lbl   = factor(sub("BabyLM-", "", as.character(model)),
                        levels = c("125M", "350M", "1.3B")),
    display_sd = pmax(sd_r2, 0.010),
    condition  = recode_condition(condition),
    panel      = factor(recode(condition,
                               "Full context"     = "Full context",
                               "Attention-zeroed" = "Attention-zeroed"),
                        levels = c("Full context", "Attention-zeroed"))
  ) |>
  ggplot(aes(x = size_lbl, y = mean_r2, color = model, shape = split)) +
  geom_pointrange(aes(ymin = mean_r2 - display_sd, ymax = mean_r2 + display_sd),
                  position = position_dodge(0.5), linewidth = 0.5, size = 0.35) +
  facet_wrap(~panel, nrow = 1) +
  scale_color_manual(values = babylm_size_colors, guide = "none") +
  scale_shape_manual(values = c("pair_novel" = 19, "word_novel" = 17),
                     labels = c("pair_novel" = "Pair-held-out",
                                "word_novel" = "Word-held-out"),
                     name = NULL) +
  labs(x = "Size", y = expression(R^2)) +
  theme_paper(base_size = 10) +
  theme(legend.position = "bottom",
        axis.text.x     = element_text(angle = 30, hjust = 1, size = 8))
```

### Memorization vs. Abstraction

We report the results of the frequency-residual regression results here. Full numerical results and fitted curves are in Appendix [-@sec-appendix-exp1]. Under the full-context condition, the quadratic coefficient $\hat\beta_2$ is positive for all three models, with 95% credible intervals excluding zero. This indicates a U-shaped relationship between corpus frequency and probe error: the probe makes the smallest errors for pairs at intermediate log-frequency values, with errors increasing toward both the low- and high-frequency tails. Under the attention-zeroed condition, the curvature is attenuated relative to full context and remains reliably positive for BabyLM-125M and BabyLM-350M; though the effect falls below statistical significance levels for BabyLM-1.3B. The attenuation under the attention-zeroed condition indicates that the high-frequency penalty on probe accuracy is driven in part by cross-word attention: blocking the second word's attention to the first flattens the frequency-residual relationship.

## Discussion

The results of Experiment 1 demonstrate that, contrary to the results of previous research [e.g., @houghton2025role], LLMs are actually learning some abstract preferences. This is demonstrated by the positive $r^2$ values for held-out pairs, and go beyond simple lexical preferences (such as simply preferring the word *bread* first in general), evidenced by the positive $r^2$ values for the novel word condition where the probe generalizes to binomials whose words were not in the probe's training data.

The results also demonstrate that the BabyLMs, in addition to using abstract preferences, are also still memorizing a good deal. This is evidenced by the frequency-residual regression model, where the probe's preference prediction is worst for high- and low-frequency items. For low-frequency binomials, this decrease in probe accuracy is likely due to underdeveloped lexical representations (i.e., noisy-representations). For higher-frequency binomials, this is likely due to attention effects reflecting context-specific representations (i.e., memorization).

# Experiment 2: Cross-Family Final-Layer Representations {#sec-exp2}

Experiment 1 demonstrates that BabyLMs are relying on both abstract preferences and memorization for binomial ordering preferences. However, those are models that are both smaller in terms of number of parameters as well as in terms of training data. In Experiment 2, we extend this to a larger variety of models.

## Methods

In Experiment 2, we probe all 16 models spanning BabyLM, Pythia, GPT-2, OLMo, and Llama families (117M to 8B parameters, trained on 150M to approximately 15T tokens) at their final training checkpoints. One design consideration differs from Experiment 1: the novel binomials were selected to be absent from BabyLM's training corpus, but they are not truly novel for the Pile-trained models; approximately 98% appear in the Pile at non-zero frequency, most at low-to-medium exposure (Pile-equivalent counts spanning 14 to 443 occurrences for the middle 50% of items). The probe R² results for these models therefore reflect representations of low-to-mid-frequency attested pairs rather than those of novel ones; validity checks are discussed in @sec-exp2-validity. Probe architecture, attention conditions, and frequency regression are identical to Experiment 1.

<!--# Framing is a bit odd because we aren't probing BabyLM here, we already did. I'm wondering if we should also give the full range of frequencies, not just the middle 50% of items. -->

## Results

### Probe Performance

Full numerical results are in Appendix @tbl-r2; @fig-r2 visualizes the pair-held-out results. Control probe R² was consistently near zero across all models and conditions (mean ± SD = `r sprintf("%.4f ± %.4f", ctrl_mean_exp2, ctrl_sd_exp2)`; full control results in Appendix [-@sec-appendix-exp2]). As in Experiment 1, the binomial-level conditions (full-context and attention-zeroed) are directly comparable; word-level results are provided in Appendix [-@sec-appendix-exp2]. All Pile-trained models substantially outperform BabyLM: pair-held-out R² under the full-context, binomial-level condition ranges from 0.43 to 0.65 across the four Pile-trained families, compared to 0.29–0.34 for BabyLM. Word-held-out R² is consistently lower than pair-held-out R² for every model and condition, matching the pattern in Experiment 1. GPT-2 shows the most striking result: under the attention-zeroed condition, pair-held-out R² reaches 0.79 for GPT-2 (117M) — substantially higher than its full-context value of 0.65 — with attention-condition advantages across all four GPT-2 models: GPT-2-M (0.76 vs. 0.53), GPT-2-L (0.67 vs. 0.52), and GPT-2-XL (0.73 vs. 0.49). Pythia, OLMo, and Llama-1.3B show the same direction: attention-zeroed outperforms full context at the binomial level. Llama-3-8B is an exception, matching the BabyLM pattern with full context slightly outperforming attention-zeroed (0.54 vs. 0.52). Scale effects within families are non-monotonic: in GPT-2, R² decreases with model size under full context (117M highest at 0.65, XL lowest at 0.49); in Pythia, the pattern is non-monotonic (160M highest at 0.59, 410M lowest at 0.53, 1B at 0.55); OLMo's 7B models slightly outperform 1B counterparts; Llama's 8B model outperforms 1.3B.

```{r}
#| label: fig-r2
#| fig-cap: "Cross-validated R² at the final hidden layer for all 16 models (binomial-level), by attention condition (columns) and cross-validation split (shape). Circles = pair-held-out; triangles = word-held-out. Bars = 95\\% CI across 10 folds ($t$-distribution, $df = 9$). Models ordered by parameter count within each family and colored by family. Control R² $\\approx$ 0 (Appendix [-@sec-appendix-exp2])."
#| fig-width: 6.5
#| fig-height: 2.8
#| fig-env: "figure*"
#| eval: !expr nrow(all_r2) > 0

size_labels <- c(
  "BabyLM-125M" = "125M", "BabyLM-350M" = "350M", "BabyLM-1.3B" = "1.3B",
  "GPT-2" = "S", "GPT-2-M" = "M", "GPT-2-L" = "L", "GPT-2-XL" = "XL",
  "Pythia-160M" = "160M", "Pythia-410M" = "410M", "Pythia-1B" = "1B",
  "OLMo-1B" = "1B", "OLMo-7B" = "7B", "OLMo-2-1B" = "2-1B", "OLMo-2-7B" = "2-7B",
  "Llama-1.3B" = "1.3B", "Llama-3-8B" = "8B"
)

all_r2 |>
  filter(mode == "mean_pooled", !is_control) |>
  mutate(
    condition = recode_condition(condition),
    split_lbl = factor(recode(split,
                              "pair_novel" = "Pair-held-out",
                              "word_novel" = "Word-held-out"),
                       levels = c("Pair-held-out", "Word-held-out")),
    family    = factor(family, levels = c("BabyLM", "GPT-2", "Pythia", "OLMo", "Llama")),
    model_fct = factor(model, levels = dumbbell_order),
    ci_half   = qt(0.975, df = 9) * sd_r2 / sqrt(n_folds),
    r2_lo     = mean_r2 - pmax(ci_half, 0.02),
    r2_hi     = mean_r2 + pmax(ci_half, 0.02)
  ) |>
  ggplot(aes(x = model_fct, y = mean_r2, color = family, shape = split_lbl)) +
  geom_linerange(aes(ymin = r2_lo, ymax = r2_hi), linewidth = 0.7,
                 position = position_dodge(width = 0.6)) +
  geom_point(size = 2.0, position = position_dodge(width = 0.6)) +
  facet_wrap(~condition, nrow = 1) +
  scale_color_manual(values = family_colors, name = "Family",
                     guide = guide_legend(nrow = 1, order = 1)) +
  scale_shape_manual(values = c("Pair-held-out" = 19, "Word-held-out" = 17), name = NULL,
                     guide = guide_legend(nrow = 1, order = 2)) +
  scale_x_discrete(labels = size_labels) +
  scale_y_continuous(breaks = c(0.25, 0.5, 0.75)) +
  labs(y = expression(R^2), x = NULL) +
  theme_paper(base_size = 9) +
  theme(panel.spacing  = unit(0.8, "lines"),
        legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 7))
```

```{r}
#| label: load-brms-all
#| include: false

all_coefs <- brms_summary |>
  filter(is.na(step), label %in% names(family_map)) |>
  left_join(model_meta, by = "label") |>
  mutate(family = family_map[label])
```

### Memorization vs. Abstraction

Full numerical results are in Appendix @tbl-beta and visualized in @fig-beta. $\hat{\beta}_2$ is positive for all 16 models under both attention conditions, with 95% credible intervals excluding zero in all cases except BabyLM-1.3B under the attention-zeroed condition ($\hat{\beta}_2 = 0.008$; 75% of posterior above zero). The U-shaped frequency-residual relationship is therefore a robust finding across model families, training corpora, and parameter scales. Two patterns stand out. First, BabyLM models show larger $\hat{\beta}_2$ magnitudes under full context (0.064–0.073) than any Pile-trained family (0.007–0.018), despite far less training — an effect likely driven by BabyLM's extreme frequency distribution (70% singleton pairs), which creates a sharper contrast between low- and high-frequency items in the regression. Second, under the attention-zeroed condition, $\hat{\beta}_2$ is consistently attenuated relative to full context across all families, confirming that cross-word attention contributes to the high-frequency memorization effect regardless of model family or scale.

<!--# need to back up the statement that $\hat{\beta}_2 is consistently attentuated under attention-zeroed condition with a statistic -->

It is worth noting that the frequency-residual regression results do not reflect an underlying correlation between preference strength and frequency: all $|r| < 0.03$ between $y_{true}$ and frequency. @fig-scatter (Appendix) corroborates this visually: mid-frequency pairs cluster tightly around the prediction diagonal, while low- and high-frequency pairs scatter in opposite directions — precisely the pattern implied by positive $\hat{\beta}_2$.

<!--# Need to check if this is the correlation betwee ytrue and frequency or |ytrue| and frequency -- because frequency should lead to *stronger* preferences, not necessarily more positive -->

```{r}
#| label: fig-beta
#| fig-cap: "Quadratic curvature ($\\hat{\\beta}_2$) of the frequency-residual relationship at the final hidden layer, binomial-level by attention condition. Circles = full context; triangles = attention-zeroed. Positive values indicate that probe error is U-shaped: lowest at intermediate corpus frequencies, highest at extremes. Points = posterior mean; bars = 95\\% credible intervals (minimum bar enforced for visibility)."
#| fig-width: 3.3
#| fig-height: 4.8
#| eval: !expr nrow(all_coefs) > 0

all_coefs |>
  filter(mode == "mean_pooled") |>
  mutate(
    condition = recode_condition(condition),
    family    = factor(family, levels = c("BabyLM", "GPT-2", "Pythia", "OLMo", "Llama")),
    model_fct = factor(label, levels = dumbbell_order),
    b_lo      = pmin(b_quad_sq_lo, b_quad_sq - 0.006),
    b_hi      = pmax(b_quad_sq_hi, b_quad_sq + 0.006)
  ) |>
  ggplot(aes(y = model_fct, x = b_quad_sq, color = family, shape = condition)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.4) +
  geom_linerange(aes(xmin = b_lo, xmax = b_hi), linewidth = 0.6,
                 position = position_dodge(width = 0.6)) +
  geom_point(size = 2.0, position = position_dodge(width = 0.6)) +
  facet_grid(family ~ ., scales = "free_y", space = "free_y") +
  scale_color_manual(values = family_colors, guide = "none") +
  scale_shape_manual(values = c("Full context" = 19, "Attention-zeroed" = 17), name = NULL) +
  scale_x_continuous(breaks = c(0, 0.05, 0.10),
                     labels = c("0", "0.05", "0.10"),
                     expand = expansion(mult = c(0.05, 0.12))) +
  labs(x = expression(hat(beta)[2]~"(quadratic curvature)"), y = NULL) +
  theme_paper(base_size = 9) +
  theme(strip.text.y  = element_blank(),
        panel.spacing = unit(0.2, "lines"),
        legend.position = "bottom")
```

## Discussion

The results of Experiment 2 replicate the results of Experiment 1, showing that across several different models we see similar effects of both abstraction-driven preferences as well as memorization.

<!--# Need to flesh this out -->

# Experiment 3: Developmental Trajectories {#sec-exp3}

Experiments 1 and 2 establish that language models encode binomial preferences at the end of training, but they leave open the timecourse in which these representations emerge, and whether the memorization-abstraction tradeoff is present from early on or crystallizes with additional exposure. Experiment 3 addresses this by examining the same probe at multiple training checkpoints. We ask two questions: (1) Does probe accuracy increase monotonically with training, or does it plateau early, suggesting that preference representations are rapidly acquired? (2) Does the U-shaped frequency-residual relationship hold from early checkpoints, or does it only emerge as training progresses and memorization of high-frequency items accumulates?

## Methods

We probe six checkpoints across two different model families (BabyLM and Pythia). The same BabyLMs from the previous Experiments were used for this one(125M, 350M, and 1.3B) and the comparable Pythia models were chosen for rough comparison (160M, 410M, and 1B). Checkpoints were sampled at approximately even intervals across the BabyLM training. For Pythia, however, rather than sampling even intervals across training, we instead opted to choose comparable checkpoints such that the Pythia models had seen roughly the same number of tokens as the BabyLMs at each checkpoint. These checkpoints span roughly 20M to 2B effective training tokens for each model.

The probe, input modes, attention conditions, and cross-validation splits are identical to Experiments 1 and 2. For the frequency-residual analysis, we fit the quadratic Bayesian regression described in Experiment 1 at each checkpoint and report $\hat\beta_2$ (the curvature of the U-shape) as a function of training step.

<!--# Maybe should go back to reporting the vertex? -->

## Results

```{r}
#| label: load-r2-checkpoints
#| cache: false
#| include: false

load_ckpt_r2 <- function(registry) {
  map_dfr(seq_len(nrow(registry)), function(i) {
    slug  <- registry$slug[i]
    label <- registry$label[i]
    step  <- registry$step[i]
    tok   <- registry$tokens_M[i]
    bind_rows(
      load_mlp(slug, label, FALSE),
      load_mlp(slug, label, TRUE)
    ) |>
      mutate(step = step, tokens_M = tok)
  }) |>
    filter(split %in% c("pair_novel", "word_novel")) |>
    group_by(model, condition, mode, split, is_control, step) |>
    filter(layer == max(layer)) |>
    ungroup()
}

babylm_ckpt_r2  <- load_ckpt_r2(babylm_ckpt_registry)
pythia_ckpt_r2  <- load_ckpt_r2(pythia_ckpt_registry)

ckpt_r2 <- bind_rows(
  babylm_ckpt_r2 |> mutate(family_lbl = "BabyLM"),
  pythia_ckpt_r2 |> mutate(family_lbl = "Pythia")
) |> mutate(family_lbl = factor(family_lbl, levels = c("BabyLM", "Pythia")))
```

```{r}
#| label: load-brms-checkpoints
#| include: false

babylm_ckpt_coefs <- brms_summary |>
  filter(label %in% c("BabyLM-125M", "BabyLM-350M", "BabyLM-1.3B"), !is.na(step)) |>
  left_join(babylm_ckpt_registry |> select(label, step, tokens_M),
            by = c("label", "step"))

pythia_ckpt_coefs <- brms_summary |>
  filter(label %in% c("Pythia-160M", "Pythia-410M", "Pythia-1B"),
         !is.na(step)) |>
  left_join(pythia_ckpt_registry |> select(label, step, tokens_M),
            by = c("label", "step"))

ckpt_coefs <- bind_rows(
  babylm_ckpt_coefs |> mutate(family_lbl = "BabyLM"),
  pythia_ckpt_coefs |> mutate(family_lbl = "Pythia")
) |> mutate(family_lbl = factor(family_lbl, levels = c("BabyLM", "Pythia")))

ckpt_colors <- c(
  "BabyLM-125M" = "#4E9D69",
  "BabyLM-350M" = "#007030",
  "BabyLM-1.3B" = "#154733",
  "Pythia-160M"  = "#F0B323",
  "Pythia-410M"  = "#B8860B",
  "Pythia-1B"    = "#875E00"
)
```

BabyLM pair-held-out R² is already high at the earliest checkpoint (\~20M tokens: 0.67–0.72) and broadly declines to 0.32–0.38 by the end of training. At the earliest checkpoint, the attention-zeroed condition outperforms full context (0.78–0.80 vs. 0.67–0.72); this ordering reverses between \~120M and \~750M tokens, so that full context exceeds attention-zeroed at the final checkpoint for all three model sizes. Pythia shows an inverted-U trajectory instead: R² rises from 0.68–0.74 at 33.6M tokens to a peak of 0.80–0.86 at 537M tokens, then declines to 0.71–0.76 at 1074M–2097M tokens — consistent with Pile-trained models still consolidating preference representations during this early window (the first 0.7% of 300B tokens). The attention condition flip for Pythia follows the same qualitative pattern as BabyLM. Full numerical results are in Appendix [-@sec-appendix-exp3]; word-level results are in Appendix [-@sec-appendix-exp3].

For $\hat{\beta}_2$, BabyLM starts negative at all sizes (-0.013 to -0.029), remains negative until approximately 750M tokens, and reaches 0.045–0.093 at the final checkpoint, matching the positive curvature from Experiment 1. Pythia's $\hat{\beta}_2$ is positive from its earliest checkpoint (33.6M tokens), with credible intervals confidently excluding zero by 537M tokens. As discussed in Appendix [-@sec-appendix-corpus-freq], this contrast reflects the far greater absolute token exposure per pair in the Pile relative to BabyLM. Full numerical results are in Appendix [-@sec-appendix-exp3].

```{r}
#| label: fig-checkpoints
#| fig-cap: "Probe R² (top) and quadratic frequency-residual curvature ($\\hat{\\beta}_2$, bottom) across BabyLM and Pythia training checkpoints, binomial-level by attention condition. BabyLM spans ~20M–1.9B effective training tokens; Pythia spans ~34M–2.1B tokens (~0.7\\% of the full 300B training run). Ribbons show 95\\% credible intervals. $\\hat{\\beta}_2$ y-axes are free per family row (BabyLM and Pythia have very different scales). Control R² $\\approx$ 0 (Appendix)."
#| fig-width: 3.3
#| fig-height: 4.5
#| eval: !expr nrow(ckpt_r2) > 0

p_r2 <- ckpt_r2 |>
  filter(mode == "mean_pooled", !is_control) |>
  mutate(
    model     = factor(model, levels = model_order),
    condition = recode_condition(condition),
    panel     = factor(recode(condition,
                              "Full context"     = "Full context",
                              "Attention-zeroed" = "Attention-zeroed"),
                       levels = c("Full context", "Attention-zeroed")),
    split_lbl = factor(recode(split,
                              "pair_novel" = "Pair-held-out",
                              "word_novel" = "Word-held-out"),
                       levels = c("Pair-held-out", "Word-held-out"))
  ) |>
  ggplot(aes(x = tokens_M, y = mean_r2, color = model,
             linetype = split_lbl, group = interaction(model, split_lbl))) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 0.7) +
  facet_grid(family_lbl ~ panel, scales = "free_x") +
  scale_x_log10(breaks = c(100, 1000),
               labels = label_number(suffix = "M", accuracy = 1, big.mark = ""),
               expand = expansion(mult = c(0.05, 0.1))) +
  scale_color_manual(values = ckpt_colors, name = "Model") +
  scale_linetype_manual(values = c("Pair-held-out" = "solid", "Word-held-out" = "dashed"),
                        guide = "none") +
  labs(x = NULL, y = expression(R^2)) +
  theme_paper(base_size = 9) +
  theme(legend.position = "none")

p_beta <- ckpt_coefs |>
  filter(mode == "mean_pooled") |>
  mutate(
    model     = factor(label, levels = model_order),
    condition = recode_condition(condition),
    panel     = factor(recode(condition,
                              "Full context"     = "Full context",
                              "Attention-zeroed" = "Attention-zeroed"),
                       levels = c("Full context", "Attention-zeroed"))
  ) |>
  ggplot(aes(x = tokens_M, y = beta2_mean, color = model, fill = model, group = model)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.3) +
  geom_ribbon(aes(ymin = beta2_ci025, ymax = beta2_ci975), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 0.7) +
  facet_grid(family_lbl ~ panel, scales = "free") +
  scale_x_log10(breaks = c(100, 1000),
               labels = label_number(suffix = "M", accuracy = 1, big.mark = ""),
               expand = expansion(mult = c(0.05, 0.1))) +
  scale_y_continuous(n.breaks = 3) +
  scale_color_manual(values = ckpt_colors, name = "Model") +
  scale_fill_manual(values  = ckpt_colors, guide = "none") +
  labs(x = "Tokens (M, log scale)", y = expression(hat(beta)[2])) +
  theme_paper(base_size = 9) +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 3, title = "Model"))

p_r2 / p_beta + plot_layout(heights = c(1, 1.4))
```

## Discussion

# Conclusion {#sec-discussion}

All 16 models encode decodable binomial ordering preferences in their final-layer representations, with probe R² well above near-zero control values in every condition. BabyLM models produce lower R² (0.24–0.34) than their Pile-trained counterparts (0.43–0.79), consistent with their smaller training corpora, and scale effects within families are non-monotonic. Across Pile-trained families, attention-zeroed outperforms full-context at the binomial level — indicating that cross-token attention in data-rich models encodes pair-specific, idiosyncratic representations that reduce probe generalizability — while BabyLM shows the reverse, consistent with attention computing abstract compositional features rather than retrieving memorized pair orderings. Positive quadratic curvature of probe error over log-frequency ($\hat{\beta}_2 > 0$) holds across all models and conditions, revealing a mid-frequency advantage in which pairs at intermediate frequencies are easiest to decode, with both rare pairs (insufficient signal) and highly frequent pairs (idiosyncratic memorization) producing elevated prediction error.

```{=latex}
\bibliography{references}
```

\clearpage
\onecolumn
```{=latex}
\appendix
```

\clearpage

# Model Details {#sec-appendix-models}

Table \ref{tbl-models} lists all 16 models used across experiments, organized by family, with approximate parameter counts and training token counts.

```{r}
#| label: tbl-models
#| tbl-cap: "Models used across experiments. Training token counts for full-scale models are approximate; GPT-2 counts are estimated from the reported 40GB WebText corpus. BabyLM models train for 20 epochs over 150M tokens (3B effective token exposures)."

model_table <- tribble(
  ~Family,  ~Model,          ~Parameters, ~`Tokens`,
  "BabyLM", "BabyLM-125M",  "125M",      "150M",
  "BabyLM", "BabyLM-350M",  "350M",      "150M",
  "BabyLM", "BabyLM-1.3B",  "1.3B",      "150M",
  "GPT-2",  "GPT-2",        "117M",      "~8B",
  "GPT-2",  "GPT-2-medium", "345M",      "~8B",
  "GPT-2",  "GPT-2-large",  "762M",      "~8B",
  "GPT-2",  "GPT-2-XL",     "1.5B",      "~8B",
  "Pythia", "Pythia-160M",  "160M",      "~300B",
  "Pythia", "Pythia-410M",  "410M",      "~300B",
  "Pythia", "Pythia-1B",    "1B",        "~300B",
  "OLMo",   "OLMo-1B",      "1B",        "3T",
  "OLMo",   "OLMo-7B",      "7B",        "2.5T",
  "OLMo",   "OLMo-2-1B",    "1B",        "~4T",
  "OLMo",   "OLMo-2-7B",    "7B",        "~4T",
  "Llama",  "Llama-1.3B",   "1.2B",      "~9T",
  "Llama",  "Llama-3-8B",   "8B",        "~15T"
)

model_table |>
  kbl(booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 9, latex_options = "HOLD_position") |>
  collapse_rows(columns = 1, latex_hline = "major")
```

\clearpage

# Experiment 1: Supplementary Results {#sec-appendix-exp1}

This appendix provides supplementary materials for Experiment 1 (§\ref{sec-exp1}), which examined final-layer probe performance and frequency-residual curvature for the three BabyLM models. Subsections report full R² tables, control probe results, a word-level replication, and the frequency-residual regression coefficients.

## Full Probe R² {#sec-appendix-full-r2-babylm}

Probe R² at the final hidden layer for BabyLM models, at the binomial level by attention condition. P = pair-held-out; W = word-held-out. SD across 10 folds in parentheses.

```{r}
#| label: tbl-r2-babylm
#| tbl-cap: "Probe R² at the final hidden layer for BabyLM models, at the binomial level by attention condition. P = pair-held-out split; W = word-held-out split. SD across 10 folds in parentheses."
#| eval: !expr nrow(all_r2) > 0

all_r2 |>
  filter(family == "BabyLM", mode == "mean_pooled") |>
  select(model, condition, split, mean_r2, sd_r2, is_control) |>
  pivot_wider(names_from  = is_control,
              values_from = c(mean_r2, sd_r2),
              names_sep   = "_") |>
  rename(r2_main = mean_r2_FALSE, sd_main = sd_r2_FALSE,
         r2_ctrl = mean_r2_TRUE,  sd_ctrl = sd_r2_TRUE) |>
  mutate(
    condition = recode_condition(condition),
    model     = factor(model, levels = model_order),
    Split     = if_else(split == "pair_novel", "P", "W"),
    `R²`      = sprintf("%.3f (%.3f)", r2_main, sd_main),
    Ctrl      = sprintf("%.3f (%.3f)", r2_ctrl,  sd_ctrl),
  ) |>
  arrange(model, condition, split) |>
  select(Model = model, Condition = condition, Split, `R²`, Ctrl) |>
  kbl(booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 8, latex_options = "HOLD_position") |>
  collapse_rows(columns = 1:2, latex_hline = "major")
```

## Control Probe R² {#sec-appendix-ctrl-exp1}

Control probe R² for BabyLM models at the final hidden layer, at the binomial level by attention condition. Labels are randomly permuted; values near zero confirm the main probe recovers genuine linguistic structure rather than label artifacts or probe capacity.

```{r}
#| label: tbl-ctrl-r2-babylm
#| tbl-cap: "Control probe R² for BabyLM models (final layer), at the binomial level by attention condition. SD across 10 folds in parentheses. P = pair-held-out; W = word-held-out."

all_r2 |>
  filter(family == "BabyLM", mode == "mean_pooled", is_control) |>
  mutate(
    condition    = recode(condition, !!!condition_labels),
    model        = factor(model, levels = model_order),
    Split        = if_else(split == "pair_novel", "P", "W"),
    `Control R²` = sprintf("%.4f (%.4f)", mean_r2, sd_r2)
  ) |>
  arrange(model, condition, split) |>
  select(Model = model, Condition = condition, Split, `Control R²`) |>
  kbl(booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 8, latex_options = "HOLD_position") |>
  collapse_rows(columns = 1:2, latex_hline = "major")
```

## Word-Level Replication {#sec-appendix-words-only-exp1}

We trained the MLP probe using word-level representations (`words_only`) for the three BabyLM models to verify that the preference signal is not an artifact of mean pooling across the full binomial phrase. Under this condition the probe receives a concatenation of two word vectors extracted from isolated sentence contexts ("*the bread*", "*the butter*"), so it cannot rely on any co-occurrence signal from the two words' shared sentence context.

Pair-held-out R² ranges from 0.31 to 0.33 — within a few percentage points of the corresponding binomial-level values (0.29–0.34). The generalization pattern is preserved: word-held-out R² is consistently lower than pair-held-out R² for all three models (word-held-out 0.18–0.20; pair-minus-word gap 0.127–0.129), confirming that the lexical generalization cost seen at the binomial level is a property of the preference encoding itself rather than an artifact of the mean-pooled representation. $\hat{\beta}_2$ is positive and credible intervals exclude zero for all three models, matching the curvature found at the binomial level (words-only $\hat{\beta}_2$: 0.039–0.079; full-context binomial-level: 0.064–0.073). Full results are in @tbl-r2-babylm-words-only.

```{r}
#| label: tbl-r2-babylm-words-only
#| tbl-cap: "Word-level probe R² and selectivity (main $-$ control R²) for BabyLM models (final layer). P = pair-held-out split; W = word-held-out split. SD across 10 folds in parentheses."

all_r2 |>
  filter(family == "BabyLM", mode == "words_only", condition == "default") |>
  select(model, condition, mode, split, mean_r2, sd_r2, is_control) |>
  pivot_wider(names_from  = is_control,
              values_from = c(mean_r2, sd_r2),
              names_sep   = "_") |>
  rename(r2_main = mean_r2_FALSE, sd_main = sd_r2_FALSE,
         r2_ctrl = mean_r2_TRUE,  sd_ctrl = sd_r2_TRUE) |>
  mutate(
    selectivity = r2_main - r2_ctrl,
    mode        = recode(mode, !!!mode_labels),
    model       = factor(model, levels = model_order),
    Split       = if_else(split == "pair_novel", "P", "W"),
    `R²`        = sprintf("%.3f (%.3f)", r2_main, sd_main),
    Ctrl        = sprintf("%.3f (%.3f)", r2_ctrl,  sd_ctrl),
    Sel.        = sprintf("%.3f", selectivity)
  ) |>
  arrange(model, split) |>
  select(Model = model, Mode = mode, Split, `R²`, Ctrl, Sel.) |>
  kbl(booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 8, latex_options = "HOLD_position") |>
  collapse_rows(columns = 1:2, latex_hline = "major")
```

## Frequency-Residual Regression: BabyLM Models {#sec-appendix-beta-babylm}

Bayesian frequency-residual regression results for BabyLM models (final layer), at the binomial level by attention condition. 95% CIs in brackets. Vertex = $-\hat\beta_1 / (2\hat\beta_2)$, the log-frequency at which probe error is minimized.

```{r}
#| label: tbl-beta-babylm
#| tbl-cap: "Bayesian frequency-residual regression results for BabyLM models (final layer), at the binomial level by attention condition. 95\\% CIs in brackets. Vertex = $-\\hat\\beta_1 / (2\\hat\\beta_2)$, the log-frequency at which probe error is minimized."
#| eval: !expr nrow(babylm_coefs) > 0

babylm_coefs |>
  filter(mode == "mean_pooled") |>
  mutate(
    model     = factor(label, levels = model_order),
    condition = recode_condition(condition),
    `$\\hat{\\beta}_1$` = sprintf("%.4f [%.4f, %.4f]",
                        b_quad_linear, b_quad_linear_lo, b_quad_linear_hi),
    `$\\hat{\\beta}_2$` = sprintf("%.5f [%.5f, %.5f]",
                        b_quad_sq, b_quad_sq_lo, b_quad_sq_hi),
    Vertex    = sprintf("%.2f", vertex)
  ) |>
  arrange(model, condition) |>
  select(Model = model, Condition = condition,
         `$\\hat{\\beta}_1$`, `$\\hat{\\beta}_2$`, Vertex) |>
  kbl(booktabs = TRUE, linesep = "", escape = FALSE) |>
  kable_styling(font_size = 9, latex_options = "HOLD_position") |>
  collapse_rows(columns = 1, latex_hline = "major")
```

\clearpage

# Experiment 2: Supplementary Results {#sec-appendix-exp2}

This appendix provides supplementary materials for Experiment 2 (§\ref{sec-exp2}), which examined final-layer representations across all 16 models from five families. Subsections report full R² tables, control probe results, a word-level replication, frequency-residual regression coefficients for all models, and a predicted-vs.-observed scatter plot.

## Full Probe R² {#sec-appendix-full-r2}

Probe R² for all 16 models at the final hidden layer, binomial-level input mode, by attention condition and cross-validation split. P = pair-held-out; W = word-held-out. SD across 10 folds in parentheses.

```{r}
#| label: tbl-r2
#| tbl-cap: "Probe R² at the final hidden layer for all 16 models, at the binomial level by attention condition. P = pair-held-out split; W = word-held-out split. SD across 10 folds in parentheses."
#| eval: !expr nrow(all_r2) > 0

all_r2 |>
  filter(mode == "mean_pooled") |>
  select(model, family, condition, mode, split, mean_r2, sd_r2, is_control) |>
  pivot_wider(names_from  = is_control,
              values_from = c(mean_r2, sd_r2),
              names_sep   = "_") |>
  rename(r2_main = mean_r2_FALSE, sd_main = sd_r2_FALSE,
         r2_ctrl = mean_r2_TRUE,  sd_ctrl = sd_r2_TRUE) |>
  mutate(
    condition = recode_condition(condition),
    model     = factor(model, levels = model_order),
    Split     = if_else(split == "pair_novel", "P", "W"),
    `R²`      = sprintf("%.3f (%.3f)", r2_main, sd_main),
    Ctrl      = sprintf("%.3f (%.3f)", r2_ctrl,  sd_ctrl),
  ) |>
  arrange(model, condition, split) |>
  select(Model = model, Condition = condition, Split, `R²`, Ctrl) |>
  kbl(booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 7, latex_options = c("scale_down", "HOLD_position")) |>
  collapse_rows(columns = 1:2, latex_hline = "major")
```

## Control Probe R² {#sec-appendix-ctrl-exp2}

Control probe R² for all 16 models at the final hidden layer, at the binomial level by attention condition.

```{r}
#| label: tbl-ctrl-r2
#| tbl-cap: "Control probe R² for all 16 models (final layer), at the binomial level by attention condition. SD across 10 folds in parentheses. P = pair-held-out; W = word-held-out."

all_r2 |>
  filter(mode == "mean_pooled", is_control) |>
  mutate(
    condition    = recode(condition, !!!condition_labels),
    model        = factor(model, levels = model_order),
    Split        = if_else(split == "pair_novel", "P", "W"),
    `Control R²` = sprintf("%.4f (%.4f)", mean_r2, sd_r2)
  ) |>
  arrange(family, model, condition, split) |>
  select(Family = family, Model = model, Condition = condition, Split, `Control R²`) |>
  kbl(booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 8, latex_options = "HOLD_position") |>
  collapse_rows(columns = 1:3, latex_hline = "major")
```

## Word-Level Replication {#sec-appendix-words-only-exp2}

Across all 16 models, word-level probing reproduces the main findings from Experiment 2. Pair-held-out R² is slightly lower than the corresponding binomial-level values — by roughly 5–10 percentage points for most Pile-trained models — but rank order across families and the generalization pattern (pair-held-out \> word-held-out) are preserved in every case. For GPT-2, pair-held-out R² under word-level probing ranges from 0.46 to 0.62 (vs. 0.49–0.65 at the binomial level); for Pythia, 0.51–0.57 (vs. 0.53–0.59); for OLMo, 0.52–0.57 (vs. 0.56–0.62); for Llama, 0.36–0.45 (vs. 0.43–0.55). The generalization gaps (pair-held-out minus word-held-out) range from 0.044 to 0.073 across Pile-trained models, closely matching the corresponding binomial-level gaps (0.035–0.080). $\hat{\beta}_2$ is positive for all 16 models, with magnitudes similar to the full-context binomial-level condition. Full results are in @tbl-r2-words-only.

```{r}
#| label: tbl-r2-words-only
#| tbl-cap: "Word-level probe R² and selectivity (main $-$ control R²) for all 16 models (final layer). P = pair-held-out split; W = word-held-out split. SD across 10 folds in parentheses."

all_r2 |>
  filter(mode == "words_only", condition == "default") |>
  select(family, model, condition, mode, split, mean_r2, sd_r2, is_control) |>
  pivot_wider(names_from  = is_control,
              values_from = c(mean_r2, sd_r2),
              names_sep   = "_") |>
  rename(r2_main = mean_r2_FALSE, sd_main = sd_r2_FALSE,
         r2_ctrl = mean_r2_TRUE,  sd_ctrl = sd_r2_TRUE) |>
  mutate(
    selectivity = r2_main - r2_ctrl,
    mode        = recode(mode, !!!mode_labels),
    model       = factor(model, levels = model_order),
    Split       = if_else(split == "pair_novel", "P", "W"),
    `R²`        = sprintf("%.3f (%.3f)", r2_main, sd_main),
    Ctrl        = sprintf("%.3f (%.3f)", r2_ctrl,  sd_ctrl),
    Sel.        = sprintf("%.3f", selectivity)
  ) |>
  arrange(family, model, split) |>
  select(Family = family, Model = model, Mode = mode, Split, `R²`, Ctrl, Sel.) |>
  kbl(booktabs = TRUE, linesep = "") |>
  kable_styling(font_size = 8, latex_options = "HOLD_position") |>
  collapse_rows(columns = 1:3, latex_hline = "major")
```

## Frequency-Residual Regression: All Models {#sec-appendix-full-beta}

Bayesian frequency-residual regression results for all 16 final-layer models, binomial-level input mode. 95% CIs in brackets. Vertex = $-\hat\beta_1 / (2\hat\beta_2)$, the log-frequency at which probe error is minimized.

```{r}
#| label: tbl-beta
#| tbl-cap: "Bayesian frequency-residual regression results for all 16 final-layer models, at the binomial level by attention condition. 95\\% CIs in brackets. Vertex = $-\\hat\\beta_1 / (2\\hat\\beta_2)$, the log-frequency at minimum probe error."
#| eval: !expr nrow(all_coefs) > 0

all_coefs |>
  filter(mode == "mean_pooled") |>
  mutate(
    model     = factor(label, levels = model_order),
    condition = recode_condition(condition),
    `$\\hat{\\beta}_1$` = sprintf("%.4f [%.4f, %.4f]",
                        b_quad_linear, b_quad_linear_lo, b_quad_linear_hi),
    `$\\hat{\\beta}_2$` = sprintf("%.5f [%.5f, %.5f]",
                        b_quad_sq, b_quad_sq_lo, b_quad_sq_hi),
    Vertex    = sprintf("%.2f", vertex)
  ) |>
  arrange(model, condition) |>
  select(Model = model, Condition = condition,
         `$\\hat{\\beta}_1$`, `$\\hat{\\beta}_2$`, Vertex) |>
  kbl(booktabs = TRUE, linesep = "", escape = FALSE) |>
  kable_styling(font_size = 7, latex_options = c("scale_down", "HOLD_position")) |>
  collapse_rows(columns = 1, latex_hline = "major")
```

## Predicted vs. Observed {#sec-appendix-scatter}

Predicted vs. observed binomial preference for all 16 models at the final hidden layer, full-context condition, binomial-level input mode. Each point is a corpus binomial pair colored by log-frequency tertile (low/medium/high). Perfect prediction falls on the diagonal. Models are ordered by family (BabyLM, GPT-2, Pythia, OLMo, Llama) and then by size within family.

```{r}
#| label: fig-scatter
#| fig-cap: "Predicted vs. observed binomial preference for all 16 models at the final hidden layer, full-context condition, binomial-level input mode. Each point is a corpus binomial pair colored by log-frequency tertile (low/medium/high). Perfect prediction falls on the diagonal."
#| fig-width: 6.5
#| fig-height: 9.0

scatter_slugs <- names(all_slugs)
scatter_labels <- unname(all_slugs)

freq_pile <- read_csv(
  file.path(RES, "corpus_binomials_infinigram_piletrain.csv"),
  show_col_types = FALSE
) |> select(word1, word2, freq_total)

freq_babylm <- read_csv(
  "../Data/corpus_binomials.csv",
  show_col_types = FALSE
) |>
  mutate(freq_total = freq_w1_w2 + freq_w2_w1) |>
  select(word1, word2, freq_total)

load_scatter <- function(slug, label) {
  xz  <- file.path(RES, slug, "by_layer_corpus_pred.csv.xz")
  csv <- file.path(RES, slug, "by_layer_corpus_pred.csv")
  if (!file.exists(xz) && !file.exists(csv)) return(NULL)
  pred <- if (file.exists(xz)) {
    con <- xzfile(xz, "rb"); d <- read_csv(con, show_col_types = FALSE); close(con); d
  } else {
    read_csv(csv, show_col_types = FALSE)
  }
  last_layer <- as.character(max(suppressWarnings(as.integer(unique(pred$layer))), na.rm = TRUE))
  freq_df <- if (grepl("babylm", slug)) freq_babylm else freq_pile
  pred |>
    filter(as.character(layer) == last_layer,
           condition == "default", mode == "mean_pooled") |>
    left_join(freq_df, by = c("word1", "word2")) |>
    filter(!is.na(freq_total), freq_total > 0) |>
    mutate(
      label    = label,
      log_freq = log(freq_total),
      freq_bin = cut(log_freq, breaks = 3,
                     labels = c("Low frequency", "Medium frequency", "High frequency"))
    )
}

scatter_data <- map_dfr(seq_along(scatter_slugs), function(i) {
  load_scatter(scatter_slugs[i], scatter_labels[i])
})

if (nrow(scatter_data) > 0) {
  scatter_data |>
    mutate(label = factor(label, levels = scatter_labels)) |>
    ggplot(aes(x = y_true, y = y_pred, color = freq_bin)) +
    geom_point(size = 0.2, alpha = 0.12) +
    geom_abline(slope = 1, intercept = 0,
                linetype = "dashed", color = "gray40", linewidth = 0.4) +
    facet_wrap(~label, nrow = 4) +
    scale_color_manual(
      values = c(
        "Low frequency"    = "#4477AA",
        "Medium frequency" = "#CCBB44",
        "High frequency"   = "#EE6677"
      )
    ) +
    labs(x = "Observed preference (log-odds)",
         y = "Predicted preference",
         color = "Corpus frequency") +
    theme_paper() +
    theme(legend.position = "bottom",
          axis.text = element_text(size = 6),
          strip.text = element_text(size = 7))
}
```

\clearpage

# Experiment 3: Supplementary Results {#sec-appendix-exp3}

This appendix provides supplementary materials for Experiment 3 (§\ref{sec-exp3}), which tracked probe R² and frequency-residual curvature across training checkpoints for BabyLM and Pythia. Subsections report full numerical checkpoint tables and a word-level replication.

## Checkpoint Probe Tables {#sec-appendix-ckpt-tables}

Full numerical results for the Experiment 3 checkpoint analysis. Tables report probe R² (pair-held-out and word-held-out) and frequency-residual regression coefficients ($\hat{\beta}_2$) for BabyLM and Pythia across training checkpoints, for the binomial-level (mean-pooled) representation under full-context and attention-zeroed conditions.

```{r}
#| echo: false

make_ckpt_r2_tbl <- function(df, cond_val, mode_val) {
  df |>
    filter(condition == cond_val, mode == mode_val,
           split %in% c("pair_novel", "word_novel")) |>
    select(model, step, tokens_M, split, mean_r2, sd_r2, is_control) |>
    pivot_wider(names_from = is_control, values_from = c(mean_r2, sd_r2),
                names_sep = "_") |>
    rename(r2_main = mean_r2_FALSE, sd_main = sd_r2_FALSE,
           r2_ctrl = mean_r2_TRUE,  sd_ctrl = sd_r2_TRUE) |>
    mutate(
      model  = factor(model, levels = model_order),
      Split  = if_else(split == "pair_novel", "P", "W"),
      Tokens = sprintf("%.0f M", tokens_M),
      `R²`   = sprintf("%.3f (%.3f)", r2_main, sd_main),
      Ctrl   = sprintf("%.3f (%.3f)", r2_ctrl,  sd_ctrl),
      Sel.   = sprintf("%.3f", r2_main - r2_ctrl)
    ) |>
    arrange(model, step, split) |>
    select(Model = model, Step = step, Tokens, Split, `R²`, Ctrl, Sel.) |>
    kbl(booktabs = TRUE, linesep = "") |>
    kable_styling(font_size = 8, latex_options = "HOLD_position") |>
    collapse_rows(columns = 1:2, latex_hline = "major")
}

make_ckpt_beta_tbl <- function(df, cond_val, mode_val) {
  df |>
    filter(condition == cond_val, mode == mode_val) |>
    mutate(
      model = factor(label, levels = model_order),
      `$\\hat{\\beta}_2$` = sprintf("%.4f", beta2_mean),
      SE      = sprintf("%.4f", beta2_se),
      CI      = sprintf("[%.4f, %.4f]", beta2_ci025, beta2_ci975),
      `P(>0)` = sprintf("%.1f\\%%", beta2_pct_gt0)
    ) |>
    arrange(model, tokens_M) |>
    mutate(Tokens = sprintf("%.0f M", tokens_M)) |>
    select(Model = model, Step = step, Tokens,
           `$\\hat{\\beta}_2$`, SE, CI, `P(>0)`) |>
    kbl(booktabs = TRUE, linesep = "", escape = FALSE) |>
    kable_styling(font_size = 8, latex_options = "HOLD_position") |>
    collapse_rows(columns = 1, latex_hline = "major")
}
```

```{r}
#| label: tbl-r2-checkpoints-babylm-default-mean
#| tbl-cap: "BabyLM checkpoint probe R² — full context, binomial-level. P = pair-held-out; W = word-held-out. SD in parentheses."

if (nrow(babylm_ckpt_r2) > 0)
  make_ckpt_r2_tbl(babylm_ckpt_r2, "default", "mean_pooled")
```

```{r}
#| label: tbl-r2-checkpoints-babylm-attn-mean
#| tbl-cap: "BabyLM checkpoint probe R² — attention-zeroed, binomial-level."

if (nrow(babylm_ckpt_r2) > 0)
  make_ckpt_r2_tbl(babylm_ckpt_r2, "attn_zeroed", "mean_pooled")
```

```{r}
#| label: tbl-r2-checkpoints-pythia-default-mean
#| tbl-cap: "Pythia checkpoint probe R² — full context, binomial-level."

if (nrow(pythia_ckpt_r2) > 0)
  make_ckpt_r2_tbl(pythia_ckpt_r2, "default", "mean_pooled")
```

```{r}
#| label: tbl-r2-checkpoints-pythia-attn-mean
#| tbl-cap: "Pythia checkpoint probe R² — attention-zeroed, binomial-level."

if (nrow(pythia_ckpt_r2) > 0)
  make_ckpt_r2_tbl(pythia_ckpt_r2, "attn_zeroed", "mean_pooled")
```

```{r}
#| label: tbl-beta-checkpoints-babylm-default-mean
#| tbl-cap: "BabyLM checkpoint frequency-residual regression ($\\hat{\\beta}_2$) — full context, binomial-level. SE = posterior SD; CI = 95\\% credible interval; P(>0) = percentage of posterior samples with $\\hat{\\beta}_2 > 0$."

if (nrow(babylm_ckpt_coefs) > 0)
  make_ckpt_beta_tbl(babylm_ckpt_coefs, "default", "mean_pooled")
```

```{r}
#| label: tbl-beta-checkpoints-babylm-attn-mean
#| tbl-cap: "BabyLM checkpoint frequency-residual regression ($\\hat{\\beta}_2$) — attention-zeroed, binomial-level."

if (nrow(babylm_ckpt_coefs) > 0)
  make_ckpt_beta_tbl(babylm_ckpt_coefs, "attn_zeroed", "mean_pooled")
```

```{r}
#| label: tbl-beta-checkpoints-pythia-default-mean
#| tbl-cap: "Pythia checkpoint frequency-residual regression ($\\hat{\\beta}_2$) — full context, binomial-level."

if (nrow(pythia_ckpt_coefs) > 0)
  make_ckpt_beta_tbl(pythia_ckpt_coefs, "default", "mean_pooled")
```

```{r}
#| label: tbl-beta-checkpoints-pythia-attn-mean
#| tbl-cap: "Pythia checkpoint frequency-residual regression ($\\hat{\\beta}_2$) — attention-zeroed, binomial-level."

if (nrow(pythia_ckpt_coefs) > 0)
  make_ckpt_beta_tbl(pythia_ckpt_coefs, "attn_zeroed", "mean_pooled")
```

## Word-Level Replication {#sec-appendix-words-only-exp3}

Word-level probe results across BabyLM and Pythia checkpoints replicate the key patterns from the binomial-level analysis. The generalization gap (pair-held-out minus word-held-out) is largest at the earliest checkpoints and shrinks over the course of training. For BabyLM, the gap at the first checkpoint ranges from 0.15 to 0.36 across model sizes, narrowing to 0.13–0.14 by the final checkpoint — the same qualitative trajectory observed at the binomial level. $\hat{\beta}_2$ shows the same sign trajectory as in the main analysis: negative early in BabyLM training, transitioning to positive after approximately 750M tokens. Full R² results are in @tbl-r2-checkpoints-babylm-default-words and @tbl-r2-checkpoints-pythia-default-words; frequency-residual regression results are in @tbl-beta-checkpoints-babylm-default-words and @tbl-beta-checkpoints-pythia-default-words.

```{r}
#| label: tbl-r2-checkpoints-babylm-default-words
#| tbl-cap: "BabyLM checkpoint probe R² — full context, word-level. P = pair-held-out; W = word-held-out. SD in parentheses."

if (nrow(babylm_ckpt_r2) > 0)
  make_ckpt_r2_tbl(babylm_ckpt_r2, "default", "words_only")
```

```{r}
#| label: tbl-r2-checkpoints-pythia-default-words
#| tbl-cap: "Pythia checkpoint probe R² — full context, word-level. P = pair-held-out; W = word-held-out. SD in parentheses."

if (nrow(pythia_ckpt_r2) > 0)
  make_ckpt_r2_tbl(pythia_ckpt_r2, "default", "words_only")
```

```{r}
#| label: tbl-beta-checkpoints-babylm-default-words
#| tbl-cap: "BabyLM checkpoint frequency-residual regression ($\\hat{\\beta}_2$) — full context, word-level. SE = posterior SD; CI = 95\\% credible interval; P(>0) = percentage of posterior samples with $\\hat{\\beta}_2 > 0$."

if (nrow(babylm_ckpt_coefs) > 0)
  make_ckpt_beta_tbl(babylm_ckpt_coefs, "default", "words_only")
```

```{r}
#| label: tbl-beta-checkpoints-pythia-default-words
#| tbl-cap: "Pythia checkpoint frequency-residual regression ($\\hat{\\beta}_2$) — full context, word-level. SE = posterior SD; CI = 95\\% credible interval; P(>0) = percentage of posterior samples with $\\hat{\\beta}_2 > 0$."

if (nrow(pythia_ckpt_coefs) > 0)
  make_ckpt_beta_tbl(pythia_ckpt_coefs, "default", "words_only")
```

\clearpage

# Corpus Frequency Distributions {#sec-appendix-corpus-freq}

The figure below compares the frequency distributions of the approximately 47,000 corpus binomials in the BabyLM training corpus (150M tokens, 20 epochs) and the Pile, illustrating the stark difference in lexical coverage between the two training regimes. BabyLM's distribution is dominated by singleton pairs (70.7% appear exactly once), while Pile frequencies span several orders of magnitude.

```{r}
#| label: fig-corpus-freq
#| fig-cap: "Frequency distributions for the ~47K corpus binomials in the BabyLM training corpus (150M tokens, 20 epochs) and the Pile. Left: histogram of log$_{10}$ raw counts. Right: rank-frequency plot treating counts as a probability distribution (probability = count / $\\sum$ counts); both axes log-scaled. BabyLM's distribution is heavily concentrated at a single occurrence per pair (70.7% of pairs appear exactly once), whereas Pile frequencies span several orders of magnitude with a median of 365 occurrences per pair."
#| fig-width: 7
#| fig-height: 3.2

corpus_freq_df <- local({
  babylm_raw <- read_csv("../Data/corpus_binomials.csv", show_col_types = FALSE) |>
    mutate(freq = freq_w1_w2 + freq_w2_w1) |>
    filter(freq > 0) |>
    select(word1, word2, freq) |>
    mutate(corpus = "BabyLM (150M tok, 20 ep)")

  pile_raw <- read_csv("../Results/corpus_binomials_infinigram_piletrain.csv",
                       show_col_types = FALSE) |>
    rename(freq = freq_total) |>
    filter(freq > 0) |>
    select(word1, word2, freq) |>
    mutate(corpus = "Pile (~300B tok, 1 ep)")

  bind_rows(babylm_raw, pile_raw) |>
    group_by(corpus) |>
    mutate(prob = freq / sum(freq),
           rank = rank(-freq, ties.method = "first")) |>
    ungroup()
})

corpus_colors <- c("BabyLM (150M tok, 20 ep)" = "#154733",
                   "Pile (~300B tok, 1 ep)"    = "#F0B323")

p_hist <- corpus_freq_df |>
  ggplot(aes(x = log10(freq), fill = corpus, color = corpus)) +
  geom_histogram(aes(y = after_stat(count)),
                 binwidth = 0.25, position = "identity", alpha = 0.55) +
  scale_fill_manual(values  = corpus_colors, name = NULL) +
  scale_color_manual(values = corpus_colors, name = NULL) +
  scale_y_continuous(labels = label_comma()) +
  labs(x = expression(log[10]~"(count)"), y = "Number of pairs") +
  theme_paper() +
  theme(legend.position = "none")

p_rank <- corpus_freq_df |>
  ggplot(aes(x = rank, y = prob, color = corpus)) +
  geom_line(linewidth = 0.7, alpha = 0.85) +
  scale_x_log10(labels = label_comma()) +
  scale_y_log10(labels = label_scientific()) +
  scale_color_manual(values = corpus_colors, name = NULL) +
  labs(x = "Rank (log scale)", y = "Probability (log scale)") +
  theme_paper() +
  theme(legend.position = "bottom")

p_hist + p_rank + plot_layout(ncol = 2, widths = c(1, 1))
```

The BabyLM training corpus is a 150M-token collection of child-directed and informal text (CHILDES, BNC Spoken, children's books, OpenSubtitles, Wikipedia), trained for 20 epochs. The Pile is a \~300B-token web-scraped corpus used to train the Pythia family. For each of the \~47K corpus binomials, we extracted 3-gram counts (i.e., the frequency of "word$_1$ and word$_2$" and its reverse) from each corpus via infinigram. As @fig-corpus-freq shows, the BabyLM corpus produces a highly degenerate frequency distribution: 70.7% of pairs appear exactly once, 87.4% appear at most twice, and 98.2% appear at most ten times, with a maximum of 1,095 occurrences. The Pile distribution is far richer: the median pair appears 365 times, the 25th percentile is 62 occurrences, and the 99th percentile exceeds 77,000.

In absolute training exposure terms (accounting for BabyLM's 20 epochs), the most frequent BabyLM binomials are seen at most \~21,900 times total across all training steps, while the equivalent Pile binomials are seen over 1.5 million times in a single pass. Even the median Pile pair (365 occurrences per pass) substantially exceeds BabyLM's total exposure for a typical singleton pair (1 × 20 = 20 exposures). This gap in absolute exposure is the proximate explanation for why idiosyncratic memorization — reflected in positive $\hat{\beta}_2$ — emerges immediately in Pythia but requires hundreds of millions of tokens to appear in BabyLM.
